---
title: "Stata (.dta) Files"
author: "Daniel Manrique-Castano"
date: "2025-12-01"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
params:
  target_dir: "."
---

## Overview

This notebook inspects **Stata data files (`.dta`)**. Stata datasets are widely used in economics and epidemiology. A well-curated Stata file should be self-documenting, meaning every column should have a clear descriptive label.

**Key Curation Objectives:**

1.  **Inventory:** Verify the number of observations (rows) and variables (columns).
2.  **Metadata Integrity:** Ensure that variable names (e.g., `educ`) are mapped to descriptive labels (e.g., *"Highest year of school completed"*).
3.  **Code Validity:** Check if categorical variables use embedded value labels (e.g., `1="Married"`) rather than raw numbers.

This process generates a **Data Dictionary** in CSV format, providing a complete overview of the dataset's metadata structure.

---

## Setup

We use the `haven` package, which provides a high-fidelity interface for reading Stata files into R, preserving the specific attributes used by Stata.

### R Packages

If you do not have the required packages, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "haven", "rstudioapi"))

```

### Load libraries

```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(haven)
library(rstudioapi)
```


## Select Target Directory

Select the folder containing the Stata files you wish to inspect.

**Note:** If running interactively (RStudio), a dialog box will appear. If running via command line or rendering, it defaults to the target_dir parameter defined at the top of this file.

```{r}
#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select Stata Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))
```

## Find Stata Files

We scan the directory for files ending in .dta.

```{r}
#| label: find-files

dta_files <- list.files(
  path = target_dir,
  pattern = "\\.dta$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

print(paste("Found", length(dta_files), "Stata files."))
head(dta_files)
```

## Generate Data Dictionary

This code reads the metadata headers of each Stata file. By using n_max = 100, we avoid loading large datasets into memory while still capturing all variable definitions.

```{r}
#| label: generate-report
#| warning: false
#| message: false

message("Generating Data Dictionary...")

report <- purrr::map_dfr(dta_files, function(file_path) {
  
  fname <- basename(file_path)
  
  tryCatch({
    # Read header only
    data <- read_dta(file_path, n_max = 100)
    
    purrr::map_dfr(names(data), function(var) {
      
      col <- data[[var]]
      
      # 1. Get Variable Label
      lbl <- attr(col, "label", exact = TRUE)
      if (is.null(lbl)) lbl <- "(No Label)"
      
      # 2. Get Value Labels
      # Stata labels are stored similarly to SPSS in 'haven' objects
      val_lbls <- attr(col, "labels", exact = TRUE)
      
      if (!is.null(val_lbls)) {
        val_str <- paste(val_lbls, names(val_lbls), sep="=", collapse="; ")
      } else {
        val_str <- ""
      }
      
      # 3. Create Summary Row
      tibble(
        FileName = fname,
        VariableName = var,
        VariableLabel = lbl,
        DataType = class(col)[1], # e.g., haven_labelled, double
        ValueLabels = substr(val_str, 1, 100)
      )
    })
    
  }, error = function(e) {
    # Error Handling
    tibble(
      FileName = fname, 
      VariableName = "ERROR", 
      VariableLabel = e$message, 
      DataType = "Error", 
      ValueLabels = ""
    )
  })
})

# Display preview
print("--- Stata Data Dictionary Preview ---")
print(head(report))
```


## Save Results

Save the dictionary to a CSV file for review.

```{r}
#| label: save-results

# Define output directory
output_dir <- file.path("Results", "FileOpen_dta")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Define filename
output_file <- file.path(output_dir, paste0("Stata_Dictionary_", Sys.Date(), ".csv"))

# Write CSV
write.csv(report, output_file, row.names = FALSE)

print(paste("Data Dictionary saved to:", output_file))
```

## Curation Insights

Use the output CSV to check for these common issues:

- Truncated Labels: Older versions of Stata (v12 and earlier) had strict character limits (80 characters) for variable labels. Check if descriptions are cut off mid-sentence.

- Date/Time Formatting: Stata handles dates differently than R. Check the DataType column. If you see raw numbers (e.g., 14832) where you expect dates, confirm that the file format version is compatible or note that conversion will be required.

- Missing Descriptions: Filter for VariableLabel == "(No Label)". In Stata datasets, abbreviated variable names (e.g., x1, y_cnt) are common; without labels, these variables are functionally useless to secondary users.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server (HPC), in a batch job, or from the command line, here is the pure R script version.

### The `FileOpen_dta_Script.R` Script

Download the **R Script:** **[`FileOpen_dta_Script.R`](Scripts/FileOpen_dta_Script.R)**

### Example HPC Submission Script (`FileOpen_dta_submit.sh`)

```bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:10:00
#SBATCH --job-name=stata_check

# Load R module
module load R

# Define data directory
DATA_DIR="/scratch/your_user/your_project/economics_data"

# Run Script
Rscript FileOpen_dta_Script $DATA_DIR
```

