---
title: "Image Files (General)"
author: "Natalie Williams"
date: "2025-10-14"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
params:
  target_dir: "data/FileChecks_Images/" 
---

## Overview

This notebook provides a workflow for performing standardized quality control checks on image files. It automatically scans a directory for common image formats (`.jpg`, `.jpeg`, `.png`, `.tiff`), extracts comprehensive metadata, and generates summary reports.

The process uses two specialized R packages to create two distinct levels of reporting:

1.  **`magick`**: Provides a high-level technical summary that allows the verification of valid images and checking elemental properties like dimensions, colorspace, and file size.
2.  **`exiftoolr`**: Performs a deep extraction of all embedded EXIF metadata, including camera/scanner model, software versions, GPS coordinates, and creation timestamps.

Finally, we offer a guide on how to use the extracted metadata to identify common curation issues, such as inconsistent image dimensions, missing location data, or potential privacy concerns.

------------------------------------------------------------------------

## One-time setup

Before running this notebook, ensure the required R packages and the `ExifTool` command-line software are installed.

### R Packages

The following R packages are required. If you don't have them, uncomment this code and run it once in your R console:

```{r}
# install.packages(c("tidyverse", "magick", "exiftoolr", "rstudioapi"))
```

### ExifTool software

The `exiftoolr` package is a wrapper around a command-line tool called **ExifTool**. You must install this tool for the detailed metadata extraction to work. **Run the following command *once* in your console to install the software.**

```{r}
# exiftoolr::install_exiftool()
```

------------------------------------------------------------------------

## Load libraries

Load all the necessary libraries for the session.

```{r load-libraries}
#| label: load-libraries
#| message: false
library(tidyverse)
library(magick)
library(exiftoolr)
# The rstudioapi package is needed for the interactive folder selection
library(rstudioapi)
```

------------------------------------------------------------------------

## Select a target directory

Here, we interactively select the directory to be analyzed.

> **Note:** If you are running this interactively in RStudio, a dialog box will appear. If you are rendering this document (where no user interaction is possible), it will default to the `params$target_dir` defined in the YAML header.

```{r select-directory}
#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  # .Platform check helps avoid errors in some non-interactive rendering contexts
  selected_dir <- rstudioapi::selectDirectory(caption = "Select Image Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory (Interactive vs Parameter)
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))
```

### Find Image Files

The code uses a single command to find all files ending in `.jpg`, `.jpeg`, `.png`, or `.tiff` within the selected directory.

```{r find-files}
#| label: find-files

# Find all image files using a regular expression for the extensions
image_files <- list.files(
  path = target_dir,
  pattern = "\\.(jpg|jpeg|png|tiff|tif)$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

# Print the number of files found and show the first few paths
print(paste("Found", length(image_files), "image files."))
head(image_files)
```

------------------------------------------------------------------------

## Technical Summary with `magick`

The `magick` package allow us to confirm these files are readable and for extracting basic technical properties. This step loops through each file, reads its information, and collects it into a single table.

```{r magick-summary}
#| label: magick-summary
#| message: false
#| warning: false

# Use purrr::map_dfr to efficiently loop through files and bind results
magick_info <- purrr::map_dfr(image_files, function(file_path) {
  tryCatch({
    # image_read() validates the file, image_info() extracts metadata
    image_read(file_path) %>%
      image_info() %>%
      # Add the filename to the summary table
      mutate(SourceFile = file_path)
  }, error = function(e) {
    # If a file is corrupt or not a valid image, print a warning
    message("Could not process file with magick: ", file_path, " | Error: ", e$message)
    # Return NULL for this file so map_dfr skips it
    return(NULL)
  })
})

# Display the first few rows of the technical summary table
print("--- Magick Technical Summary ---")
head(magick_info)

# Save the results to a CSV file
# Create output directory if it doesn't exist
dir.create("Results/FileChecks_Images", recursive = TRUE, showWarnings = FALSE)

write.csv(
  magick_info, 
  file = paste0("Results/FileChecks_Images/Image_Metadata_Magick_", Sys.Date(), ".csv"), 
  row.names = FALSE
)
```

------------------------------------------------------------------------

## Detailed EXIF Metadata with `exiftoolr`

This step uses `exiftoolr` to extract all available embedded metadata from the image files. This can include hundreds of fields detailing everything from the camera settings and lens information to GPS coordinates and software versions. Please note that the output can have many columns.

```{r exiftool-summary}
#| label: exiftool-summary
#| message: false
#| warning: false
#| results: 'hide'

# exif_read processes all files at once (no loop needed)
exif_info <- tryCatch({
  exif_read(image_files)
}, error = function(e) {
  message("Error running ExifTool. Ensure it is installed. Error: ", e$message)
  return(data.frame())
})

# Save the detailed results to a CSV file if data exists
if (nrow(exif_info) > 0) {
  write.csv(
    exif_info,
    file = paste0("Results/FileChecks_Images/Image_Metadata_Exif_", Sys.Date(), ".csv"),
    row.names = FALSE
  )
  print("ExifTool metadata saved.")
} else {
  print("No EXIF metadata could be extracted.")
}

head(exif_info)
```

------------------------------------------------------------------------

## Additional tools

We can perform additional checks for images using the following chunks.

### Identify inconsistent image dimensions

In many datasets, all images are expected to have the same dimensions (width and height). A simple summary can quickly flag any outliers.

```{r curation-check-dimensions}
#| label: curation-check-dimensions
#| code-fold: false

if (nrow(magick_info) > 0) {
  dimension_summary <- magick_info %>%
    count(width, height, format, name = "file_count") %>%
    arrange(desc(file_count))
  
  print("Summary of Image Dimensions:")
  print(dimension_summary)
}
```

### Summarize a Key EXIF Field (e.g., Megapixels)

Verifying that all images were captured with the same device can be a critical quality check.

```{r curation-check-model}
#| label: curation-check-model
#| code-fold: false

if (nrow(exif_info) > 0) {
  # We only need the Make and Model columns from the exif_info table
  device_summary <- exif_info %>%
    # Check if the columns exist before trying to use them
    { if ("Megapixels" %in% names(.)) 
        count(., Megapixels, name = "file_count") 
      else 
        tibble(Message = "Megapixels metadata not found.")
    }
  
  print("Summary of Megapixels:")
  print(device_summary)
}
```

### Check for Geospatial Information (Privacy/Completeness)

It's important to know if your image files contain embedded GPS coordinates. This can be a privacy concern if the data is meant to be anonymous, or a completeness issue if the location data is required.

```{r curation-check-gps}
#| label: curation-check-gps
#| code-fold: false

if (nrow(exif_info) > 0) {
  # Check for the presence of the GPSLatitude column
  if ("GPSLatitude" %in% names(exif_info)) {
    
    gps_summary <- exif_info %>%
      summarise(
        files_with_gps = sum(!is.na(GPSLatitude)),
        files_missing_gps = sum(is.na(GPSLatitude)),
        total_files = n()
      )
    
    print("Summary of GPS Metadata:")
    print(gps_summary)
    
  } else {
    print("No GPS metadata columns found in any of the files.")
  }
}
```

------------------------------------------------------------------------

## 🚀 Using the Non-Interactive R Script

For users who want to run this analysis on a server, in a batch job, or from the command line, here is a pure R script that performs the same process.

**Prerequisites:**
* R with `tidyverse`, `magick`, and `exiftoolr` installed.
* **ExifTool** command-line software must be installed on the system.

Download the **R Script:** **[`FileChecks_Images_Script.R`](Scripts/FileChecks_Images_Script.R)**


### Example HPC Submission Script 

```bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:20:00
#SBATCH --job-name=img_check

module load R
# module load exiftool # Uncomment if your cluster requires this module

DATA_DIR="/scratch/your_user/your_project/images_v1"

Rscript FileChecks_Images_Script.R $DATA_DIR
```