---
title: "MATLAB Files (.mat)"
author: "Daniel Manrique-Castano"
date: "2025-11-14"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
params:
  target_dir: "."
---

## Overview

This notebook provides a robust workflow for inspecting **MATLAB data files (`.mat`)**. It supports both:

* **Classic Files (v4/v5):** Unpacked using `R.matlab`.
* **Modern Files (v7.3+):** Unpacked using `hdf5r` (HDF5 format).

**The process reports:**
1.  **Variable Name**: The specific object identifier.
2.  **Class**: The data type (e.g., `double`, `cell`, `H5Group`).
3.  **Dimensions**: The size of the array or matrix.

---

## Setup

This workflow relies on the `R.matlab` package.

### R Packages

If you don't have the packages installed, run this command once in your R console:

```{r}
#install.packages(c("tidyverse", "R.matlab", "hdf5r",  "rstudioapi"))
```


```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(R.matlab)
library(hdf5r)
library(rstudioapi)
```

```{r}
# FUNCTION: Checks the file's first 4 bytes for the HDF5 signature
# This is a robust way to identify MATLAB v7.3 files.
is_v73_hdf5 <- function(path) {
  # Open file in binary read mode
  con <- file(path, "rb")
  # Ensure connection is closed when the function exits
  on.exit(close(con)) 
  
  # Read the first 4 bytes (HDF5 signature starts with 0x89HDF)
  header <- tryCatch({
    readBin(con, what = "raw", n = 4, size = 1)
  }, error = function(e) {
    # File might be too small or inaccessible
    return(as.raw(0))
  })
  
  # Check if the header matches the HDF5 signature
  return(identical(header, as.raw(c(0x89, 0x48, 0x44, 0x46))))
}
```


## Select Target Directory

Here we search the selected directory (and its subfolders) for all files ending in .mat.The output gives the number of files detected and prints their full paths.

Note: Interactive selection works in RStudio. If rendering, it defaults to params$target_dir.

```{r}
#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select MATLAB Data Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))

```

## Generate Metadata Report

MATLAB files come in two broad categories:

- v4/v5 files, stored in a proprietary binary format

- v7.3 files, stored as HDF5 (Hierarchical Data Format)

We try to read each .mat file as HDF5 first. If that succeeds, we extract:

- variable (dataset) names
- data type (HDF5 class)
- array dimensions
- number of dimensions
- number of elements

If the file is not HDF5, the code falls back to R.matlab::readMat() and extracts the same information from a v4/v5 structure.

```{r}
# Recompute .mat file list inside this chunk
mat_files <- list.files(
  path        = target_dir,
  pattern     = "\\.mat$",
  recursive   = TRUE,
  full.names  = TRUE,
  ignore.case = TRUE
)

cat("Analyzing directory:", normalizePath(target_dir), "\n")
cat("Found", length(mat_files), "MATLAB files.\n")
print(mat_files)

if (length(mat_files) == 0) {
  warning("No .mat files found. Check 'target_dir' in the YAML params.")
}

# Helper to parse dims from either numeric vector or "3000 x 180 x 180" string
parse_dims <- function(d) {
  # Missing
  if (is.null(d) || (length(d) == 1 && is.na(d))) {
    return(list(str = "-", n_dim = 0L, n_el = NA_integer_))
  }

  # Character like "3000 x 180 x 180"
  if (is.character(d)) {
    parts <- strsplit(d, "x|×")[[1]]
    parts <- trimws(parts)
    parts <- parts[parts != ""]
    nums  <- suppressWarnings(as.integer(parts))

    if (length(nums) == 0 || all(is.na(nums))) {
      # Fallback: just show raw string
      return(list(str = d, n_dim = NA_integer_, n_el = NA_integer_))
    } else {
      return(list(
        str   = paste(nums, collapse = " x "),
        n_dim = length(nums),
        n_el  = as.integer(prod(nums))
      ))
    }
  }

  # Numeric or integer vector
  nums <- suppressWarnings(as.integer(d))
  nums <- nums[!is.na(nums)]
  if (length(nums) == 0) {
    return(list(str = "-", n_dim = 0L, n_el = NA_integer_))
  } else {
    return(list(
      str   = paste(nums, collapse = " x "),
      n_dim = length(nums),
      n_el  = as.integer(prod(nums))
    ))
  }
}

file_summary_list <- list()
var_summary_list  <- list()
idx <- 1L

for (file_path in mat_files) {
  file_name <- basename(file_path)
  cat("\n--- Processing file:", file_name, "---\n")

  file_size_mb <- round(file.info(file_path)$size / (1024^2), 3)

  # 1) TRY: read as HDF5 (MAT v7.3)
  res <- tryCatch({
    f <- hdf5r::H5File$new(file_path, mode = "r")

    # Make sure closing the file never throws a hard error
    on.exit(
      tryCatch(
        f$close_all(),
        error = function(e) {
          message("  close_all() warning for ", file_name, ": ", e$message)
        }
      ),
      add = TRUE
    )

    listing <- f$ls(recursive = TRUE)

    # Keep only datasets
    ds <- listing[listing$obj_type == "H5I_DATASET", , drop = FALSE]

    if (nrow(ds) == 0) stop("HDF5 file contains no datasets")

    dims_list <- ds$dataset.dims

    parsed <- lapply(dims_list, parse_dims)

    dims_str  <- vapply(parsed, function(x) x$str,   character(1))
    num_dims  <- vapply(parsed, function(x) x$n_dim, integer(1))
    num_elems <- vapply(parsed, function(x) x$n_el,  integer(1))

    out <- tibble::tibble(
      FileName      = file_name,
      VariableName  = ds$name,
      Class         = as.character(ds$dataset.type_class),
      Dimensions    = dims_str,
      NumDimensions = num_dims,
      NumElements   = num_elems,
      Format        = "v7.3 (HDF5)"
    )

    cat("  Read as HDF5 (v7.3);", nrow(out), "datasets.\n")
    out
  },
  error = function(e) {
    cat("  HDF5 attempt failed for ", file_name, ": ",
        conditionMessage(e), "\n", sep = "")
    NULL
  })

  # 2) FALLBACK: classic MAT (v4/v5) via readMat, only if HDF5 failed
  if (is.null(res)) {
    res <- tryCatch({
      mat_data  <- R.matlab::readMat(file_path)
      var_names <- names(mat_data)

      if (length(var_names) == 0) {
        cat("  readMat(): file appears empty.\n")
        tibble::tibble(
          FileName      = file_name,
          VariableName  = "(Empty)",
          Class         = "-",
          Dimensions    = "-",
          NumDimensions = 0L,
          NumElements   = NA_integer_,
          Format        = "v4/v5"
        )
      } else {
        cat("  readMat(): ", length(var_names), " variables.\n", sep = "")
        purrr::map_dfr(var_names, function(var) {
          obj  <- mat_data[[var]]
          dims <- dim(obj)
          dims_str <- if (is.null(dims)) {
            paste(length(obj), "(len)")
          } else {
            paste(as.integer(dims), collapse = " x ")
          }
          n_dim <- if (is.null(dims)) 1L else length(dims)
          n_el  <- if (is.null(dims)) length(obj) else as.integer(prod(dims))

          tibble::tibble(
            FileName      = file_name,
            VariableName  = var,
            Class         = class(obj)[1],
            Dimensions    = dims_str,
            NumDimensions = n_dim,
            NumElements   = n_el,
            Format        = "v4/v5"
          )
        })
      }
    },
    error = function(e) {
      cat("  readMat() failed for ", file_name, ": ",
          conditionMessage(e), "\n", sep = "")
      tibble::tibble(
        FileName      = file_name,
        VariableName  = "ERROR",
        Class         = "Error",
        Dimensions    = paste("Failed to read:", conditionMessage(e)),
        NumDimensions = NA_integer_,
        NumElements   = NA_integer_,
        Format        = "Unknown"
      )
    })
  }

  # File-level summary
  file_summary_list[[idx]] <- tibble::tibble(
    FileName     = file_name,
    FullPath     = normalizePath(file_path),
    Format       = unique(res$Format)[1],
    FileSize_MB  = file_size_mb,
    NumVariables = nrow(res)
  )

  # Variable-level summary
  var_summary_list[[idx]] <- res
  idx <- idx + 1L
}

# Combine summaries
file_report <- dplyr::bind_rows(file_summary_list)
mat_report  <- dplyr::bind_rows(var_summary_list)

cat("\n=== File-level MATLAB summary ===\n")
knitr::kable(file_report)

cat("\n\n=== Variable-level MATLAB summary ===\n")
knitr::kable(mat_report)

# Write detailed variable-level report to CSV
if (!dir.exists("Results/FileOpen_mat")) {
  dir.create("Results/FileOpen_mat", recursive = TRUE, showWarnings = FALSE)
}

utils::write.csv(
  mat_report,
  file = file.path(
    "Results/FileOpen_mat",
    paste0("Matlab_Detailed_Report_", Sys.Date(), ".csv")
  ),
  row.names = FALSE
)
```


## Notes and limitations

MATLAB .mat files can vary significantly in format and complexity. We have some limitations to keep in mind:

- Very large datasets may not be fully loaded and may be summarized based on sampling.
- Deeply nested MATLAB structs or objects may require custom handling.
- Character encoding in .mat files may differ across MATLAB versions.
- Complex cell arrays can contain heterogeneous types that cannot be summarized numerically.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server, in a batch job, or from the command line, here is a pure R script that performs the same process.

Download the **R Script:** **[`FileOpen_mat_Script.R`](Scripts/FileOpen_mat_Script.R)**

### Example HPC Submission Script (`FileOpen_mat_Script.sh`)

```bash
#!/bin/bash
#SBATCH --job-name=matlab_inspect
#SBATCH --cpus-per-task=2
#SBATCH --mem=8G
#SBATCH --time=01:00:00
#SBATCH --output=matlab_inspect_%j.out
#SBATCH --error=matlab_inspect_%j.err

module load R/4.3.0   # adapt to your cluster's module system

Rscript matlab_inspector.R \
  --target_dir /path/to/your/mat/files \
  --output_dir /path/to/output/folder \
  --max_full_numeric 1000000 \
  --max_sample_numeric 100000
  
  ```
