---
title: "SPSS (.sav) Files"
author: "Daniel Manrique-Castano"
date: "2025-11-19"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
params:
  target_dir: "."
---

## Overview

This notebook inspects **SPSS data files (`.sav`)** to extract the internal **Data Dictionary**.

**The report includes:**
1.  **File Info:** Row and Column counts.
2.  **Variable Metadata:** The Variable Name (e.g., `q1`) and the embedded Variable Label (e.g., "What is your age?").
3.  **Value Labels:** A summary of categorical definitions (e.g., `1=Yes, 0=No`).

## Setup

We use the `haven` package to read SPSS files. It is specifically designed to handle the "labelled" vectors that are unique to statistical software like SPSS, SAS, and Stata.If you do not have the required packages, run this command once in your R console:

### R Packages

```{r}
# install.packages(c("tidyverse", "```", "rstudioapi"))
```

### Load Libraries
```{r}
#| message: false
library(tidyverse)
library(haven)
library(rstudioapi)
```

## Select Directory

Select the folder containing the SPSS files you wish to inspect.

**Note:** If running interactively (RStudio), a dialog box will appear. If running via command line or rendering, it defaults to the target_dir parameter defined at the top of this file.

```{r}
#| label: select-target-dir
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select SPSS Directory")
} else { selected_dir <- NULL }

target_dir <- if (!is.null(selected_dir)) selected_dir else params$target_dir
print(paste("Analyzing:", target_dir))
```
## Find SPSS files

We scan the directory for files ending in .sav.

```{r}
#| label: find-files

spss_files <- list.files(
  path = target_dir,
  pattern = "\\.sav$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

print(paste("Found", length(spss_files), "SPSS files."))
head(spss_files)
```

## Generate Data Dictionary

Here, the code loops through every file found and performs the following:

- Reads the Header: It reads only the first 100 rows (n_max = 100). This makes the process extremely fast, even for multi-gigabyte files, because we only need the metadata headers, not the full dataset.

- Extracts Metadata: It uses the attr() function to pull out the hidden "label" and "labels" attributes attached to each column.

- Formats Output: It creates a tidy table listing every variable, its type, and its definitions.


```{r}
#| label: generate-report
#| warning: false
#| message: false

message("Generating Data Dictionary...")

# Map over every file found
report <- purrr::map_dfr(spss_files, function(file_path) {
  
  fname <- basename(file_path)
  
  tryCatch({
    # Read a small subset to get metadata without memory overhead
    data <- read_sav(file_path, n_max = 100)
    
    # Map over every column (variable) in the dataset
    purrr::map_dfr(names(data), function(var) {
      
      # Extract the column object
      col <- data[[var]]
      
      # 1. Get Variable Label (The "Question")
      lbl <- attr(col, "label", exact = TRUE)
      if (is.null(lbl)) lbl <- "(No Label)"
      
      # 2. Get Value Labels (The "Answers")
      # These come as a named vector: c(Yes = 1, No = 0)
      val_lbls <- attr(col, "labels", exact = TRUE)
      
      # Format value labels into a readable string: "1=Yes; 0=No"
      if (!is.null(val_lbls)) {
        val_str <- paste(val_lbls, names(val_lbls), sep="=", collapse="; ")
      } else {
        val_str <- ""
      }
      
      # 3. Create Summary Row
      tibble(
        FileName = fname,
        TotalRowsInFile = nrow(data), # Note: If n_max used, this is partial
        VariableName = var,
        VariableLabel = lbl,
        DataType = class(col)[1], # e.g., haven_labelled, numeric, character
        ValueLabels = substr(val_str, 1, 100) # Truncate long lists for readability
      )
    })
    
  }, error = function(e) {
    # Error Handling: If file is corrupt
    tibble(
      FileName = fname, 
      VariableName = "ERROR", 
      VariableLabel = e$message, 
      DataType = "Error", 
      ValueLabels = ""
    )
  })
})

# Display the first few rows of the dictionary
print("--- SPSS Data Dictionary Preview ---")
print(head(report))
```

## Save the results

We save the resulting table to a CSV file. This CSV acts as the Data Dictionary for the curated deposit.

```{r}
#| label: save-results

# Define output directory
output_dir <- file.path("Results", "OpenFile_sav")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Define filename with timestamp
output_file <- file.path(output_dir, paste0("OpenFile_sav_", Sys.Date(), ".csv"))

# Write CSV
write.csv(report, output_file, row.names = FALSE)

print(paste("Data Dictionary saved to:", output_file))
```


## Curation Insights

Use the generated CSV to perform these checks:

- Missing Variable Labels: Filter the CSV for VariableLabel == "(No Label)".

- Missing Value Labels: Look for variables that are DataType = haven_labelled or numeric but have empty ValueLabels.

- Personal identifiers detection: Scan the VariableLabel column for keywords like "Name", "Phone", "Address", or "IP". SPSS labels often contain the full original survey question, which might inadvertently reveal that PII was collected.

## Using the Non-Interactive R Script
For users who want to run this analysis on a server (HPC), in a batch job, or from the command line, here is the pure R script version.

### The `FileOpen_sav_Script.R` Script

Download the **R Script:** **[`FileOpen_sav_Script.R`](Scripts/FileOpen_sav_Script.R)**

### Example HPC Submission Script (`FileOpen_sav_submit.sh`)

```bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:10:00
#SBATCH --job-name=spss_check

# Load R module
module load R

# Define data directory
DATA_DIR="/scratch/your_user/your_project/survey_data"

# Run Script
Rscript FileOpen_sav_Script.R $DATA_DIR
```

