---
title: "NetCDF Files"
author: "Natalie Williams"
date: "2025-10-16"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
params:
  target_dir: "."
---

## Overview

This tool provides an automated workflow for extracting and summarizing metadata from a collection of NetCDF (`.nc`) file, a self-describing format common in scientific fields. The workflow performs three main tasks:

1.  **File discovery**: It scans a specified directory for all NetCDF files.
2.  **Metadata extraction**: It uses the `tidync` package to pull out three fundamental components from each file:
    * **Dimensions**: The axes of the data grid (e.g., latitude, longitude, time).
    * **Variables**: The actual data stored in the file (e.g., temperature, pressure).
    * **Attributes**: Descriptive metadata, which can be global (applying to the whole file) or attached to specific variables (e.g., units).
3.  **Report generation**: It reshapes this metadata into user-friendly summary tables and saves them as `.csv` files.

------------------------------------------------------------------------

## Setup

Before running this notebook, you need to ensure the required R packages are installed.

### R Packages

The following R packages are required. If you don't have them, run this code once in your R console:

```{r}
# install.packages(c("tidyverse", "tidync", "rstudioapi"))
```

------------------------------------------------------------------------

## Load Libraries

This chunk loads all the necessary libraries for the session.

```{r load-libraries}
#| label: load-libraries
#| message: false

library(tidyverse)
library(tidync)
library(rstudioapi) # Added for interactive selection
```

------------------------------------------------------------------------

## Select target directory with NetCDF files

This section identifies the directory to be analyzed and finds all `.nc` files within it.

> **Note:** If you are running this interactively in RStudio, a dialog box will appear. If you are rendering this document (where no user interaction is possible), it will default to the `params$target_dir` defined in the YAML header.

```{r select-directory}
#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select NetCDF Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory (Interactive vs Parameter)
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))
```

### Find NetCDF files

Now we scan the selected directory for all `.nc` files.

```{r find-files}
#| label: find-files

# Find all NetCDF files
nc_files <- list.files(
  path = target_dir,
  pattern = "\\.nc$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

# Print the number of files found and show the first few paths
print(paste("Found", length(nc_files), "NetCDF files."))
head(nc_files)
```

------------------------------------------------------------------------

## Extract and organize metadata

Here, we will process each NetCDF file using `purrr::map` functions to extract its dimensions, variables, and attributes.

```{r extract-metadata}
#| label: extract-metadata

# Create a "safely" version of tidync to handle potentially corrupt files
safe_tidync <- purrr::safely(tidync)

# 1. Process all files, safely handling any errors
processed_files <- purrr::map(nc_files, ~safe_tidync(.x)) %>% 
  set_names(nc_files)

# 2. Separate successful results from errors
successful_results <- purrr::map(processed_files, "result") %>% 
  purrr::compact()

errors <- purrr::map(processed_files, "error") %>% 
  purrr::compact()

# Report any files that failed to process
if (length(errors) > 0) {
  message("The following files failed to process and were skipped:")
  walk(names(errors), message)
}

# 3. Extract and combine dimensions, variables, and attributes
nc_dimensions <- purrr::map(successful_results, ~.x$dimension) %>% 
  bind_rows(.id = "FileName")

nc_variables <- purrr::map(successful_results, ~.x$variable) %>% 
  bind_rows(.id = "FileName")

nc_attributes <- purrr::map(successful_results, ~.x$attribute) %>% 
  bind_rows(.id = "FileName")

print("Metadata extraction complete.")
```

------------------------------------------------------------------------

## Reshape metadata for comparison

We reshape the tables to facilitate cross-file comparison. We create two primary summary tables: one for global attributes and another that joins variables with their specific attributes.

### Global attributes summary

This table gives a wide-format overview of the global attributes, with one row per file. It helps checking the consistency at the file-level, including `title`, `history`, or `source`.

```{r reshape-global-attributes}
#| label: reshape-global-attributes

nc_attributes_global <- nc_attributes %>%
  filter(variable == "NC_GLOBAL") %>%
  # Pivot to make each global attribute its own column
  pivot_wider(
    id_cols = FileName,
    names_from = name,
    values_from = value,
    values_fn = ~paste(., collapse = "; ")
  )

print("--- Global Attributes Summary ---")
glimpse(nc_attributes_global)
```

### Variables and Attributes Summary

This table provides a detailed, wide-format view of each variable from each file and its associated attributes (like `units` or `long_name`).

```{r reshape-variable-attributes}
#| label: reshape-variable-attributes

nc_variables_with_attributes <- nc_variables %>%
  # Join the variables with their non-global attributes
  left_join(
    filter(nc_attributes, variable != "NC_GLOBAL"),
    by = c("name" = "variable", "FileName")
  ) %>%
  # Pivot to make each attribute (units, long_name, etc.) its own column
  pivot_wider(
    names_from = name.y, # The attribute name
    values_from = value,
    values_fn = ~paste(., collapse = "; ")
  )

print("--- Variables and Attributes Summary ---")
head(nc_variables_with_attributes)
```

------------------------------------------------------------------------

## Save Summary Reports

Finally, we save the most useful summary tables to `.csv` files for documentation and further analysis.

```{r write-files}
#| label: write-files

# Create output directory if it doesn't exist
dir.create("Results/NCChecker", recursive = TRUE, showWarnings = FALSE)

write.csv(
  nc_dimensions,
  file = paste0("Results/NCChecker/NetCDF_Dimensions_", Sys.Date(), ".csv"),
  row.names = FALSE
)
write.csv(
  nc_attributes_global,
  file = paste0("Results/NCChecker/NetCDF_Global_Attributes_", Sys.Date(), ".csv"),
  row.names = FALSE
)
write.csv(
  nc_variables_with_attributes,
  file = paste0("Results/NCChecker/NetCDF_Variables_with_Attributes_", Sys.Date(), ".csv"),
  row.names = FALSE
)

print("Summary reports have been saved to CSV files.")
```

------------------------------------------------------------------------

## Additional tools

### Verify consistent global attributes

Verify if all files in a dataset have the same `title`, `institution`, `source`, and CF `Conventions`.

```{r curation-check-global}
#| label: curation-check-global
#| code-fold: false

if (nrow(nc_attributes_global) > 0) {
  # Select a few key attributes and count the unique combinations
  global_consistency_check <- nc_attributes_global %>%
    select(FileName, contains("title"), contains("institution"), contains("source"), contains("Conventions")) %>%
    # The line below groups by all columns except FileName
    group_by(across(-FileName)) %>%
    summarise(file_count = n(), .groups = "drop")
  
  print("Consistency Check of Key Global Attributes:")
  print(global_consistency_check)
}
```

### Check for essential variable attributes

For data to be reusable, variables should always have attributes like `long_name` and `units`. This section allows checking for missing attributes across all variables.

```{r curation-check-var-attr}
#| label: curation-check-var-attr
#| code-fold: false

if (nrow(nc_variables_with_attributes) > 0) {
  missing_attribute_check <- nc_variables_with_attributes %>%
    # Summarise the number of variables missing these key attributes
    summarise(
      missing_long_name = sum(is.na(long_name)),
      missing_units = sum(is.na(units))
    )
  
  print("Check for Missing Essential Variable Attributes:")
  print(missing_attribute_check)
}
```


## Using the Non-Interactive R Script

For users who want to run this analysis on a server, in a batch job, or from the command line, here is a pure R script that performs the same process.

### The `FileOpen_nc_Script.R` Script

Download the **R Script:** **[`FileOpen_nc_Script.R`](Scripts/FileOpen_nc_Script.R)**

### Example HPC Submission Script (`FileOpen_nc_submit.sh`)

```bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:15:00
#SBATCH --job-name=nc_check

# Load R module (often includes NetCDF/HDF5 libraries)
module load R 

DATA_DIR="/scratch/your_user/your_project/netcdf_data"

Rscript FileOpen_nc_Script.R $DATA_DIR
```