---
title: "Excel Files (.xls, .xlsx)"
author: "Daniel Manrique-Castano"
date: "2025-11-17"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
params:
  target_dir: "."
---

## Overview

This notebook provides a workflow for inspecting and summarizing the contents of multiple Excel workbooks (`.xls` and `.xlsx`) within a directory. Because Excel files are complex containers, this tool is designed to help curators quickly assess their structure and content *before* attempting a full data conversion.

**The process performs four key checks for each workbook:**

1.  **Lists all sheet names** to identify hidden sheets or unexpected content.
2.  **Gets the true dimensions** (rows and columns) of each sheet.
3.  **Reads the first 5 rows** of each sheet to preview headers and content.
4.  **Attempts to detect non-tabular data** (like charts, pictures, or complex merged-cell headers) by checking for default or minimal column names.

---

## Setup

This workflow relies on the `readxl` package to parse Excel files and `rstudioapi` for the interactive directory selection.

### R Packages

If you don't have these packages installed, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "readxl", "rstudioapi"))
```

---

## Load Libraries

This chunk loads all the necessary libraries for the session.

```{r load-libraries}
#| label: load-libraries
#| message: false

library(tidyverse)
library(readxl)
library(rstudioapi)
```

---

## Select Target Directory

This section identifies the directory to be analyzed and finds all Excel files within it.

> **Note:** If you are running this interactively in RStudio, a dialog box will appear. If you are rendering this document (where no user interaction is possible), it will default to the `params$target_dir` defined in the YAML header.

```{r select-directory}
#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select Excel Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory (Interactive vs Parameter)
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))
```

### Find Excel Files

Now we scan the selected directory for all `.xls` and `.xlsx` files.

```{r find-files}
#| label: find-files

# Find all Excel files using a regex pattern
excel_files <- list.files(
  path = target_dir,
  pattern = "\\.xlsx?$", # Matches .xls or .xlsx
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

# Print the number of files found and show the first few paths
print(paste("Found", length(excel_files), "Excel files."))
head(excel_files)
```

---

## 📊 Generate Structural Report

This is the core of the notebook. The code will loop through each Excel file, then loop through each *sheet* within that file, and generate a summary report.

```{r generate-report}
#| label: generate-report
#| warning: false
#| message: false

# 1. Initialize an empty list to store results
all_sheets_summary <- list()

# 2. Loop over each Excel file
for (file_path in excel_files) {
  
  file_name <- basename(file_path)
  
  # Use a tryCatch in case the file is corrupt or password-protected
  tryCatch({
    
    # Check 1: List all sheet names
    sheet_names <- excel_sheets(file_path)
    
    if (length(sheet_names) == 0) {
      # Handle empty workbooks
      all_sheets_summary[[length(all_sheets_summary) + 1]] <- tibble(
        FileName = file_name,
        SheetName = "(No Sheets Found)",
        Dimensions = "0 x 0",
        IsLikelyNonTabular = TRUE,
        HeaderPreview = "(Empty)"
      )
      next # Skip to the next file
    }
    
    # 3. Loop over each sheet in the workbook
    for (sheet_name in sheet_names) {
      
      # Use a tryCatch for each sheet (a sheet might be corrupt)
      tryCatch({
        
        # Check 2: Get true dimensions by reading the whole sheet
        # We specify .name_repair to prevent messages
        full_sheet_data <- read_excel(file_path, sheet = sheet_name, .name_repair = "minimal")
        true_dims <- paste(dim(full_sheet_data), collapse = " x ")
        
        # Check 3 & 4: Read headers and check for non-tabular data
        # We use .name_repair = "minimal" to detect default '...#' names
        header_preview <- read_excel(file_path, sheet = sheet_name, n_max = 5, .name_repair = "minimal")
        col_names <- colnames(header_preview)
        
        # This is a proxy. It's TRUE if:
        # a) Any column name is a default '...#' (often means merged headers)
        # b) There is only 1 column (often means a chart or text block)
        is_non_tabular <- any(str_detect(col_names, "^\\.\\.\\.\\d+$")) || length(col_names) <= 1
        
        # Store the first 5 rows' headers as a single string
        header_as_string <- paste(col_names, collapse = " | ")
        
        # Add results to our list
        all_sheets_summary[[length(all_sheets_summary) + 1]] <- tibble(
          FileName = file_name,
          SheetName = sheet_name,
          Dimensions = true_dims,
          IsLikelyNonTabular = is_non_tabular,
          HeaderPreview = header_as_string
        )
        
      }, error = function(e_sheet) {
        # Handle a single bad sheet
        message(paste("Error on sheet:", sheet_name, "in file:", file_name, "|", e_sheet$message))
        all_sheets_summary[[length(all_sheets_summary) + 1]] <- tibble(
          FileName = file_name,
          SheetName = sheet_name,
          Dimensions = "Error",
          IsLikelyNonTabular = TRUE,
          HeaderPreview = paste("ERROR:", e_sheet$message)
        )
      })
    }
    
  }, error = function(e_file) {
    # Handle a single bad file
    message(paste("Error on file:", file_name, "|", e_file$message))
    all_sheets_summary[[length(all_sheets_summary) + 1]] <- tibble(
      FileName = file_name,
      SheetName = "Error",
      Dimensions = "Error",
      IsLikelyNonTabular = TRUE,
      HeaderPreview = paste("ERROR:", e_file$message)
    )
  })
}

# 4. Combine all results into one data frame
summary_report <- bind_rows(all_sheets_summary)

print("--- Excel Structure Report ---")
print(summary_report)

# 5. Save the report to a CSV file
dir.create("Results/OpenFile_xlsx", recursive = TRUE, showWarnings = FALSE)
write.csv(
  summary_report,
  file = paste0("Results/OpenFile_xlsx/Excel_Structure_Report_", Sys.Date(), ".csv"),
  row.names = FALSE
)

print("Excel structure report saved.")
```

---

## Curation Insights

The `summary_report` table provides:

* **`SheetName`**: Scan this column for unexpected names like `"Sheet1"`, `"Chart2"`, or `"DELETE_ME"`. This often reveals which sheets are data and which are artifacts.
* **`Dimensions`**: Look for sheets with `0 x 0` or `1 x 1` dimensions. These are almost always empty or non-data sheets that can be ignored or flagged.
* **`IsLikelyNonTabular = TRUE`**: **This is a key flag.** `TRUE` suggests the sheet does not start with a simple, clean table. It could be a chart, a block of text, or have a complex multi-row header (which will require special handling during conversion).
* **`HeaderPreview`**: This gives you a direct look at the column names. You can quickly see if they make sense or if they are `...1 | ...2 | ...3`, confirming the `IsLikelyNonTabular` flag.

---

## Using the Non-Interactive R Script

For users who want to run this analysis on a server or from the command line, here is a pure R script that performs the same process.

This "hybrid" script will:
1.  If run **interactively** (e.g., in RStudio), pop up a dialog to select a **directory**.
2.  If run **non-interactively** (e.g., on an HPC cluster), accept a **directory** path as a command-line argument.
3.  Process **all** `.xls` and `.xlsx` files found in that directory.

Download the **R Script:** **[`FileOpen_xlsx_Script.R`](Scripts/FileOpen_xlsx_Script.R)**


### Example HPC Submission Script (`submit_excel_check.sh`)

```bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:20:00 # Excel parsing can be slow
#SBATCH --job-name=excel_check

# Load R module
module load R 

# Define the folder containing your Excel files
DATA_DIR="/scratch/your_user/your_project/excel_data"

# Run the R script, passing the directory as an argument
Rscript FileOpen_xlsx_Scriptl.R $DATA_DIR
```