---
title: "File extensions"
author: "Natalie Williams"
date: "2025-11-13"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
params:
  target_dir: "data/FileChecks_Extensions/"
---

## Overview

This notebook provides a utility to inventory the contents of a data directory. Its primary purpose is to scan a target folder, identify all files, and generate a summary table counting the occurrences of each unique file extension.

This process helps answer key questions:

-   What types of files are present in the dataset?
-   Are there any unexpected or disallowed file formats?
-   Are there hidden files (e.g., `.DS_Store`, `Thumbs.db`) that should be removed before publication?
-   Does the file count match the dataset's documentation?

The output is a clean CSV file that serves as an inventory of the entire data package.

------------------------------------------------------------------------

## Setup

This workflow uses the `tidyverse` suite of packages, which includes `dplyr` for data manipulation and `stringr` for text processing.

### R Packages

If you don't have the `tidyverse` package installed, run this command once in your R console:

```{r}
# install.packages("tidyverse")
```

------------------------------------------------------------------------

## Load libraries

This chunk loads the necessary library for the session.

```{r load-libraries}
#| label: load-libraries
#| message: false

library(tidyverse)
```

------------------------------------------------------------------------

## Select the target directory

This section identifies the directory to be analyzed and creates a complete list of every file within it, including those in sub-folders and hidden files. The directory is set by the `target_dir` parameter at the top of this document.

```{r find-files}
#| label: find-files

# Get the directory path from the document's parameters
target_dir <- params$target_dir
print(paste("Analyzing directory:", target_dir))

# Find all files, including hidden files (`all.files = TRUE`)
all_files <- list.files(
  path = target_dir,
  recursive = TRUE,
  full.names = TRUE, # Get the full path for robustness
  all.files = TRUE   # Include dotfiles (e.g., .git, .Rproj)
)

# Print the total number of files found
print(paste("Found", length(all_files), "total files (including hidden)."))
```

------------------------------------------------------------------------

## Extract extensions and generate summary table

The code takes the complete list of file paths and performs two steps:

1.  **Extract extensions**: It isolates the file extension from each path. The logic is designed to handle files with multiple dots (e.g., `archive.tar.gz`) and files with no extension at all.
2.  **Count and summarize**: It then counts the occurrences of each unique extension and sorts the results to show the most common file types first.

```{r generate-summary}
#| label: generate-summary

file_summary <- tibble(FilePath = all_files) %>%
  mutate(
    # basename() gets just the filename from the full path
    FileName = basename(FilePath),
    # Use stringr for a more readable and robust extraction
    Extension = str_extract(FileName, "(?<=\\.)[^.]+$") %>%
      # If no extension is found (returns NA), replace with a descriptor
      replace_na("(no extension)")
  ) %>%
  count(Extension, name = "Count", sort = TRUE)

print("--- Summary of File Extensions ---")
print(file_summary)

# Save the results to a CSV file
write.csv(
  file_summary,
  file = paste0("Results/FileCheks_Extensions/File_Extension_Summary_", Sys.Date(), ".csv"),
  row.names = FALSE
)

print("File extension summary has been saved.")
```

------------------------------------------------------------------------

## Additional tools

### Identify Unexpected Formats

Quickly scan the `Extension` column. Do you see any proprietary formats (`.xlsx`, `.psd`, `.paj`) that should have been converted to open, preservation-friendly formats (`.csv`, `.tiff`, `.txt`)?

### Find and Remove System "Junk" Files

Many operating systems create hidden files that are not part of the actual dataset and should be removed before publication.

**Common examples to look for:** \* `DS_Store` (macOS) \* `Thumbs.db` (Windows) \* `__MACOSX` (Created by macOS when zipping files) \* `.Rhistory` or `.Rproj.user` (RStudio project files)

If these appear in your summary, you can locate and delete them to clean the dataset.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server or from the command line, a pure R script is available to perform the same process.

**Prerequisites:**
* R with `tidyverse`, `magick`, and `exiftoolr` installed.
* **ExifTool** command-line software must be installed on the system.

Download the **R Script:** **[`FileChecks_Extensions_Script.R`](Scripts/FileChecks_Extensions_Script.R)**

### Example HPC Submission Script 

```bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:20:00
#SBATCH --job-name=img_check

module load R
# module load exiftool # Uncomment if your cluster requires this

DATA_DIR="/scratch/your_user/your_project/images_v1"

Rscript FileChecks_Extensions_Script.R $DATA_DIR
