---
title: "Automated CSV File Checker"
author: "Natalie Williams"
date: "2025-10-10"
format: html
editor: visual
execute:
  echo: true
  warning: false
---

## Overview

This notebook provides a standardized procedure for checking and summarizing the contents of multiple `.csv` files within a directory. The goal is to automatically generate metadata summaries that can be used for data quality control, documentation, and initial exploratory analysis.

**With this procedure curators can:**

* **Discover files:** Automatically find all `.csv` files in the project directory.
* **Extract metadata:** For each file, gather information on column names, data types, and row counts.
* **Calculate descriptive statistics:** Compute summary statistics (min, max, mean, median) and the count of missing values (`NA`) for each column.
* **Generate summary outputs:** Create two final `.csv` files:
    1.  `MetaData_CSV.csv`: A detailed, long-format summary of every column from every file.
    2.  `ColumnCheck_CSV.csv`: A wide-format summary to easily compare column names and types across all files.

**Assumptions:**

* This notebook is located and run in the root directory of the data deposit.
* All compressed files (e.g., `.zip`, `.tar`) have already been extracted.

---

## Setup: Loading libraries

First, we need to load the `tidyverse` library. We also load the `rstudioapi` library to enable the interactive directory selection prompt (It comes pre-installed with RStudio).

```{r setup}
#| label: load-libraries
library(tidyverse)
library(rstudioapi)
```

---


## Select target directory

First, we select the directory to analyze and find `.csv` files within it.

### Select a Directory

In the chunk below, select the directory you want to analize.

```{r select-directory}
#| label: set-target-dir-manual

# Manually define the path to the directory you want to analyze.
# You must update this path every time you want to check a new folder.
target_dir <- "D:/Alliance/FRDR/Curation/2025/1337/files/v3"

# Print the directory being used to confirm
print(paste("Using directory:", target_dir))
```

### Find CSV Files in the Selected Directory

Now that we have the path to the target directory, this code searches that location and all its subdirectories (`recursive = TRUE`) for files ending with the `.csv` extension.

We set `full.names = TRUE` to ensure that `csv.files` stores the complete file path for each file, which is necessary for the script to read them correctly.

```{r get-files}
#| label: get-csv-files

csv.files <- list.files(
  path = target_dir,       # The directory to search
  pattern = "*.csv",       # The file pattern to look for
  recursive = TRUE,        # Search in sub-folders as well
  full.names = TRUE        # Return the full file path, not just the name
)

# Let's print the full paths of the files found to check
print(csv.files)
```

---

## Generate metadata summaries

Here, the code will loop through each full file path in our `csv.files` list, read the file, and extract the required metadata.

### Initializing the summary table

```{r initialize-df}
#| label: initialize-summary-df

csv.summ <- data.frame(
  file.name = character(),
  column.name = character(),
  column.type = character(),
  rows = integer(),
  min = character(),
  mean = character(),
  med = character(),
  max = character(),
  NAs = character()
)
```

### Processing each file

Here, we systematically loops through each CSV file selected to summarize the data and handle errors. For valid data, we extract metadata (like column names, data types, and row count), and calculate descriptive statistics (min, mean, max, and a count of missing values). When not valid columns are found (corrupted or contains unexpected data types) the script print a detailed message in the console identifying the problematic file and the specific error.

```{r process-files}
#| label: process-files-loop

# The Main Loop: Processing Each File
for (i in csv.files) {
  
  # Use tryCatch to handle potential errors in a file without stopping the whole script
  tryCatch({
    
    # --- All the original processing code is placed inside the tryCatch block ---
    tmp <- read.csv(i)
    
    tmp.meta <- data.frame(
      file.name = basename(i),
      column.name = colnames(tmp),
      column.type = sapply(tmp, class),
      rows = nrow(tmp)
    )
    
    tmp.stat <- tmp %>%
      summarise(
        across(
          .cols = everything(),
          .fns = list(
            min = ~min(.x, na.rm = TRUE),
            mean = ~mean(.x, na.rm = TRUE),
            med = ~median(.x, na.rm = TRUE),
            max = ~max(.x, na.rm = TRUE),
            NAs = ~sum(is.na(.x))
          ),
          .names = "{.col}|{.fn}"
        )
      ) %>%
      mutate(
        across(everything(), as.character)
      ) %>%
      pivot_longer(
        cols = everything(),
        names_to = c("column.name", ".value"),
        names_pattern = "(.*)\\|(.*)"
      )
      
    tmp.meta <- full_join(tmp.meta, tmp.stat, by = "column.name")
    
    csv.summ <- bind_rows(csv.summ, tmp.meta)
    
  }, error = function(e) {
    # If an error occurs for a file, this code will run instead of crashing
    message(paste("--- ERROR --- Failed to process file:", i))
    message(paste("Specific error message:", e$message))
    message("Skipping this file and continuing with the next one.\n")
  })
}

# Clean up temporary variables from the loop
rm(tmp, tmp.stat, tmp.meta)

# View the first few rows of the final detailed summary table
head(csv.summ)
```

---

## Create a comparison table

This step creates the wide-format `col.compare` table, making it easy to spot inconsistencies in column names and data types across files.

```{r create-comparison-table}
#| label: create-col-compare

col.compare <- csv.summ %>%
  pivot_wider(
    id_cols = "file.name",
    names_from = "column.name",
    values_from = "column.type",
    values_fn = ~paste(., collapse = ", ")
  )

# View the column comparison table to confirm it's a flat table
print(col.compare)
```

---

## Write the results to file

Finally, we save our two summary data frames as new `.csv` files. These files will be saved in your **current working directory**, not necessarily the directory you selected for analysis.

```{r write-files}
#| label: write-output-files

write.csv(csv.summ, file = "Results/FileChecks/MetaData_CSV.csv", row.names = FALSE)
write.csv(col.compare, file = "Results/FileChecks/ColumnCheck_CSV.csv", row.names = FALSE)

print("Summary files 'MetaData_CSV.csv' and 'ColumnCheck_CSV.csv' have been created in your current working directory.")
```