---
title: "JSON Files"
author: "Daniel Manrique-Castano"
date: "2025-11-19"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
params:
  target_dir: "."
---

## Overview

This notebook provides a workflow for inspecting and summarizing **JSON (JavaScript Object Notation)** files. JSON files are flexible and can store anything from simple flat tables to complex, deeply nested hierarchies.

**The process performs four key checks for each file:**

1.  **Validation:** Confirms the file is valid JSON and can be parsed.
2.  **Structure Identification:** Determines if the file represents a table (data frame), a list (hierarchy), or an array.
3.  **Top-Level Key Extraction:** Lists the main "branches" or column names to help understand the schema.
4.  **Dimension Check:** Reports the number of rows/columns (for tables) or the number of elements (for lists).

---

## ⚙️ Setup

This workflow relies on `jsonlite` for parsing and `tidyverse` for data manipulation.

### R Packages

If you don't have these packages installed, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "jsonlite", "rstudioapi"))

```

## Load libraries

```{r}

#| label: load-libraries
#| message: false

library(tidyverse)
library(jsonlite)
library(rstudioapi)

```


## Select Target Directory
Select the directory containing your .json files.

Note: Interactive selection works in RStudio. If rendering, it defaults to params$target_dir.

```{r}

#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select JSON Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))
```


## Find Json files

```{r}
#| label: find-files

# Find all .json files
json_files <- list.files(
  path = target_dir,
  pattern = "\\.json$", 
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

print(paste("Found", length(json_files), "JSON files."))
head(json_files)
```

## Generate a Report

This loop reads each file and attempts to summarize its contents. It handles errors gracefully (e.g., malformed syntax).


```{r}
#| label: generate-report
#| warning: false
#| message: false

# 1. Initialize results list
json_summary_list <- list()

# 2. Loop through files
for (file_path in json_files) {
  file_name <- basename(file_path)
  
  tryCatch({
    # Attempt to parse the JSON
    # simplifyVector = TRUE allows R to detect tables automatically
    json_data <- fromJSON(file_path, simplifyVector = TRUE)
    
    # Determine Class (Data Frame, List, or Vector)
    obj_class <- class(json_data)[1]
    
    # Determine Dimensions & Keys based on type
    if (is.data.frame(json_data)) {
      dims_str <- paste(dim(json_data), collapse = " x ")
      keys_str <- paste(head(colnames(json_data), 5), collapse = ", ")
      if (ncol(json_data) > 5) keys_str <- paste(keys_str, "...")
      
    } else if (is.list(json_data)) {
      dims_str <- paste(length(json_data), "elements")
      keys_str <- paste(head(names(json_data), 5), collapse = ", ")
      if (length(json_data) > 5) keys_str <- paste(keys_str, "...")
      
    } else {
      # Atomic vectors or arrays
      dims_str <- paste(length(json_data), "length")
      keys_str <- "(No Keys - Flat Array)"
    }
    
    # Store success result
    json_summary_list[[length(json_summary_list) + 1]] <- tibble(
      FileName = file_name,
      IsValid = TRUE,
      StructureType = obj_class,
      Dimensions = dims_str,
      TopLevelKeys = keys_str
    )
    
  }, error = function(e) {
    # Store error result
    json_summary_list[[length(json_summary_list) + 1]] <- tibble(
      FileName = file_name,
      IsValid = FALSE,
      StructureType = "Error",
      Dimensions = "NA",
      TopLevelKeys = paste("Parse Error:", e$message)
    )
  })
}

# 3. Combine and Print
json_report <- bind_rows(json_summary_list)

print("--- JSON Structure Report ---")
print(json_report)

# 4. Save to CSV
dir.create("Results/FileOpen_json", recursive = TRUE, showWarnings = FALSE)
write.csv(
  json_report,
  file = paste0("Results/FileOpen_json/JSON_Report_", Sys.Date(), ".csv"),
  row.names = FALSE
)

print("Report saved.")
```


## Curation Insights

The StructureType column is your primary indicator of the data's shape:

- data.frame: Good news! This JSON is tabular (an array of objects) and acts like a CSV. The Dimensions column tells you how many records it holds.

- list: This indicates a hierarchical or nested JSON. Use the TopLevelKeys to understand what categories of data it holds (e.g., metadata, data, citations).

- IsValid = FALSE: These files are corrupted or contain invalid syntax (like trailing commas). They will break any downstream analysis pipeline and should be flagged immediately.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server, in a batch job, or from the command line, here is a pure R script that performs the same process.


Download the **R Script:** **[`FileOpen_json_Script.R`](Scripts/FileOpen_json_Script.R)**


- **For clusters**

Alternatively, adapt the folling code script to match your cluster's environment and your data path, then submit it using `sbatch`.

```bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:10:00
#SBATCH --job-name=json_check

module load R 

DATA_DIR="/scratch/your_user/your_project/json_data"

Rscript check_json.R $DATA_DIR
```

