---
title: "SQLite Databases (.sqlite, .db)"
author: "Daniel Manrique-Castano"
date: "2025-01-05"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
bibliography: references.bib
params:
  target_dir: "data/Inspect_sqlite/"
---

## Overview

**SQLite** is the most widely deployed database engine in the world (see [SQLite.org](https://sqlite.org/about.html)). Unlike client-server database management systems (DBMS) like PostgreSQL or Oracle, an SQLite database is a single, self-contained disk file.

::: {.callout-note title="Curation Goal"}
From a curation perspective, SQLite is a highly sustainable format. [The Library of Congress](https://www.loc.gov/preservation/digital/formats/fdd/fdd000461.shtml) lists it as a preferred format for datasets due to its open specification and long-term stability. Our goal is to "unbox" the database to ensure all tables are documented and accessible.
:::

::: {.callout-warning title="Identifying Risks"}
"Self-Contained" does not imply "Self-Documenting." A curator receiving a `.db` file often faces a **Black Box**. Without a schema map, relationships between tables (Foreign Keys) can be lost, making the data contextually unusable.
:::

------------------------------------------------------------------------

## Setup

We use `RSQLite` [@RSQLite] to interact with the database files directly from R.

### R Packages

```{r}
# install.packages(c("tidyverse", "RSQLite", "DBI", "rstudioapi", "DT"))
```

### Load Libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(DBI)        
library(RSQLite)    
library(rstudioapi) 
library(DT)
```

## Select Target Directory

```{r}
#| label: select-directory

if (interactive() && .Platform$OS.type == "windows") {
  selected_dir <- rstudioapi::selectDirectory(caption = "Select SQLite Directory")
} else {
  selected_dir <- NULL
}

target_dir <- if (!is.null(selected_dir)) selected_dir else params$target_dir
print(paste("Analyzing directory:", target_dir))
```

## Inventory and Schema Extraction

This process iterates through all SQLite files, connects to them, and extracts the structure of every table found.

```{r}
#| label: database-inspection

db_files <- list.files(target_dir, pattern = "\\.(sqlite|db|sqlite3)$", full.names = TRUE, recursive = TRUE)

inspect_sqlite <- function(file_path) {
  con <- dbConnect(SQLite(), file_path)
  tables <- dbListTables(con)
  
  details <- map_dfr(tables, function(tbl) {
    # Get column info
    cols <- dbGetQuery(con, paste0("PRAGMA table_info(", tbl, ")"))
    # Get row count
    row_count <- dbGetQuery(con, paste0("SELECT COUNT(*) as count FROM ", tbl))$count
    
    tibble(
      file = basename(file_path),
      table_name = tbl,
      columns = paste(cols$name, collapse = ", "),
      row_count = row_count
    )
  })
  
  dbDisconnect(con)
  return(details)
}

if (length(db_files) > 0) {
  db_inventory <- map_dfr(db_files, inspect_sqlite)
  datatable(db_inventory, caption = "Inventory of Tables and Row Counts")
} else {
  message("No SQLite files found.")
}
```

## Archival Reporting

We save the database manifest as supplementary documentation to support long-term re-use of the dataset.

```{r}
#| label: save-results

if (exists("db_inventory")) {
  output_dir <- "Results/Inspect_sqlite"
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

  dir_label <- gsub("[^A-Za-z0-9_.-]", "_", basename(sub("[/\\\\]+$", "", target_dir)))
  timestamp <- format(Sys.Date(), "%Y%m%d")
  write_excel_csv(db_inventory, file.path(output_dir, paste0("Database_Manifest_", dir_label, "_", timestamp, ".csv")))
  
  message("Report saved to: ", output_dir)
}
```

## References

::: {#refs}
:::
