---
title: "SQLite Databases (.sqlite, .db)"
author: "Daniel Manrique-Castano"
date: "2025-01-05"
format: 
  html:
    toc: true
    code-fold: true
params:
  target_dir: "data/Inspect_sqlite/"
---

## 1. Overview

**SQLite** is the most widely deployed database engine in the world (see [SQLite.org](https://sqlite.org/about.html)). Different to client-server database management systems (DBMS) like PostgreSQL or Oracle, which often require a running server process, an SQLite database is a single, self-contained disk file. This portability makes it a preferred format for numerous applications such as:

-   **Research Software:** Storing simulation outputs or intermediate data processing steps.
-   **Digital Humanities:** Exporting datasets from web scrapers or network analysis tools.
-   **Mobile & Embedded Systems:** It is the standard local storage format for Android and iOS (see [documentation](https://developer.android.com/training/data-storage/sqlite?hl=es-419)\].

From a curation perspective, SQLite can be considered as a highly sustainable format. Indeed, [The Library of Congress](https://www.loc.gov/preservation/digital/formats/fdd/fdd000461.shtml) lists it as a preferred format for datasets due to its open specification, ubiquity, and long-term stability. However, it is important to consider that "Self-Contained" does not implies "Self-Documenting." A curator receiving a `.db` file often faces a "Black Box" containing an unknown number of tables and relationshipsm if not properly documented.

Because a single file can contain hundreds of separate "tables" (essentially spreadsheets linked together), curators cannot simply "open" it to check quality. Therefore, this notebook processes the following approach:

1.  **Inventory Tables:** List every table inside the container to ensure no data is overlooked.
2.  **Map the Schema:** Identify column names and data types to verify if the data matches the documentation.
3.  **Volume Check:** Count rows in each table to detect "Empty" tables (orphaned structures) or "Massive" tables that may require special handling.

------------------------------------------------------------------------

## Setup

We use `RSQLite` to interact with the database files directly from R.

### R Packages

```{r}
# install.packages(c("tidyverse", "RSQLite", "DBI", "rstudioapi"))
```

### Load Libraries

```{r}
#| message: false
#| 
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(DBI)        # Standard Database Interface
library(RSQLite)    # SQLite Engine
library(rstudioapi) # UI interaction
```
