---
title: "CSV Files"
author: "Natalie Williams"
date: "2025-10-10"
format: html
editor: visual
execute:
  echo: true
  warning: false
params:
  target_dir: "data/Inspect_csv/"
---

## Overview

This notebook provides a comprehensive toolkit for inspecting tabular data stored in **Comma Separated Values (CSV)** format.While seemingly simple, CSVs are the backbone of long-term data preservation due to their text-based, human-readable nature.

However, a "valid" CSV file is not necessarily a "usable" one. To ensure data reuse, datasets should adhere to **Tidy Data** principles, where variables are columns, observations are rows, and each observational unit forms a table [@wickham2014]. Common structural errors—such as "ragged" rows, empty header cells, or mixed data types—can severely hinder downstream analysis [@broman2018].

**Curation Objectives:**

1.  **Format Compliance:** Verify files use standard UTF-8 encoding to prevent character corruption (e.g., replacement characters like \`\`).
2.  **Structural Integrity:** Detect "dirty" data patterns such as duplicate rows or irregular shapes (non-rectangular data).
3.  **Content Profiling:** Automatically generate a statistical dictionary (Mean, Min, Max, Nulls) for every variable to validate the semantic meaning of the data.
4.  **Privacy Screening:** Scan for Personally Identifiable Information (PII) embedded within text fields.

------------------------------------------------------------------------

## Setup

We utilize `tidyverse` for data manipulation, `skimr` for detailed profiling, and `DT` for interactive result exploration.

### R Packages

If you do not have the required packages, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "rstudioapi", "skimr", "DT", "tools"))
```

### Load libraries

```{r}
#| label: load-libraries
#| message: false
#| warning: false

library(tidyverse)
library(skimr)
library(DT)
library(tools)
library(rstudioapi)
```

------------------------------------------------------------------------

## Select target directory

First, we select the directory to analyze and find `.csv` files within it.

```{r select-directory}
#| label: select-target-dir

# 1. Attempt interactive selection (Windows/RStudio)
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select Data Directory")
} else {
  selected_dir <- NULL
}

# 2. Fallback to parameter
target_dir <- if (!is.null(selected_dir)) selected_dir else params$target_dir

message("Analyzing directory: ", target_dir)
```

### Inventory Files

We locate all .csv files and extract basic filesystem metadata (size, modification date).

```{r get-files}
#| label: find-files

csv_files <- list.files(
  path = target_dir,
  pattern = "\\.csv$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

message("Found ", length(csv_files), " CSV files.")
```

------------------------------------------------------------------------

## Batch Health Check

This section assesses the "Archival Readiness" of the files. We perform a high-level scan to detect structural defects that might prevent the file from being opened or interpreted correctly [@broman2018].

Metrics Calculated:

-   Encoding: Non-UTF-8 encodings (like Latin-1 or Windows-1252) are a major source of data loss when moving between operating systems.

-   Completeness: The percentage of non-null cells. High missingness often indicates data entry errors or improper formatting (e.g., blank rows).

-   Duplication: Exact duplicate rows often signal version control errors (e.g., pasting data twice).

-   PII Flag: Detection of potential email addresses in text columns.

```{r initialize-df}
#| label: batch-health-check
#| message: false
#| warning: false

analyze_csv_health <- function(file_path) {
  
  fname <- basename(file_path)
  file_info <- file.info(file_path)
  
  # 1. Detect Encoding
  guess <- readr::guess_encoding(file_path, n_max = 1000)
  likely_encoding <- if (nrow(guess) > 0) guess$encoding[1] else "Unknown"
  
  tryCatch({
    # 2. Safe Read
    df <- read_csv(file_path, locale = locale(encoding = likely_encoding), 
                   show_col_types = FALSE, progress = FALSE)
    
    # 3. Structural Metrics
    n_rows <- nrow(df)
    n_cols <- ncol(df)
    n_missing <- sum(is.na(df))
    total_cells <- n_rows * n_cols
    pct_complete <- if (total_cells > 0) round(100 * (1 - n_missing / total_cells), 2) else 0
    n_duplicates <- sum(duplicated(df))
    
    # 4. Privacy Scan (PII - Email Regex)
    char_cols <- select(df, where(is.character))
    email_pattern <- "[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}"
    pii_found <- FALSE
    
    if (ncol(char_cols) > 0) {
      sample_size <- min(n_rows, 1000)
      pii_check <- char_cols %>%
        slice_head(n = sample_size) %>%
        summarise(across(everything(), ~ any(str_detect(., email_pattern), na.rm = TRUE)))
      pii_found <- any(unlist(pii_check))
    }
    
    tibble(
      FileName = fname,
      Size_MB = round(file_info$size / 1024^2, 2),
      Encoding = likely_encoding,
      Rows = n_rows,
      Cols = n_cols,
      Pct_Complete = pct_complete,
      Duplicate_Rows = n_duplicates,
      PII_Risk = pii_found,
      Status = "Success"
    )
    
  }, error = function(e) {
    tibble(
      FileName = fname,
      Size_MB = round(file_info$size / 1024^2, 2),
      Encoding = likely_encoding,
      Rows = NA, Cols = NA, Pct_Complete = NA, Duplicate_Rows = NA, PII_Risk = NA,
      Status = paste("Read Failed:", e$message)
    )
  })
}

if (length(csv_files) > 0) {
  health_report <- purrr::map_dfr(csv_files, analyze_csv_health)
  datatable(health_report, 
            caption = "Table 1: CSV Health Check Summary",
            options = list(scrollX = TRUE))
} else {
  message("No CSV files found.")
}
```

## Visualization: Structural Overview

We visualize the dimensions of the datasets. This helps identify outliers (e.g., empty files or unexpectedly small files in a time series).

```{r}
#| label: viz-structure
#| fig-cap: "Dataset Dimensions (Rows vs Columns)"
#| warning: false

ggplot(health_report, aes(x = Cols, y = Rows, size = Size_MB, color = Status)) +
  geom_point(alpha = 0.7) +
  scale_y_log10() + # Log scale for rows as they can vary wildly
  labs(
    title = "Dataset Dimensions",
    subtitle = "Log Scale on Y-Axis (Rows)",
    x = "Number of Columns",
    y = "Number of Rows (Log)"
  ) +
  theme_minimal()
```

## Detailed Batch Profiling

Here, we generate a "Master Data Dictionary" that characterizes every variable in the dataset. This allows curators to spot Type Inconsistency (e.g., a "Weight" column that is numeric in one file but character in another due to "N/A" strings).

-   Numeric: Mean, SD, quartiles, histograms (as text).

-   Character: Unique counts, whitespace checks.

-   Missing: NA counts per column.

```{r}
#| label: batch-profiling
#| message: false
#| warning: false

message("Generating detailed profiles...")

safe_skim <- function(file_path) {
  tryCatch({
    df <- read_csv(file_path, show_col_types = FALSE)
    skim(df) %>% 
      as_tibble() %>% 
      mutate(FileName = basename(file_path)) %>%
      select(FileName, everything())
  }, error = function(e) NULL)
}

if (length(csv_files) > 0) {
  full_profile_data <- map_dfr(csv_files, safe_skim)
  
  datatable(head(full_profile_data, 50), 
            caption = "Table 2: Master Data Dictionary (Preview)",
            options = list(scrollX = TRUE))
}
```

------------------------------------------------------------------------

## Save Results

We export two reports:

-   Health Check: The high-level summary of file validity and risks.

-   Detailed Profile: The granular statistics for every variable.

```{r write-files}
#| label: save-results

output_dir <- file.path("Results", "Inspect_CSV")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# File names
health_file <- file.path(output_dir, paste0("CSV_Health_Check_", Sys.Date(), ".csv"))
profile_file <- file.path(output_dir, paste0("CSV_Full_Profile_", Sys.Date(), ".csv"))

# Save
write_csv(health_report, health_file)
write_csv(full_profile_data, profile_file)

message("Reports saved:")
message("1. ", health_file)
message("2. ", profile_file)
```

------------------------------------------------------------------------

## Curation Insights

-   **Rectangularity:** If Read Failed occurs in Table 1, the file may have "ragged rows" (varying numbers of columns per row). This violates the Tidy Data principle that "each type of observational unit forms a table".

-   **Encoding Hygiene:** Files with encodings like ISO-8859-1 should be converted to UTF-8 to ensure long-term accessibility.

-   **Privacy:** If PII_Risk is TRUE, the file requires manual review. Publishing raw PII (like email addresses) is a severe ethical and legal violation.

## Additional Tools for Researchers

R is an excellent tool for data curation, but other tools are also available for cleaning messy tabular data.

-   **OpenRefine:** The industry standard for cleaning messy data. It excels at standardizing inconsistent text (e.g., "USA", "U.S.A.", "United States") via clustering algorithms [@dewilde2013] (https://openrefine.org).

-   **Frictionless Data (Table Schema):** A specification for defining a schema (types, constraints) for CSV files. It allows you to validate data automatically against a defined structure [\@fowler2018](https://frictionlessdata.io/).

-   csvkit / xsv: Command-line utilities for slicing, dicing, and analyzing massive CSV files that are too large to open in Excel or R (https://csvkit.readthedocs.io).

## Using the Non-Interactive R Script

For users who want to run this analysis on a server or from the command line, a non-interactive version of this script is available.

-   **R Script:** [**`Inspect_csv_Script.R`**](Scripts/Inspect_csv_Script.R)
-   **Example SLURM Submission Script:** [**`FileCkecks_csv.sh`**](Scrum/FileCkecks_csv.sh)

### How to Use

-   **For RStudio:**

1.  Download the `Inspect_csv_Script.R` script.
2.  Open in Rstudio and run the whole script. or run it from your terminal, passing the path to your data folder as an argument:

-   **For clusters**

Alternatively, adapt the `FileCkecks_csv.sh` script to match your cluster's environment and your data path, then submit it using `sbatch`.

```bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:30:00
#SBATCH --job-name=csv_check

# Load R module
module load R

# Define directories
DATA_DIR="/scratch/your_user/data_folder"
OUTPUT_DIR="/scratch/your_user/csv_results"

# Run Script
Rscript Inspect_CSV_Script.R $DATA_DIR $OUTPUT_DIR
```


## References

::: {#refs}
:::
