---
Me gusta. Procede. ---
title: "Jupyter Notebooks (.ipynb) Files"
author: "Daniel Manrique-Castano"
date: "2025-12-08"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
bibliography: references.bib
params:
  target_dir: "data/Inspect_ipynb/"
---

## Overview

Jupyter Notebooks are **Literate Programming**, a paradigm introduced by Donald Knuth (1984), where code and natural language documentation are interwoven to create a "computational narrative." In data curation, preserving this narrative is as important as preserving the raw data.

::: {.callout-note title="Curation Goal"}
Preserve the computational narrative. Our objective is to assess the structural integrity and documentation quality of notebooks to ensure they remain understandable and potentially reproducible for future researchers.
:::

::: {.callout-warning title="Preservation Risk"}
Notebooks are notoriously difficult to re-execute. Studies show that as few as 4% of notebooks produce identical results over time due to non-linear execution, hidden state, and undocumented software environments [@pimentel2019].
:::

**Key Curation Objectives:**

1.  **Format Obsolescence:** Checking `nbformat` versions to ensure compatibility.
2.  **Documentation Quality:** Calculating the ratio of Markdown to Code cells. A low ratio indicates poor explainability [@rule2018].
3.  **Dependency Awareness:** Scanning imports to identify the software environment required for re-execution.

------------------------------------------------------------------------

## Setup

We use `jsonlite` to parse the notebook structure and `tidyverse` for data manipulation.

### R Packages

If you do not have the required packages, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "jsonlite", "rstudioapi", "stringr"))
```

### Load libraries

```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(jsonlite)   # To parse .ipynb files
library(rstudioapi) # For directory selection
library(stringr)    # For dependency regex
```

## Select Target Directory

We select the folder containing the `.ipynb` files.

**Note:** If running interactively, a dialog box will appear. Otherwise, it defaults to the target_dir parameter.

```{r}
#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select Notebook Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))

```

## Find Jupyter Notebooks

We scan the directory for files ending in .ipynb.

```{r}
#| label: find-files

ipynb_files <- list.files(
  path = target_dir,
  pattern = "\\.ipynb$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

print(paste("Found", length(ipynb_files), "Jupyter Notebook files."))
head(ipynb_files)
```

## Robust Inventory & Extraction

We perform a "Fail-Safe" extraction. If a notebook is corrupted (invalid JSON), the script flags it as INVALID_JSON instead of crashing.

```{r}
#| label: generate-report
#| warning: false
#| message: false

message("Generating Notebook Report...")

# Regex for imports (Basic Python/R)
# Python: import numpy, from sklearn import ...
# R: library(tidyverse), require(data.table)
import_pattern_py <- "^\\s*(?:import|from)\\s+([a-zA-Z0-9_\\.]+)"
import_pattern_r <- "(?:library|require|p_load)\\s*\\(\\s*[\"']?([a-zA-Z0-9\\.]+)[\"']?\\s*\\)"

report <- purrr::map_dfr(ipynb_files, function(file_path) {
  
  fname <- basename(file_path)
  
  tryCatch({
    # Read the JSON structure
    nb_json <- jsonlite::fromJSON(file_path, flatten = FALSE)
    
    # SAFETY CHECK: Ensure it's a list (JSON object)
    if (!is.list(nb_json)) {
      stop("Parsed content is not a JSON object (is it a list?)")
    }

    # --- Metadata Extraction ---
    # Handle metadata being NULL or not a list
    metadata <- nb_json$metadata
    if (is.null(metadata)) {
        display_name <- "Unknown"
        language <- "Unknown"
    } else if (!is.list(metadata)) {
        display_name <- "Error: Metadata not a list"
        language <- "Unknown"
    } else {
        kernelspec <- metadata$kernelspec
        # Kernelspec check
        if (!is.null(kernelspec) && is.list(kernelspec)) {
             display_name <- if(!is.null(kernelspec$display_name)) kernelspec$display_name else "Unknown"
             language <- if(!is.null(kernelspec$language)) kernelspec$language else "Unknown"
        } else {
             display_name <- "Unknown"
             language <- "Unknown"
        }
    }

    format_version <- paste(nb_json$nbformat, nb_json$nbformat_minor, sep=".")
    
    # --- Cell Analysis ---
    cells <- nb_json$cells
    
    if (is.null(cells) || length(cells) == 0) {
      return(tibble(
        FileName = fname, Language = language, Kernel = display_name, NbFormat = format_version,
        TotalCells = 0, MarkdownCells = 0, CodeCells = 0,
        HasOutputs = FALSE, ErrorOutputs = 0, ExecOrderLinear = TRUE,
        Libraries = "", Status = "Empty"
      ))
    }
    
    # Counts
    num_cells <- nrow(cells)
    markdown_cells <- sum(cells$cell_type == "markdown")
    code_cells_df <- cells[cells$cell_type == "code", ]
    num_code_cells <- nrow(code_cells_df)
    
    # --- Code Cell Deep Dive ---
    has_output <- FALSE
    error_count <- 0
    exec_order_linear <- TRUE
    detected_libs <- character(0)
    
    if (num_code_cells > 0) {
      # 1. Check Outputs
      if (!is.null(code_cells_df$outputs)) {
        has_output <- sum(sapply(code_cells_df$outputs, length) > 0) > 0
        
        # Check for Errors
        all_outputs <- unlist(code_cells_df$outputs, recursive = FALSE)
        if (length(all_outputs) > 0) {
          # Safely extract output_type
          # Ensure x is a list/env before $ access
          output_types <- unlist(lapply(all_outputs, function(x) {
              if (is.list(x)) return(x$output_type) else return(NA)
          }))
          error_count <- sum(output_types == "error", na.rm = TRUE)
        }
      }
      
      # 2. Check Execution Order
      exec_counts <- unlist(code_cells_df$execution_count)
      if (!is.null(exec_counts) && length(exec_counts) > 1) {
        valid_counts <- exec_counts[!is.na(exec_counts)]
        if (length(valid_counts) > 1) {
          exec_order_linear <- !is.unsorted(valid_counts)
        }
      }
      
      # 3. Detect Libraries
      # Unlist all source lines directly to handle line-by-line regex anchors correctly
      # (code_cells_df$source is a list of character vectors)
      all_lines <- unlist(code_cells_df$source)
      
      if (length(all_lines) > 0) {
        if (tolower(language) == "python") {
           matches <- str_match(all_lines, import_pattern_py)
           # str_match returns matrix [full_match, group1]
           detected_libs <- unique(matches[,2])
           detected_libs <- detected_libs[!is.na(detected_libs)]
        } else if (tolower(language) == "r") {
           matches <- str_match(all_lines, import_pattern_r)
           detected_libs <- unique(matches[,2])
           detected_libs <- detected_libs[!is.na(detected_libs)]
        }
      }
    }
    
    tibble(
      FileName = fname,
      Language = language,
      Kernel = display_name,
      NbFormat = format_version,
      TotalCells = num_cells,
      MarkdownCells = markdown_cells,
      CodeCells = num_code_cells,
      doc_ratio = round(MarkdownCells / (CodeCells + 0.001), 2),
      HasOutputs = has_output,
      ErrorOutputs = error_count,
      ExecOrderLinear = exec_order_linear,
      Libraries = paste(detected_libs, collapse = ", "),
      Status = "Success"
    )
    
  }, error = function(e) {
    tibble(
      FileName = fname, Language = "ERROR", Kernel = "", NbFormat = "",
      TotalCells = NA, MarkdownCells = NA, CodeCells = NA,
      HasOutputs = NA, ErrorOutputs = NA, ExecOrderLinear = NA,
      Libraries = "", Status = paste("Failed:", e$message)
    )
  })
})

# Display preview
print("--- Jupyter Notebook Report Preview ---")
print(head(report))
```

## Visualization: Documentation Quality

We visualize the ratio of Markdown (Text) to Code. A ratio near 0 implies a script with no explanation. A ratio \> 1 implies a well-documented narrative.

```{r}
#| label: visual-doc
#| fig-cap: "Documentation Quality: Markdown vs. Code Cells"

ggplot(report, aes(x = reorder(FileName, doc_ratio), y = doc_ratio)) +
  geom_col(aes(fill = doc_ratio > 0.5)) +
  coord_flip() +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("FALSE" = "#D55E00", "TRUE" = "#009E73"), name = "Well Documented?") +
  labs(
    title = "Notebook Documentation Ratio",
    subtitle = "Ratio of Markdown Cells to Code Cells (Target > 0.5)",
    x = "Notebook",
    y = "Markdown / Code Ratio"
  ) +
  theme_minimal()
```

## Save Results

Save the report to a CSV file for review.

```{r}
#| label: save-results

output_dir <- "Results/Inspect_ipynb"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

output_file <- file.path(output_dir, paste0("Jupyter_Report_", format(Sys.Date(), "%Y%m%d"), ".csv"))

write.csv(report, output_file, row.names = FALSE)
print(paste("Report saved to:", output_file))
```

## Curation Insights

Use the generated CSV to perform these checks:

-   **Residual Outputs (has_outputs):** If TRUE, the notebook was saved with graphs, tables, or text logs visible. For archival purposes, it is generally recommended to clear all outputs (using nbstripout or the Jupyter interface). This reduces file size significantly and prevents the accidental leakage of sensitive data (PII) that might be displayed in a table output.

-   **Kernel Consistency:** Check for mismatches, such as a notebook using a python2 kernel when the project standard is python3, or a file named Analysis_R.ipynb that actually runs on a python kernel. These inconsistencies often break reproducibility. Ensure the kernel metadata matches the environment provided in requirements.txt or environment.yml.

-   **Format Obsolescence (nbformat):** The standard format version is currently 4.x.If you find notebooks with nbformat \< 4 (created before \~2015), they may not render correctly in modern JupyterLab. Convert them using nbconvert or open and re-save them in a modern interface.

-   **JSON Validity:** Jupyter notebooks are just JSON text files. If the script fails to parse a file, it is likely a "Corrupt Notebook" (invalid JSON structure). Attempt to fix the JSON syntax manually in a text editor, or check if the file was truncated during a previous transfer.

## Additional Tools & Resources

While R parses the metadata efficiently, these specialized tools are essential for managing the notebook lifecycle and reproducibility.

-   **nbstripout:** Is a [command-line tool](https://github.com/kynan/nbstripout) that automatically strips output cells from notebooks. It is essential for version control (Git) and preparing "clean" copies for the archive.

-   **Jupytext:** This [utility](https://jupytext.readthedocs.io/en/latest/) converts .ipynb files into paired plain-text scripts (Markdown or Python/R scripts). This allows you to diff changes easily and preserves the code even if the notebook JSON corrupts.

-   **nbconvert:** The official [tool](https://nbconvert.readthedocs.io/en/latest/) to convert notebooks into static formats like HTML, PDF, or LaTeX. It is recommended to generate a static HTML or PDF copy of the notebook with outputs executed to ensure the visual results are preserved even if the code stops working in the future.

-   **Binder / Repo2Docker:** This [utility](https://mybinder.org/) allows you to build the exact software environment required to run the notebook.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server (HPC), in a batch job, or from the command line, here is the pure R script version.

### The `Inspect_ipynb_Script.R` Script

Download the **R Script:** [**`Inspect_ipynb_Script.R`**](Scripts/Inspect_ipynb_Script.R)

### Example HPC Submission Script (`Inspect_ipynb_submit.sh`)

``` bash
#!/bin/bash
#SBATCH --job-name=ipynb_inspect
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:15:00
#SBATCH --mem=4G
#SBATCH --output=logs/inspect_ipynb_%j.log

# Load R module (Adjust based on your cluster's module name)
module load R

# Define target directory containing notebooks
TARGET_DIR="/scratch/user/project_data/notebooks"

# Ensure the Results directory exists (optional, script handles it too)
mkdir -p Results/Inspect_ipynb

# Run the R script
echo "Starting Notebook Inspection on $TARGET_DIR"
Rscript Scripts/Inspect_ipynb_Script.R "$TARGET_DIR"
```

## References

::: {#refs}
:::