---
title: "GeoPackage (.gpkg) Files"
format:
  html:
    toc: true
    toc-location: left
    code-fold: true
bibliography: references.bib
params:
  target_dir: "data/Inspect_gpkg"
---

## Overview

This notebook inspects **GeoPackage (`.gpkg`)** files—an open, standards-based, and self-describing format for geospatial information.

::: {.callout-note title="Curation Goal"}
Validate structural and spatial integrity. Our objective is to inventory the layers within the GeoPackage database, verify Coordinate Reference Systems (CRS), and ensure all geometries conform to the Simple Features standard (ISO 19125).
:::

::: {.callout-warning title="Identifying Risks"}
GeoPackage eliminates the multi-file fragility of Shapefiles. However, a missing CRS is a primary cause of non-reproducibility in spatial analysis, as it prevents data from correctly aligning with other map layers [@obe2021].
:::

**Curation Objectives:**

1.  **CRS Validation:** Confirm every layer has a defined coordinate system.
2.  **Layer Inventory:** Map all datasets stored within the container.
3.  **Geometry Profiling:** Determine topological types and feature counts.
4.  **Standard Compliance:** Detect invalid geometries (e.g., self-intersecting lines).

## Setup

We use the **`sf`** package [@sf] for structural analysis, which provides an R implementation of the Simple Features standard [@pebesma2018] and `tmap`[@tmap] for thematic mapping. For visual inspection, we use **`leaflet`** [@leaflet].

### R Packages

If you don't have these packages installed, run this command once in your R console:

```{r}
# install.packages(c("sf", "tidyverse", "leaflet", "tmap", "tools", "DT"))
```

### Load Libraries

This chunk loads the necessary libraries for the session.

```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(sf)      
library(tmap) 
library(leaflet)
library(tools)   
library(DT)     
library(rstudioapi)

# Validation check for system dependencies (GDAL/GEOS)
if (!require("sf", quietly = TRUE)) {
  stop("The 'sf' package is missing. Please install it to proceed.")
}
```

## Batch Inventory

A GeoPackage is a database that may contain multiple datasets. We use the st_layers() function in the sf package to query the SQLite metadata tables without loading the actual geometry. This allows for rapid scanning of large archives.

### Select Target Directory

Select the directory containing the .gpkg files.

```{r}
#| label: select-target-dir

# Hybrid Selection Logic (Interactive vs Batch)
if (interactive() && requireNamespace("rstudioapi", quietly = TRUE)) {
  message("Running in interactive mode. Please select a directory.")
  selected_dir <- rstudioapi::selectDirectory(caption = "Select GeoPackage Directory")
  
  if (is.null(selected_dir)) stop("No directory selected.")
  target_dir <- selected_dir
  
} else {
  target_dir <- params$target_dir
}

if (!dir.exists(target_dir)) stop(paste("Directory not found:", target_dir))

message("Analyzing directory: ", target_dir)
```

## Inventory of Files

```{r}
#| label: find-files

gpkg_files <- list.files(
  path = target_dir,
  pattern = "\\.gpkg$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

message("Found ", length(gpkg_files), " GeoPackage files.")
```

## Batch Layer Analysis

A GeoPackage is technically a relational database (SQLite). Therefore, we cannot simply "read the file"; we must query it to list its layers (tables). The st_layers() function retrieves metadata without loading the heavy geometry data into memory.

**Information Extracted:**

-   **Layer Name:** The identifier within the database.

-   **Geometry:** The topological type (e.g., Polygon, MultiPoint).

-   **CRS:** The coordinate reference system ID (e.g., EPSG:4326).

-   **Features:** The number of rows (spatial objects).

```{r}
#| label: define-analysis-function

analyze_gpkg_layers <- function(file_path) {
  
  fname <- basename(file_path)
  file_size_mb <- round(file.size(file_path) / 1024^2, 2)
  
  tryCatch({
    # 1. List Layers (st_layers creates a summary without reading full geometry)
    layer_info <- st_layers(file_path)
    
    # 2. Extract Metadata for each layer
    # If file has 0 layers, we return a blank record
    if (length(layer_info$name) == 0) {
      return(tibble(
        FileName = fname,
        Size_MB = file_size_mb,
        Layer_Name = "EMPTY",
        Geometry_Type = NA,
        CRS = NA,
        Features = 0,
        Fields = 0,
        Status = "Empty GPKG"
      ))
    }
    
    # Map over layers to get details
    purrr::map_dfr(seq_along(layer_info$name), function(i) {
      
      crs_val <- layer_info$crs[[i]]
      # Extract the CRS input code (e.g. "EPSG:4326")
      crs_code <- if (!is.na(crs_val)) crs_val$input else "Missing"
      
      tibble(
        FileName = fname,
        Size_MB = file_size_mb,
        Layer_Name = layer_info$name[i],
        Geometry_Type = paste(layer_info$geomtype[[i]], collapse = ", "),
        CRS = as.character(crs_code),
        Features = layer_info$features[i],
        Fields = layer_info$fields[i],
        Status = "Success"
      )
    })
    
  }, error = function(e) {
    tibble(
      FileName = fname,
      Size_MB = file_size_mb,
      Layer_Name = NA, Geometry_Type = NA, CRS = NA, Features = NA, Fields = NA,
      Status = paste("Read Failed:", e$message)
    )
  })
}
```

### Execute Analysis

```{r}
#| label: run-analysis
#| message: false
#| warning: false

message("Inspecting GeoPackage layers...")
layer_report <- purrr::map_dfr(gpkg_files, analyze_gpkg_layers)

# Display interactive table
datatable(layer_report, 
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "Table 1: GeoPackage Layer Inventory")
```

### Visualization

Visualizing spatial data is the most effective way to detect Projection Errors.

If a layer lacks a CRS or has an incorrect one, points may appear in "Null Island" (0,0 coordinates) or be distorted. We use tmap to plot the first available layer as a diagnostic check.

```{r}
#| label: viz-map
#| fig-cap: "Sample Map: First Layer of First File"
#| warning: false

# 1. Select the first valid layer
valid_layers <- layer_report %>% filter(Status == "Success", Features > 0)

if (nrow(valid_layers) > 0) {
  
  target_file <- file.path(target_dir, valid_layers$FileName[1])
  target_layer <- valid_layers$Layer_Name[1]
  
  message("Visualizing: ", target_layer, " from ", basename(target_file))
  
  # Read the actual data
  geo_data <- st_read(target_file, layer = target_layer, quiet = TRUE)
  
  # Plot using tmap in static "plot" mode
  tmap_mode("plot")
  
  tm_shape(geo_data) +
    tm_borders(col = "black", lwd = 1) +
    tm_fill(col = "#4C78A8", alpha = 0.6) +
    tm_layout(
      title = paste0("Layer: ", target_layer),
      frame = TRUE
    )
    
} else {
  message("No valid spatial layers found to visualize.")
}
```

### Save Results

```{r}
#| label: save-results

output_dir <- file.path("Results", "Inspect_GPKG")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

dir_label <- gsub("[^A-Za-z0-9_.-]", "_", basename(sub("[/\\\\]+$", "", target_dir)))
output_file <- file.path(output_dir, paste0("GPKG_Layer_Report_", dir_label, "_", Sys.Date(), ".csv"))
write_excel_csv(layer_report, output_file)

message("Report saved to: ", output_file)
```

## Interactive Ground-Truthing (leaflet)

While batch scanning validates structure, Visual Inspection validates content. It is the primary method for detecting "logical errors," such as a dataset labeled "Canada" that visually appears in "Australia" due to projection issues [@explorat2024]. We inspect a single file below. This step uses leaflet to overlay the data on a reference base map (OpenStreetMap).

**Note:** Leaflet requires the WGS84 coordinate system (EPSG:4326). We perform an on-the-fly transformation to ensure the data renders correctly.

### Select File for Inspection

```{r}
#| label: select-single-file

# Default to the first valid file found, or use the parameter
if (exists("layer_report") && nrow(layer_report) > 0) {
  target_file <- file.path(target_dir, layer_report$FileName[1])
} else {
  target_file <- params$target_file
}

message("Inspecting file: ", basename(target_file))
```

### Layer Table of Contents (st_layers)

Before reading data, we examine the table of contents. This confirms the schema without loading potentially massive datasets into memory.

```{r}
#| label: inspect-layers
st_layers(target_file)
```

### Statistical Overview (summary)

We load a specific layer to check for value distribution. Anomalies in the summary() output (e.g., a "Population" column with negative values) often indicate data corruption.

```{r}
#| label: read-and-summary

# Read the first layer by default
gpkg_data <- st_read(target_file, quiet = TRUE)

# Statistical Summary
summary(gpkg_data)
```

### Render the Interactive Map

The code below adapts to the geometry type (Point, Line, or Polygon) to prevent rendering errors. It also performs an on-the-fly transformation to WGS84 (EPSG:4326), which is required for web mapping.

```{r}
#| label: interactive-map
#| fig-cap: "Figure 1: Interactive verification map. Use zoom to check alignment with the base map."

# 1. Read the first layer
gpkg_data <- st_read(target_file, quiet = TRUE)

# 2. Transform to WGS84 for Leaflet
gpkg_data_wgs84 <- st_transform(gpkg_data, crs = 4326)

# 3. Detect Geometry Type
geom_type <- as.character(st_geometry_type(gpkg_data_wgs84)[1])

# 4. Initialize Base Map
map <- leaflet(gpkg_data_wgs84) %>%
  addProviderTiles(providers$CartoDB.Positron)

# 5. Conditionally Add Layers
if (grepl("POLYGON", geom_type)) {
  map <- map %>%
    addPolygons(
      color = "#4C78A8", weight = 1, fillOpacity = 0.5,
      popup = ~paste("Feature ID:", seq_len(nrow(gpkg_data_wgs84)))
    )
} else if (grepl("POINT", geom_type)) {
  map <- map %>%
    addCircleMarkers(
      radius = 5, color = "#4C78A8", stroke = FALSE, fillOpacity = 0.8,
      popup = ~paste("Feature ID:", seq_len(nrow(gpkg_data_wgs84)))
    )
} else if (grepl("LINE", geom_type)) {
  map <- map %>%
    addPolylines(
      color = "#4C78A8", weight = 2, opacity = 0.8,
      popup = ~paste("Feature ID:", seq_len(nrow(gpkg_data_wgs84)))
    )
}

map
```

## Curation Insights

Based on the analysis above, curators should look for the following risks:

-   **Missing CRS:** If the CRS column in Table 1 is "Missing" or NA, the file is technically incomplete. Without a CRS, the data cannot be reliably combined with other datasets.

-   **Geometry Validity:** The sf package enforces strict topological rules (e.g., polygon lines cannot cross themselves). If st_read throws an error, the geometry is likely invalid and may require repair in QGIS.

-   **Empty Layers:** Layers with Features = 0 may be artifacts of a failed export process and should be flagged for removal or documentation. Confirm with the depositor that these featureless layers are correct.

## Additional Tools

While R is excellent for detection, other tools are often better suited for repairing geospatial data. Curators should be aware of:

-   **QGIS (Desktop GIS Software):** is a free, open-source, and powerful Geographic Information System. It's the industry standard for working with geospatial data outside of an programming environment (https://qgis.org).

-   **DB Browser for SQLite (Database Tool):** Since a `.gpkg` file is fundamentally an SQLite database, you can use a database browser to inspect its raw structure (https://sqlitebrowser.org/).

-   **GDAL (Command Line):** The `ogr2ogr` command-line utility is the gold standard for batch converting formats (e.g., Shapefile to GeoPackage) and reprojecting data (https://gdal.org) [@warmerdam].

-   **Mapshaper:** A web-based tool excellent for simplifying overly complex geometries (e.g., reducing the file size of a detailed coastline) while preserving topology (https://mapshaper.org).

## Using the Non-Interactive R Script

For users who want to run this analysis on a server, in a batch job, or from the command line, here is a pure R script that performs the same process.

### The `Inspect_gpkg_Script.R` Script

Download the **R Script:** [**`Inspect_gpkg_Script.R`**](Scripts/Inspect_gpkg_Script.R)

### Example HPC Submission Script (`Inspect_gpkg_submit.sh`)

``` bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:20:00
#SBATCH --job-name=gpkg_check

# Load R Module
module load R

# IMPORTANT: sf requires GDAL/GEOS. 
# On many clusters, these are separate modules that must be loaded explicitly.
# Check your cluster documentation (e.g., 'module avail gdal')
module load gdal 
module load geos
module load proj
module load R

# Define data directory
DATA_DIR="/scratch/your_user/spatial_data"
OUTPUT_DIR="/scratch/your_user/spatial_results"

# Run Script
Rscript Inspect_GPKG_Script.R $DATA_DIR $OUTPUT_DIR
```

## References

::: {#refs}
:::
