---
title: "Text (.txt) and Markdown (.md) Files"
author: "Data Curation Team"
date: "2025-12-18"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: cosmo
bibliography: references.bib
params:
  target_dir: "data/Inspect_Text/"
---

## Overview

Text files are the simplest form of documentation. However, they are often prone to encoding and structural issues that impede interoperability.

::: {.callout-note title="Curation Goal"}
Ensure universal readability and "Archival Readiness" of text documents. Our objective is to validate UTF-8 encoding, identify "invisible" characters (BOM), and normalize line endings to ensure documents remain readable across all operating systems.
:::

::: {.callout-warning title="Preservation Risk"}
Character corruption ("Mojibake") caused by legacy encodings (e.g., Windows-1252) and "link rot" from broken external URLs are the primary threats to the long-term usability of plain text documentation.
:::

**Key Curation Objectives:**

1.  **Encoding Validation:** Detect character encoding and ensure compliance with the UTF-8 standard.
2.  **Structural Integrity:** Identify Byte Order Marks (BOM) and mixed line endings (CRLF/LF).
3.  **Link & Security Scan:** Extract external URLs and scan for accidental PII leaks (e.g., email addresses).

------------------------------------------------------------------------

## Setup

We use `readr` for encoding detection and `stringr` for link extraction.

### R Packages

```{r}
# install.packages(c("tidyverse", "readr", "rstudioapi", "stringr"))
```

### Load libraries

```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(readr)      # For encoding guessing
library(stringr)    # For Regex (Links/Emails)
library(rstudioapi) # For directory selection
```

## Select Target Directory

```{r}
#| label: select-target-dir

if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select Text Directory")
} else {
  selected_dir <- NULL
}

if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))
```

## Inventory and Inspection

We scan for .txt, .md, .csv, and .rmd files. The inspection extracts encoding confidence, checks for the hidden BOM, identifies line endings, and scans for PII (Emails).

```{r}
#| label: extraction-logic
#| warning: false
#| message: false

message("Generating Text Report...")

text_files <- list.files(
  path = target_dir,
  pattern = "\\.(txt|md|csv|rmd)$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

# Regex Patterns
url_pattern <- "http[s]?://(?:[a-zA-Z]|[0-9]|[$-_@.&+]|[!*\\(\\),]|(?:%[0-9a-fA-F][0-9a-fA-F]))+"
email_pattern <- "[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}"

report <- purrr::map_dfr(text_files, function(file_path) {
  
  fname <- basename(file_path)
  
  tryCatch({
    # 1. BOM Detection (Read raw bytes)
    con <- file(file_path, "rb")
    bytes <- readBin(con, "raw", n = 4)
    close(con)
    
    # Check for UTF-8 BOM (EF BB BF)
    has_bom <- identical(bytes[1:3], as.raw(c(0xef, 0xbb, 0xbf)))
    
    # 2. Encoding Guess
    guess <- readr::guess_encoding(file_path, n_max = 1000)[1, ]
    encoding <- if (!is.na(guess$encoding)) guess$encoding else "Unknown"
    confidence <- if (!is.na(guess$confidence)) guess$confidence else 0
    
    # 3. Content Analysis (Read text)
    # Read safely with UTF-8 fallback
    content_lines <- readLines(file_path, warn = FALSE)
    full_text <- paste(content_lines, collapse = "\n")
    
    # 4. Line Ending Detection
    # We read raw again to distinguish \r\n vs \n (readLines normalizes them)
    raw_text <- readChar(file_path, nchars = 2000, useBytes = TRUE)
    eol_type <- "Unknown"
    if (grepl("\r\n", raw_text)) {
      eol_type <- "Windows (CRLF)"
    } else if (grepl("\n", raw_text)) {
      eol_type <- "Unix (LF)"
    } else if (grepl("\r", raw_text)) {
      eol_type <- "Classic Mac (CR)"
    }
    
    # 5. Extract Artifacts
    urls <- str_extract_all(full_text, url_pattern)[[1]]
    emails <- str_extract_all(full_text, email_pattern)[[1]]
    
    example_links <- paste(head(unique(urls), 3), collapse = ", ")
    
    tibble(
      FileName = fname,
      Encoding = encoding,
      Confidence = confidence,
      HasBOM = has_bom,
      LineEndings = eol_type,
      LineCount = length(content_lines),
      URL_Count = length(urls),
      Email_Count = length(unique(emails)),
      Example_Links = substr(example_links, 1, 100),
      Status = "Success"
    )
    
  }, error = function(e) {
    tibble(
      FileName = fname, Encoding = NA, Confidence = NA, HasBOM = NA,
      LineEndings = NA, LineCount = NA, URL_Count = NA, Email_Count = NA,
      Example_Links = NA, Status = paste("Failed:", e$message)
    )
  })
})

# Display preview
print("--- Text Report Preview ---")
head(report)
```

## Visualization

We can visualize the distribution of detected encodings. Ideally, the repository should be 100% UTF-8 (or ASCII). Any "ISO-8859" or "Windows-1252" files are candidates for remediation.

```{r}
#| label: visual-text
#| fig-cap: "Distribution of File Encodings"

if (nrow(report) > 0) {
  ggplot(report %>% filter(Status == "Success"), aes(x = Encoding, fill = Encoding)) +
    geom_bar() +
    labs(
      title = "Text File Encodings",
      subtitle = "Archival Standard: UTF-8 / ASCII",
      x = "Detected Encoding",
      y = "File Count"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
}
```

## Save Results

```{r}
#| label: save-results

output_dir <- "Results/Inspect_Text"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

output_file <- file.path(output_dir, paste0("Text_Report_", format(Sys.Date(), "%Y%m%d"), ".csv"))

write.csv(report, output_file, row.names = FALSE)
print(paste("Report saved to:", output_file))
```

## Curation Insights

Use the generated CSV to perform these checks:

-   **PII Check (Email_Count \> 0):** Text files (especially Readmes) often contain developer contact info.Verify if these emails are personal (e.g., gmail.com) or professional.

-   **Encoding (Encoding != UTF-8):** Legacy files (Windows-1252) or other encodings may display corrupted characters on the web. It is recommended to convert them to UTF-8 using procedures like `iconv` (see below).

-   **BOM (HasBOM = TRUE):** The Byte Order Mark (BOM) is often unnecessary for UTF-8 and can break some scripts (e.g., shebang lines in bash). Curators can remove the BOM if the file is intended for code execution.

## Additional Tools

-   **iconv:** The standard [command-line tool](https://pubs.opengroup.org/onlinepubs/007904975/functions/iconv.html) for converting text encodings (e.g., iconv -f WINDOWS-1252 -t UTF-8 in.txt \> out.txt).

-   **dos2unix:** A [tool](https://linux.die.net/man/1/dos2unix) to normalize line endings (converting Windows CRLF to Unix LF). This may be useful for ensuring scripts run correctly on Linux clusters.

-   **Internet Archive Wayback Machine:** Use this [website](https://web.archive.org/) to find live versions of broken URLs.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server (HPC), in a batch job, or from the command line, here is the pure R script version.

Download the **R Script:** [**`Inspect_Text_Script.R`**](Scripts/Inspect_Text_Script.R)

### Example HPC Submission Script (`Inspect_Text_submit.sh`)

``` bash
#!/bin/bash
#SBATCH --job-name=text_check
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:15:00
#SBATCH --mem=4G
#SBATCH --output=logs/text_check_%j.log

module load R

# Define target directory
TARGET_DIR="/scratch/user/project_data/docs"

# Prepare folders
mkdir -p Results/Inspect_Text
mkdir -p logs

# Run
echo "Starting Text Inspection on $TARGET_DIR"
Rscript Inspect_Text_Script.R "$TARGET_DIR"
```

## References

::: {#refs}
:::
