---
title: "Inspection of Text (.txt) and Markdown (.md) Files"
author: "Data Curation Team"
date: "2025-12-18"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: cosmo
params:
  target_dir: "data/Inspect_Text/"
---

## Overview

Text files are the simplest form of documentation. Howver, they are often prone to **Encoding** issues, specially when dealing with other languages or symbols other then English. The archival standard for text files is **UTF-8** (defined in ISO/IEC 10646). Legacy encodings (like **Windows-1252**) map bytes to characters differently. If a file saved in Windows-1252 is read as UTF-8, characters like "é" turn into garbage symbols ("Mojibake"), rendering the data unreadable (see https://www.unicode.org/versions/Unicode15.0.0/).

Some editors add a hidden signature (`U+FEFF`, BOM) at the start of UTF-8 files. This "invisible" character often breaks scripts and parsers. Also, ***Line Endings (EOL):** Windows uses `CRLF` (`\r\n`), while Linux/Unix uses `LF` (`\n`). "Mixed" line endings often cause version control conflicts. Text documents often include external resources via URLs, which are often found to be broken, and are common places for accidental PII leaks, such as personal email addresses.

**Key Curation Objectives:**

1.  **Encoding:** Detect character encoding. Archival standards require **UTF-8**. Legacy encodings (like Windows-1252) cause "mojibake" on other systems.

2.  **Link Validation:** Extract HTTP/HTTPS links to check if external references are valid.

------------------------------------------------------------------------

## Setup

We use `readr` for encoding detection and `stringr` for link extraction.

### R Packages

```{r}
# install.packages(c("tidyverse", "readr", "rstudioapi", "stringr"))
```

### Load libraries

```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(readr)      # For encoding guessing
library(stringr)    # For Regex (Links/Emails)
library(rstudioapi) # For directory selection
```

## Select Target Directory

```{r}
#| label: select-target-dir

if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select Text Directory")
} else {
  selected_dir <- NULL
}

if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))
```

## Inventory and Inspection

We scan for .txt, .md, .csv, and .rmd files. The inspection extracts encoding confidence, checks for the hidden BOM, identifies line endings, and scans for PII (Emails).

```{r}
#| label: extraction-logic
#| warning: false
#| message: false

message("Generating Text Report...")

text_files <- list.files(
  path = target_dir,
  pattern = "\\.(txt|md|csv|rmd)$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

# Regex Patterns
url_pattern <- "http[s]?://(?:[a-zA-Z]|[0-9]|[$-_@.&+]|[!*\\(\\),]|(?:%[0-9a-fA-F][0-9a-fA-F]))+"
email_pattern <- "[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}"

report <- purrr::map_dfr(text_files, function(file_path) {
  
  fname <- basename(file_path)
  
  tryCatch({
    # 1. BOM Detection (Read raw bytes)
    con <- file(file_path, "rb")
    bytes <- readBin(con, "raw", n = 4)
    close(con)
    
    # Check for UTF-8 BOM (EF BB BF)
    has_bom <- identical(bytes[1:3], as.raw(c(0xef, 0xbb, 0xbf)))
    
    # 2. Encoding Guess
    guess <- readr::guess_encoding(file_path, n_max = 1000)[1, ]
    encoding <- if (!is.na(guess$encoding)) guess$encoding else "Unknown"
    confidence <- if (!is.na(guess$confidence)) guess$confidence else 0
    
    # 3. Content Analysis (Read text)
    # Read safely with UTF-8 fallback
    content_lines <- readLines(file_path, warn = FALSE)
    full_text <- paste(content_lines, collapse = "\n")
    
    # 4. Line Ending Detection
    # We read raw again to distinguish \r\n vs \n (readLines normalizes them)
    raw_text <- readChar(file_path, nchars = 2000, useBytes = TRUE)
    eol_type <- "Unknown"
    if (grepl("\r\n", raw_text)) {
      eol_type <- "Windows (CRLF)"
    } else if (grepl("\n", raw_text)) {
      eol_type <- "Unix (LF)"
    } else if (grepl("\r", raw_text)) {
      eol_type <- "Classic Mac (CR)"
    }
    
    # 5. Extract Artifacts
    urls <- str_extract_all(full_text, url_pattern)[[1]]
    emails <- str_extract_all(full_text, email_pattern)[[1]]
    
    example_links <- paste(head(unique(urls), 3), collapse = ", ")
    
    tibble(
      FileName = fname,
      Encoding = encoding,
      Confidence = confidence,
      HasBOM = has_bom,
      LineEndings = eol_type,
      LineCount = length(content_lines),
      URL_Count = length(urls),
      Email_Count = length(unique(emails)),
      Example_Links = substr(example_links, 1, 100),
      Status = "Success"
    )
    
  }, error = function(e) {
    tibble(
      FileName = fname, Encoding = NA, Confidence = NA, HasBOM = NA,
      LineEndings = NA, LineCount = NA, URL_Count = NA, Email_Count = NA,
      Example_Links = NA, Status = paste("Failed:", e$message)
    )
  })
})

# Display preview
print("--- Text Report Preview ---")
head(report)
```

## Visualization

We can visualize the distribution of detected encodings. Ideally, the repository should be 100% UTF-8 (or ASCII). Any "ISO-8859" or "Windows-1252" files are candidates for remediation.

```{r}
#| label: visual-text
#| fig-cap: "Distribution of File Encodings"

if (nrow(report) > 0) {
  ggplot(report %>% filter(Status == "Success"), aes(x = Encoding, fill = Encoding)) +
    geom_bar() +
    labs(
      title = "Text File Encodings",
      subtitle = "Archival Standard: UTF-8 / ASCII",
      x = "Detected Encoding",
      y = "File Count"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
}
```

## Save Results

```{r}
#| label: save-results

output_dir <- "Results/Inspect_Text"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

output_file <- file.path(output_dir, paste0("Text_Report_", format(Sys.Date(), "%Y%m%d"), ".csv"))

write.csv(report, output_file, row.names = FALSE)
print(paste("Report saved to:", output_file))
```

## Curation Insights

Use the generated CSV to perform these checks:

- **PII Check (Email_Count > 0):** Text files (especially Readmes) often contain developer contact info.Verify if these emails are personal (e.g., gmail.com) or professional. 

- **Encoding (Encoding != UTF-8):** Legacy files (Windows-1252) or other encodings may display corrupted characters on the web. It is recommended to convert them to UTF-8 using procedures like `iconv` (see below).

- **BOM (HasBOM = TRUE):** The Byte Order Mark (BOM) is often unnecessary for UTF-8 and can break some scripts (e.g., shebang lines in bash). Curators can remove the BOM if the file is intended for code execution.

## Additional Tools

- **iconv:** The standard [command-line tool](https://pubs.opengroup.org/onlinepubs/007904975/functions/iconv.html) for converting text encodings (e.g., iconv -f WINDOWS-1252 -t UTF-8 in.txt > out.txt).

- **dos2unix:** A [tool](https://linux.die.net/man/1/dos2unix) to normalize line endings (converting Windows CRLF to Unix LF). This may be useful for ensuring scripts run correctly on Linux clusters.

- **Internet Archive Wayback Machine:** Use this [website](https://web.archive.org/) to find live versions of broken URLs.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server (HPC), in a batch job, or from the command line, here is the pure R script version.

Download the **R Script:** [**`Inspect_Text_Script.R`**](Scripts/Inspect_Text_Script.R)

### Example HPC Submission Script (`Inspect_Text_submit.sh`)

```bash
#!/bin/bash
#SBATCH --job-name=text_check
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:15:00
#SBATCH --mem=4G
#SBATCH --output=logs/text_check_%j.log

module load R

# Define target directory
TARGET_DIR="/scratch/user/project_data/docs"

# Prepare folders
mkdir -p Results/Inspect_Text
mkdir -p logs

# Run
echo "Starting Text Inspection on $TARGET_DIR"
Rscript Inspect_Text_Script.R "$TARGET_DIR"
```



## References

::: {#refs}
:::
