---
title: "XML Files"
author: "Daniel Manrique-Castano"
date: "2025-12-18"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: cosmo
bibliography: references.bib
params:
  target_dir: "data/Inspect_xml/"
---

## Overview

XML (Extensible Markup Language) is a standard for storing structured data. In curation, we distinguish between two states. A Well-Formed file follows basic syntax rules (e.g., every start tag `<tag>` has an end tag `</tag>`). If this is not well-formed, the file is not completely functional. A valid file conforms to a specific **Schema (XSD or DTD)**. For example, a "Book" schema might require every `<book>` to have an `<isbn>`. Otherwise, a file can be well-formed but be considered *invalid* given the missing of some information like ISBN.

XML often mixes vocabularies (e.g., HTML for display, Dublin Core for metadata). **Namespaces** (`xmlns`) prevent naming collisions. Identifying namespaces helps curators understand what standards the file claims to follow (e.g., DDI, TEI, METS) .

This notebook performs a rigorous structure analysis:

1.  **Parse Check:** Verifies well-formedness using `xml2` .
2.  **Schema Validation:** If an `.xsd` file is found in the directory, it validates the XMLs against it .
3.  **Complexity Scan:** Calculates the **Max Nesting Depth**. Excessive depth (\>50 levels) can indicate "XML Bombs" or poor design.
4.  **Namespace Inventory:** Extracts all declared namespaces.

------------------------------------------------------------------------

## Setup

We use the `xml2` package, which is a wrapper around the comprehensive `libxml2` C library.

### R Packages

If you do not have the required packages, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "xml2", "rstudioapi"))
```

### Load libraries

```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(xml2)       # Parsing and Validation
library(rstudioapi) # Directory selection
```

## Select Target Directory

Select the folder containing the XML files.

**Note:** If running interactively, a dialog box will appear. Otherwise, it defaults to the target_dir parameter.

```{r}
#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select XML Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))

```

## Inventory of files and schemas

We scan for XML files and any accompanying XSD schemas.

```{r}
#| label: inventory

# Find XML files
xml_files <- list.files(
  path = target_dir,
  pattern = "\\.xml$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

# Find XSD Schemas
xsd_files <- list.files(
  path = target_dir,
  pattern = "\\.xsd$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

print(paste("Found", length(xml_files), "XML files."))
print(paste("Found", length(xsd_files), "XSD Schema files."))

# Use the first XSD found for validation (if any)
active_schema <- if (length(xsd_files) > 0) read_xml(xsd_files[1]) else NULL
if (!is.null(active_schema)) message(paste("Using Schema:", basename(xsd_files[1])))
```

## Generate Data Dictionary

This code iterates through the files and extracts their internal structure.

```{r}
#| label: extraction-logic
#| warning: false
#| message: false

message("Generating XML Report...")

# Helper: Calculate Max Depth (Recursive)
get_max_depth <- function(node) {
  children <- xml_children(node)
  if (length(children) == 0) return(1)
  return(1 + max(sapply(children, get_max_depth)))
}

report <- purrr::map_dfr(xml_files, function(file_path) {
  
  fname <- basename(file_path)
  
  tryCatch({
    # 1. Parse (Well-Formedness Check)
    doc <- read_xml(file_path)
    
    # 2. Validation (if Schema exists)
    validity <- "Not Checked (No XSD)"
    if (!is.null(active_schema)) {
      is_valid <- xml_validate(doc, active_schema)
      validity <- if (is_valid) "Valid" else "Invalid (Schema Violation)"
    }
    
    # 3. Structure Analysis
    root <- xml_root(doc)
    root_name <- xml_name(root)
    num_children <- length(xml_children(root))
    
    # Namespaces
    ns <- xml_ns(doc)
    ns_str <- if (length(ns) > 0) paste(names(ns), collapse = ", ") else "None"
    
    # Complexity (Depth)
    # Note: For huge files, depth calc can be slow. We limit it implicitly by recursion.
    max_depth <- tryCatch(get_max_depth(root), error = function(e) "Error")
    
    tibble(
      FileName = fname,
      Status = "Well-Formed",
      SchemaValidation = validity,
      RootNode = root_name,
      Namespaces = ns_str,
      MaxDepth = as.character(max_depth),
      DirectChildren = num_children
    )
    
  }, error = function(e) {
    tibble(
      FileName = fname,
      Status = "Parsing Failed (Not Well-Formed)",
      SchemaValidation = "Failed",
      RootNode = "Error",
      Namespaces = "",
      MaxDepth = NA,
      DirectChildren = NA
    )
  })
})

# Display preview
print("--- XML Structure Report ---")
head(report)
```

## Save Results

Save the dictionary to a CSV file for review.

```{r}
#| label: save-results

output_dir <- "Results/Inspect_xml"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

output_file <- file.path(output_dir, paste0("XML_Structure_", format(Sys.Date(), "%Y%m%d"), ".csv"))

write.csv(report, output_file, row.names = FALSE)
print(paste("Report saved to:", output_file))
```

## Curation Insights

Use the generated CSV to guide your preservation actions:

-   **Status ("Parsing Failed"):** If the parsing fails, it may implicate the files are corrupted or contain syntax errors (e.g., unclosed tags). Curators may need to discuss with depositors the validity of these files.

-   **Namespaces (Namespaces):** It is avisable to look for standard prefixes like dc (Dublin Core) or mods. If a specific standard is detected, verify if the file strictly adheres to it (Schema Validation) .

-   **Complexity (MaxDepth \> 50):** Deeply nested XMLs can be difficult to process with standard XSLT tools. Curators can investigate these files manually.

## Additional Tools

-   **Oxygen XML Editor:** It is a [standard IDE](https://www.oxygenxml.com/) for XML development. It provides visual schema validation and XSLT debugging.

-   **xmllint:** It is a a [command-line tool](https://xmllint.com/), also part of [libxml2](https://gitlab.gnome.org/GNOME/libxml2), for checking syntax and validating against schemas in batch.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server (HPC), in a batch job, or from the command line, here is the pure R script version.

Download the **R Script:** [**`Inspect_xml_Script.R`**](Scripts/Inspect_xml_Script.R)

### Example HPC Submission Script (`Inspect_xml_submit.sh`)

``` bash
#!/bin/bash
#SBATCH --job-name=xml_check
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:15:00
#SBATCH --mem=4G
#SBATCH --output=logs/xml_check_%j.log

module load R

# Define target directory
TARGET_DIR="/scratch/user/project_data/metadata"

# Prepare folders
mkdir -p Results/Inspect_xml
mkdir -p logs

# Run
echo "Starting XML Inspection on $TARGET_DIR"
Rscript Inspect_xml_Script.R "$TARGET_DIR"
```

## References

::: {#refs}
:::
