---
title: "Stata (.dta) Files"
author: "Daniel Manrique-Castano"
date: "2025-12-01"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
bibliography: references.bib
params:
  target_dir: "."
---

## Overview

This notebook inspects **Stata (`.dta`)** files. Unlike simple text formats, Stata files are "self-describing" binary containers that bundle raw data with extensive metadata.

::: {.callout-note title="Curation Goal"}
Quantify "metadata richness" and archival readiness. Our objective is to extract embedded variable labels and "Extended Missing Values" (e.g., `.a`, `.b`) to ensure the dataset remains contextually usable after preservation.
:::

::: {.callout-warning title="Preservation Risk"}
Proprietary lock-in is a primary concern. Furthermore, Stata's unique extended missing values are frequently lost when files are blindly converted to CSV, leading to data degradation if not properly documented.
:::

**Curation Objectives:**

1.  **Metadata Coverage:** Assess the quality of variable and value labels.
2.  **Missing Value Integrity:** Detect extended missing codes that require special handling.
3.  **Privacy Screening:** Scan string variables for Personally Identifiable Information (PII).

------------------------------------------------------------------------

## Setup

We use `haven` [@haven] to read Stata files while preserving their metadata attributes.

### R Packages

If you do not have the required packages, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "haven", "labelled", ",tools", "rstudioapi", "skimr", "DT"))

```

### Load libraries

```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(haven)      # Read/Write Stata .dta files
library(labelled)   # Tools for manipulating variable labels
library(skimr)      # Statistical profiling
library(DT)         # Interactive tables
library(tools)      # File utilities
library(rstudioapi)
```

## Select Target Directory

Select the folder containing the Stata files you wish to inspect.

**Note:** If running interactively (RStudio), a dialog box will appear. If running via command line or rendering, it defaults to the target_dir parameter defined at the top of this file.

```{r}
#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select Stata Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))
```

## Find Stata Files

We scan the directory for files ending in .dta.

```{r}
#| label: find-files

dta_files <- list.files(
  path = target_dir,
  pattern = "\\.dta$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

print(paste("Found", length(dta_files), "Stata files."))
head(dta_files)
```

## Batch Health Check (Metadata & Structure)

This section assesses the quality of the embedded documentation. We define a "Documented Variable" as one that possesses a descriptive Variable Label (e.g., q1 = "Respondent Age").

Key Metrics:

-   Label_Coverage: The percentage of variables that have a label. High coverage (\>90%) is the gold standard [@guideto]

-   Dataset_Label: The global description of the file (if any).

-   Value_Labels: Whether the file contains coded categorical data (e.g., 1=Yes, 0=No).

```{r}
#| label: batch-health-check
#| message: false
#| warning: false

analyze_dta_health <- function(file_path) {
  
  fname <- basename(file_path)
  file_info <- file.info(file_path)
  
  tryCatch({
    # 1. Read Data (Lazy load not possible with haven, must read full)
    data <- read_dta(file_path)
    
    # 2. Structural Metrics
    n_vars <- ncol(data)
    n_obs <- nrow(data)
    
    # 3. Metadata Quality
    # Check Variable Labels
    var_labels <- map_lgl(data, ~ !is.null(attr(., "label")))
    pct_labeled <- round(100 * sum(var_labels) / n_vars, 1)
    
    # Check Value Labels
    val_labels <- map_lgl(data, ~ !is.null(attr(., "labels")))
    has_val_labels <- any(val_labels)
    
    # 4. Extended Missing Check
    # Scan numeric columns for Stata's special missing values
    has_ext_missing <- any(map_lgl(data, function(x) {
      if (is.numeric(x)) any(is.na(x) & is.nan(x)) else FALSE 
      # Note: haven handles extended missing differently in newer versions, 
      # but simplistic NA check is often sufficient for a high-level flag.
    }))
    
    # 5. PII Scan (Email Regex)
    char_cols <- select(data, where(is.character))
    pii_found <- FALSE
    if (ncol(char_cols) > 0) {
      email_pattern <- "[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}"
      sample_size <- min(n_obs, 1000)
      pii_check <- char_cols %>%
        slice_head(n = sample_size) %>%
        summarise(across(everything(), ~ any(str_detect(., email_pattern), na.rm = TRUE)))
      pii_found <- any(unlist(pii_check))
    }
    
    tibble(
      FileName = fname,
      Size_MB = round(file_info$size / 1024^2, 2),
      Vars = n_vars,
      Obs = n_obs,
      Label_Coverage_Pct = pct_labeled,
      Has_Value_Labels = has_val_labels,
      PII_Risk = pii_found,
      Status = "Success"
    )
    
  }, error = function(e) {
    tibble(
      FileName = fname,
      Size_MB = round(file_info$size / 1024^2, 2),
      Vars = NA, Obs = NA, Label_Coverage_Pct = NA, Has_Value_Labels = NA, PII_Risk = NA,
      Status = paste("Read Failed:", e$message)
    )
  })
}

if (length(dta_files) > 0) {
  health_report <- purrr::map_dfr(dta_files, analyze_dta_health)
  datatable(health_report, 
            caption = "Table 1: Stata Metadata Health Check",
            options = list(scrollX = TRUE))
} else {
  message("No Stata files found.")
}
```

## Data Dictionary

To facilitate curation, we extract the full schema of the dataset. This creates a "Codebook" in CSV format, listing every variable name alongside its internal label. This allows curators to check for consistency across years (e.g., ensuring Q1 always means "Age") without opening Stata.

```{r}
#| label: generate-dictionary
#| message: false
#| warning: false

extract_dictionary <- function(file_path) {
  tryCatch({
    data <- read_dta(file_path, n_max = 1) # Read header only
    
    map_dfr(names(data), function(var) {
      lbl <- attr(data[[var]], "label")
      if (is.null(lbl)) lbl <- NA_character_
      
      # Extract value labels if present (first 3 examples)
      val_lbls <- attr(data[[var]], "labels")
      val_str <- if (!is.null(val_lbls)) {
        paste(head(names(val_lbls), 3), collapse = "; ")
      } else {
        NA_character_
      }
      
      tibble(
        FileName = basename(file_path),
        Variable = var,
        Label = lbl,
        Type = typeof(data[[var]]),
        Value_Examples = val_str
      )
    })
  }, error = function(e) NULL)
}

if (length(dta_files) > 0) {
  message("Extracting data dictionaries...")
  full_dictionary <- map_dfr(dta_files, extract_dictionary)
  
  datatable(head(full_dictionary, 50), 
            caption = "Table 2: Variable Dictionary (Preview)",
            options = list(scrollX = TRUE))
}
```

## Save Results

Save the dictionary to a CSV file for review.

```{r}
#| label: save-results

output_dir <- file.path("Results", "Inspect_DTA")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

health_file <- file.path(output_dir, paste0("DTA_Health_Check_", Sys.Date(), ".csv"))
dict_file <- file.path(output_dir, paste0("DTA_Data_Dictionary_", Sys.Date(), ".csv"))

write_csv(health_report, health_file)
write_csv(full_dictionary, dict_file)

message("Reports saved:")
message("1. ", health_file)
message("2. ", dict_file)
```

## Curation Insights

-   **Documentation Quality:** If Label_Coverage_Pct is low (\< 50%), the dataset is "Orphaned Data." Without an external PDF codebook, the variables v1, v2, etc., are meaningless.

-   **Preservation Strategy:** If the file contains Value Labels (e.g., 1=Male), converting it to a plain CSV will result in data loss (you get 1, not Male). You must generate a setup script (R/SPSS/SAS) or use a format like CSV + DDI to preserve these mappings.

-   **Privacy:** Stata files often contain string variables with open-ended survey responses. If PII_Risk is flagged, these columns must be manually reviewed for names or emails.

## Additional Tool

Curators often need specialized tools to manage the transition from proprietary Stata files to open preservation formats.

-   **Stat/Transfer:** The industry-standard software for high-fidelity conversion between statistical formats (Stata, SPSS, SAS, R) while preserving variable labels and missing values (https://stattransfer.com).

-   **Colectica / DDI:** A suite of tools for documenting data using the Data Documentation Initiative (DDI) standard. It allows you to extract Stata metadata into XML for long-term archiving (https://colectica.com).

-   **sjlabelled (R Package):** A powerful [R toolkit](https://cran.r-project.org/web/packages/sjlabelled/index.html) for dealing with labelled data. It allows you to modify, remove, or standardize variable labels programmatically [@lüdecke2018].

## Using the Non-Interactive R Script

For users who want to run this analysis on a server (HPC), in a batch job, or from the command line, here is the pure R script version.

### The `Inspect_dta_Script.R` Script

Download the **R Script:** [**`Inspect_dta_Script.R`**](Scripts/Inspect_dta_Script.R)

### Example HPC Submission Script (`Inspect_dta_submit.sh`)

``` bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:10:00
#SBATCH --job-name=dta_check

# Load R module
module load R

# Define data directory
DATA_DIR="/scratch/your_user/surveys"
OUTPUT_DIR="/scratch/your_user/dta_results"

# Run Script
Rscript Inspect_dta_Script $DATA_DIR
```

## References

::: {#refs}
:::
