---
title: "NetCDF Files"
author: "Natalie Williams"
date: "2025-10-16"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
params:
  target_dir: "data/Inspect_nc/"
---

## Overview

[NetCDF](https://www.ogc.org/standards/netcdf/) (Network Common Data Form) is the OGC standard for storing multidimensional scientific data. Unlike CSVs, NetCDF files are self-describing, meaning they contain a header describing the layout, units, and dimensions of the data inside. This makes them ideal for archiving complex climate models and satellite imagery.

NetCDF defines *how* to store data but does not define *what* the data represents. Therefore, the community developed the [Climate and Forecast (CF) Conventions](https://cfconventions.org/). These metadata rules ensure that software knows, for instance, that a variable named `temp` with units `K` represents physical temperature.

This notebook evaluates NetCDF files on three levels:

1.  **Metadata Compliance:** Checks for global attributes like `Conventions` or `institution`.
2.  **Spatial Awareness:** Detects if the file contains a defined **Coordinate Reference System (CRS)** (e.g., Latitude/Longitude grids), which is essential for GIS reuse.
3.  **Data Health:** Performs a "lightweight scan" of the first data slice to check if the file is empty (100% NaNs).

------------------------------------------------------------------------

## Setup

Before running this notebook, you need to ensure the required R packages are installed.

### R Packages

The following R packages are required. If you don't have them, run this code once in your R console:

```{r}
# install.packages(c("tidyverse", "tidync", "ncmeta", "rstudioapi"))
```

------------------------------------------------------------------------

## Load Libraries

This chunk loads all the necessary libraries for the session.

```{r load-libraries}
#| label: load-libraries
#| message: false

library(tidyverse)
library(tidync)     # Tidy interface for NetCDF
library(ncmeta)     # Low-level metadata extraction
library(rstudioapi) # For directory selection
```

------------------------------------------------------------------------

## Select target directory with NetCDF files

This section identifies the directory to be analyzed and finds all `.nc` files within it.

> **Note:** If you are running this interactively in RStudio, a dialog box will appear. If you are rendering this document (where no user interaction is possible), it will default to the `params$target_dir` defined in the YAML header.

```{r select-directory}
#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select NetCDF Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory (Interactive vs Parameter)
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))
```

### Find NetCDF files

Now we scan the selected directory for all `.nc` files.

```{r find-files}
#| label: find-files

# Find all NetCDF files
nc_files <- list.files(
  path = target_dir,
  pattern = "\\.nc$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

# Print the number of files found and show the first few paths
print(paste("Found", length(nc_files), "NetCDF files."))
head(nc_files)
```

------------------------------------------------------------------------

## Usability scan

This phase evaluates the "fitness for use." We check if the files contain valid spatial coordinates (CRS) and if the data variables actually contain numbers (Data Health).

```{r}
#| label: inventory-logic
#| warning: false
#| message: false

nc_files <- list.files(target_dir, pattern = "\\.nc$", full.names = TRUE, recursive = TRUE)

# Helper: Usability Scan
inspect_nc_inventory <- function(fp) {
  fname <- basename(fp)
  
  # 1. Safe Load
  tnc <- tryCatch(tidync(fp), error = function(e) NULL)
  
  if (is.null(tnc)) {
    return(tibble(
      FileName = fname,
      Status = "Corrupt/Unreadable",
      DimsSummary = NA,
      VarCount = NA,
      HasCRS = NA,
      DataHealth = NA
    ))
  }
  
  # 2. Extract Metadata Summary
  # Active grid dimensions
  dims <- tnc %>% hyper_dims()
  dims_str <- paste(dims$name, collapse = " x ")
  
  # Variables
  vars <- tnc %>% hyper_vars()
  var_count <- length(vars$name)
  
  # 3. Spatial Check (CRS)
  # Look for standard lat/lon names or "grid_mapping" attribute
  has_lat <- any(str_detect(dims$name, "(?i)lat|y"))
  has_lon <- any(str_detect(dims$name, "(?i)lon|x"))
  
  # Robust attribute check using ncmeta
  all_atts <- ncmeta::nc_atts(fp)
  has_grid_mapping <- any(all_atts$name == "grid_mapping")
  
  spatial_status <- if (has_grid_mapping || (has_lat && has_lon)) "Georeferenced" else "No Spatial Grid"
  
  # 4. Data Health Check (Sparsity)
  # We read a tiny slice of the first active variable to see if it contains valid data
  is_empty_label <- "Unknown"
  try({
    first_var <- vars$name[1]
    # Pull first 100 values only
    sample_data <- tnc %>% 
      activate(first_var) %>% 
      hyper_slice(select_var = first_var) %>% 
      as_tibble()
    
    val_col <- names(sample_data)[ncol(sample_data)]
    if (all(is.na(sample_data[[val_col]]))) {
      is_empty_label <- "⚠️ All NaNs (Empty)"
    } else {
      is_empty_label <- "Contains Data"
    }
  }, silent = TRUE)
  
  return(tibble(
    FileName = fname,
    Status = "Valid",
    DimsSummary = dims_str,
    VarCount = var_count,
    HasCRS = spatial_status,
    DataHealth = is_empty_label
  ))
}

# Run Inventory
message(paste("Scanning", length(nc_files), "files for usability..."))
inventory_results <- map_dfr(nc_files, inspect_nc_inventory)

print(paste("Inventory complete for", nrow(inventory_results), "files."))
head(inventory_results)
```

## Metadata Extraction

Now, we perform a deep extraction of all attributes to create detailed documentation.

```{r extract-metadata}
#| label: deep-extraction
#| warning: false
#| message: false

# Create a "safely" version of tidync to handle potentially corrupt files
safe_tidync <- purrr::safely(tidync)

# 1. Process all files
processed_files <- purrr::map(nc_files, ~safe_tidync(.x)) %>% 
  set_names(nc_files)

# 2. Separate successful results
successful_results <- purrr::map(processed_files, "result") %>% 
  purrr::compact()

errors <- purrr::map(processed_files, "error") %>% 
  purrr::compact()

if (length(errors) > 0) {
  message("The following files failed deep extraction:")
  walk(names(errors), message)
}

# 3. Extract Raw Components
nc_dimensions <- purrr::map(successful_results, ~.x$dimension) %>% 
  bind_rows(.id = "FileName") %>% mutate(FileName = basename(FileName))

nc_variables <- purrr::map(successful_results, ~.x$variable) %>% 
  bind_rows(.id = "FileName") %>% mutate(FileName = basename(FileName))

nc_attributes <- purrr::map(successful_results, ~.x$attribute) %>% 
  bind_rows(.id = "FileName") %>% mutate(FileName = basename(FileName))

print("Deep extraction complete.")
```

------------------------------------------------------------------------

## Reshape metadata for comparison

We reshape the metadata into two summary tables: one for Global Attributes (file-level) and one for Variable Attributes (variable-level).

```{r reshape-global-attributes}
#| label: reshape-attributes

# A. Global Attributes Summary
nc_attributes_global <- nc_attributes %>%
  filter(variable == "NC_GLOBAL") %>%
  pivot_wider(
    id_cols = FileName,
    names_from = name,
    values_from = value,
    values_fn = ~paste(., collapse = "; ")
  )

print("--- Global Attributes Summary ---")
glimpse(nc_attributes_global)

# B. Variable Attributes Summary
nc_variables_with_attributes <- nc_variables %>%
  left_join(
    filter(nc_attributes, variable != "NC_GLOBAL"),
    by = c("name" = "variable", "FileName")
  ) %>%
  pivot_wider(
    names_from = name.y, 
    values_from = value,
    values_fn = ~paste(., collapse = "; ")
  )

print("--- Variables and Attributes Summary ---")
head(nc_variables_with_attributes)
```

------------------------------------------------------------------------

## Save Summary Reports

Finally, we save the most useful summary tables to `.csv` files for documentation and further analysis.

```{r write-files}
#| label: save-results

output_dir <- "Results/Inspect_nc"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

timestamp <- format(Sys.Date(), "%Y%m%d")

# 1. Save Inventory (High-Level Scan)
write.csv(inventory_results, file.path(output_dir, paste0("NetCDF_Inventory_", timestamp, ".csv")), row.names = FALSE)

# 2. Save Deep Metadata (Detailed Tables)
write.csv(nc_dimensions, file.path(output_dir, paste0("NetCDF_Dimensions_", timestamp, ".csv")), row.names = FALSE)
write.csv(nc_attributes_global, file.path(output_dir, paste0("NetCDF_Global_Attributes_", timestamp, ".csv")), row.names = FALSE)
write.csv(nc_variables_with_attributes, file.path(output_dir, paste0("NetCDF_Variables_Attributes_", timestamp, ".csv")), row.names = FALSE)

print(paste("All reports saved to:", output_dir))
```

------------------------------------------------------------------------

## Curation Insights

Use the generated reports to guide your preservation actions:

-   **Spatial Awareness (HasCRS):** Files marked "No Spatial Grid" lack standard latitude/longitude coordinates or a grid_mapping attribute. These files will have difficulties to load in GIS software. Check if they are non-spatial (e.g., time-series at a single station) or if the metadata is missing.

-   **Data Health (DataHealth):** Files marked "⚠️ All NaNs" are likely empty shells—the model ran but produced no output. You can verify these files manually and exclude them from the archive.

-   **Metadata Compliance:** In the NetCDF_Global_Attributes CSV, the curator can check for the attribute Conventions = "CF-1.x" and verify fullfilment of gold standards.

## Additional Tools & Resources

### Verify consistent global attributes

Verify if all files in a dataset have the same `title`, `institution`, `source`, and CF `Conventions`.

```{r curation-check-global}
#| label: curation-check-global
#| code-fold: false

if (nrow(nc_attributes_global) > 0) {
  # Select a few key attributes and count the unique combinations
  global_consistency_check <- nc_attributes_global %>%
    select(FileName, contains("title"), contains("institution"), contains("source"), contains("Conventions")) %>%
    # The line below groups by all columns except FileName
    group_by(across(-FileName)) %>%
    summarise(file_count = n(), .groups = "drop")
  
  print("Consistency Check of Key Global Attributes:")
  print(global_consistency_check)
}
```

### Check for essential variable attributes

For data to be reusable, variables should always have attributes like `long_name` and `units`. This section allows checking for missing attributes across all variables.

```{r curation-check-var-attr}
#| label: curation-check-var-attr
#| code-fold: false

if (nrow(nc_variables_with_attributes) > 0) {
  missing_attribute_check <- nc_variables_with_attributes %>%
    # Summarise the number of variables missing these key attributes
    summarise(
      missing_long_name = sum(is.na(long_name)),
      missing_units = sum(is.na(units))
    )
  
  print("Check for Missing Essential Variable Attributes:")
  print(missing_attribute_check)
}
```

-   **CDO (Climate Data Operators):** Is a [command-line suite](https://code.mpimet.mpg.de/projects/cdo) for manipulating and analyzing NetCDF data. It is the industry standard for regridding and statistical aggregation.

-   **Panoply:** A cross-platform [application](https://www.giss.nasa.gov/tools/panoply/) from NASA that plots geo-referenced arrays from NetCDF files. Excellent for "Visual QC".

-   **NCO (NetCDF Operators):** A [toolkit](https://nco.sourceforge.net/) to perform arithmetic and attribute editing on NetCDF files [@zender2008].

## Using the Non-Interactive R Script

For users who want to run this analysis on a server, in a batch job, or from the command line, here is a pure R script that performs the same process.

### The `Inspect_nc_Script.R` Script

Download the **R Script:** [**`Inspect_nc_Script.R`**](Scripts/Inspect_nc_Script.R)

### Example HPC Submission Script (`Inspect_nc_submit.sh`)

``` bash
#!/bin/bash
#SBATCH --job-name=nc_inspect
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=2
#SBATCH --mem=8G
#SBATCH --time=00:30:00
#SBATCH --output=logs/nc_inspect_%j.log

# 1. Load R Module
module load R

# 2. Define Directory
TARGET_DIR="/scratch/user/project_data/climate_models"

# 3. Prepare Environment
mkdir -p Results/Inspect_nc
mkdir -p logs

# 4. Run Analysis
echo "Starting NetCDF Inspection on $TARGET_DIR"
Rscript Inspect_nc_Script.R "$TARGET_DIR"
```


## References

::: {#refs}
:::
