---
title: "Container/Archive (.zip, .tar, .7z) Formats"
author: "Data Curation Team"
date: "2025-01-05"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: cosmo
params:
  target_dir: "data/Inspect_Containers/"
bibliography: references.bib
---

## Overview

Curators often receive data packaged in containers like **ZIP**, **TAR**, or **7-Zip**. While convenient for transfer, these formats are "black boxes" that hide potential preservation risks:

- **Corruption:** Archives are prone to bit-rot. A single corrupted bit in a solid archive (like `.7z`) can render the entire dataset unreadable.

- **"Zip Bombs" (Decompression Bombs):** Malicious or accidentally created files that are small when compressed (e.g., 42 KB) but expand to petabytes (4.5 PB) when opened, crashing the curation workstation.

- **Nested Archives:** Archives inside archives (e.g., `data.zip` containing `backup.tar.gz`) create "Russian Doll" problems for preservation systems.

This notebook uses the `archive` package (based on `libarchive`) to "peek" inside the container without fully extracting it. We assess:

1.  **Integrity:** Can the file be read? (Basic corruption check).
2.  **Expansion Ratio:** We calculate the **Compression Ratio**. If the extracted size is >100x the file size, it is flagged as a potential Zip Bomb.
3.  **Manifest:** We generate a summary of contents (e.g., "Contains 500 .csv files and 2 .exe files").

------------------------------------------------------------------------

## Setup

We use the `archive` package, which is a robust binding to the industry-standard `libarchive` C library.

### R Packages

If you do not have the required packages, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "archive", "fs", "rstudioapi"))
```

### Load libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(archive)    # The engine for reading archives
library(fs)         # File system tools
library(rstudioapi)
```

## Select Target Directory

Select the folder containing the TiFF files.

**Note:** If running interactively, a dialog box will appear. Otherwise, it defaults to the target_dir parameter.

```{r}
#| label: select-target-dir

if (interactive() && .Platform$OS.type == "windows") {
  selected_dir <- rstudioapi::selectDirectory(caption = "Select TIFF Directory")
} else {
  selected_dir <- NULL
}

if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))

```

## Inventory of files

We scan for `.tif` and `.tiff` files.

```{r}
#| label: inventory

# Find archives (zip, tar, gz, 7z, rar)
archive_files <- list.files(
  path = target_dir,
  pattern = "\\.(zip|tar|tar\\.gz|tgz|7z|rar)$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

if (length(archive_files) == 0) {
  stop("No archive files found in this folder.")
}

print(paste("Found", length(archive_files), "archive files."))
```

## Non-Invasive Inspection

This function reads the header (Table of Contents) of the archive to calculate metrics without filling up your hard drive with extracted files.

```{r}
#| label: inspection-logic
#| warning: false
#| message: false

message("Generating Archive Manifests...")

inspect_archive <- function(fp) {
  fname <- basename(fp)
  
  tryCatch({
    # Get Physical File Size (Compressed)
    size_compressed_bytes <- file.size(fp)
    
    # Read the Manifest (Does NOT extract files)
    # Returns a tibble: path, size, date, mode
    contents <- archive::archive(fp)
    
    # Calculate Metrics
    file_count <- nrow(contents)
    size_extracted_bytes <- sum(contents$size)
    
    # Zip Bomb Detection (Compression Ratio)
    # Avoid division by zero for empty archives
    ratio <- if(size_compressed_bytes > 0) size_extracted_bytes / size_compressed_bytes else 0
    
    # Content Profiling
    # What kind of files are inside? Get extensions.
    extensions <- fs::path_ext(contents$path)
    top_exts <- names(sort(table(extensions), decreasing = TRUE))[1:3]
    content_summary <- paste(top_exts, collapse = ", ")
    
    # Check for Nested Archives
    has_nested <- any(extensions %in% c("zip", "tar", "gz", "7z", "rar"))
    
    tibble(
      FileName = fname,
      FileCount = file_count,
      Compressed_MB = round(size_compressed_bytes / 1024^2, 2),
      Extracted_MB = round(size_extracted_bytes / 1024^2, 2),
      CompressionRatio = round(ratio, 1),
      ContentTypes = content_summary,
      HasNestedArchives = has_nested,
      Status = "Success"
    )
    
  }, error = function(e) {
    tibble(
      FileName = fname, FileCount = NA, Compressed_MB = NA, 
      Extracted_MB = NA, CompressionRatio = NA, ContentTypes = NA,
      HasNestedArchives = NA,
      Status = paste("Corrupt/Unreadable:", e$message)
    )
  })
}

# Execute Analysis
report <- map_dfr(archive_files, inspect_archive)

# Display
print("--- Archive Safety Report ---")
print(head(report))
```
## Expansion Risk

We visualize the difference between the Compressed Size (Storage) and Extracted Size (Risk). Huge discrepancies indicate potential decompression issues.

```{r}
#| label: viz-risk
#| fig-cap: "Compressed vs. Extracted Size (Log Scale)"

if (nrow(report) > 0 && any(report$Status == "Success")) {
  
  # Reshape for plotting
  plot_data <- report %>%
    filter(Status == "Success") %>%
    select(FileName, Compressed_MB, Extracted_MB) %>%
    pivot_longer(cols = c("Compressed_MB", "Extracted_MB"), names_to = "Type", values_to = "Size")

  ggplot(plot_data, aes(x = FileName, y = Size, fill = Type)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_y_log10() + # Use log scale because differences can be massive
    labs(
      title = "Archive Expansion Potential",
      y = "Size (MB) - Log Scale",
      x = NULL
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Save Results

Save the report to a CSV file for review.

```{r}
#| label: save-results

output_dir <- "Results/Inspect_Containers"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

output_file <- file.path(output_dir, paste0("Container_Manifest_", format(Sys.Date(), "%Y%m%d"), ".csv"))

write.csv(report, output_file, row.names = FALSE)
print(paste("Manifest saved to:", output_file))
```

## Curation Insights

Use this report to ensure safety before ingesting:

- **Compression Ratio > 100:** This is a potential Zip Bomb. A 500 MB file turning into 500 GB can fill a hard drive instantly. Inspect these files cautiously. 

- **Status "Corrupt/Unreadable":** This implies the container header may be  broken and the data inside is likely lost. Request a re-transfer from the depositor.

- **HasNestedArchives = TRUE:** Preservation tools often fail to index "Zip inside a Zip." In this case, the curator must unpack the outer layer and archive the inner files directly to ensure discoverability.

## Additional Tools

- **7-Zip:** The [open-source standard](https://www.7-zip.org/) for handling high-compression archives. It supports almost every format.

- **Droid (Digital Record Object Identification):** A [tool](https://github.com/digital-preservation/droid) from The National Archives (UK) to profile file formats inside archives without full extraction.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server (HPC), in a batch job, or from the command line, here is the pure R script version.

Download the **R Script:** [**`Inspect_Containers_Script.R`**](Scripts/Inspect_Containers_Script.R)

### Example HPC Submission Script (`Inspect_Containers_submit.sh`)

``` bash
#!/bin/bash
#SBATCH --job-name=archive_check
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:30:00
#SBATCH --mem=8G
#SBATCH --output=logs/archive_check_%j.log

module load R
# Note: The 'archive' package relies on libarchive. 
# If it fails, you might need: module load libarchive

# Define target directory
DATA_DIR="/scratch/user/project_data/deposits"

# Prepare Environment
mkdir -p Results/Inspect_Archives
mkdir -p logs

# Run
echo "Starting Archive Inspection on $DATA_DIR"
Rscript Inspect_Archive_Script.R "$DATA_DIR"
```

## References

::: {#refs}
:::
