---
title: "HDF5 Files"
author: "Daniel Manrique-Castano"
date: "2025-12-05"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
bibliography: references.bib
params:
  target_dir: "data/Inspect_hdf5/"
---

## Overview

This notebook audits **Hierarchical Data Format version 5 (HDF5)** files (`.h5`, `.hdf5`). Unlike flat text files, [HDF5](https://support.hdfgroup.org/documentation/hdf5/latest/_h5_d_m__u_g.html) is a binary container that functions like a hard drive, capable of storing complex hierarchies of data within a single object.

::: {.callout-note title="Curation Goal"}
Map the internal "filesystem-like" structure. Our objective is to inventory groups, datasets, and attributes, ensuring that the internal hierarchy is well-documented and remains navigable for future research.
:::

::: {.callout-warning title="Preservation Risk"}
HDF5 is a "Black Box." Data can be locked behind proprietary compression filters or rely on external links that break when files are moved. Without specific documentation, the internal layout is effectively invisible to standard archival tools.
:::

**Curation Objectives:**

1.  **Inventory:** Map the internal directory structure (Groups vs. Datasets).
2.  **Dependencies:** Detect required compression libraries (e.g., GZIP vs. Proprietary).
3.  **Integrity:** Identify external links that may break if files are moved.
4.  **Metadata Extraction:** Verify that the data is self-describing through internal attributes.

------------------------------------------------------------------------

## Setup

We use the `hdf5r` package, which provides an object-oriented interface to the HDF5 library.

### R Packages

If you do not have the required packages, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "hdf5r", "rstudioapi"))
```

### Load libraries

```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(hdf5r)      # Interface to HDF5 library
library(DT)         # Interactive tables
library(rstudioapi) # Directory selection

# Validation: Check if hdf5r loaded correctly
if (!require("hdf5r", quietly = TRUE)) {
  stop("The 'hdf5r' package is missing. Please install it to proceed.")
}
```

## Select Target Directory

We select the folder containing the HDF5 files.

**Note:** If running interactively, a dialog box will appear. Otherwise, it defaults to the target_dir parameter.

```{r}
#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select HDF5 Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))

```

## Find HDF5 Files

We scan the directory for files ending in .h5 or .hdf5.

```{r}
#| label: find-files

hdf5_files <- list.files(
  path = target_dir,
  pattern = "\\.(h5|hdf5)$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

print(paste("Found", length(hdf5_files), "HDF5 files."))
head(hdf5_files)
```

## File Inspection

We iterate through each file to map its internal contents. This routine extracts not just dimensions, but also storage layouts and compression filters.

**Key Technical Checks:**

-   **Filters:** We check the "creation property list" of every dataset to see if compression (e.g., GZIP, SZIP) is active.

-   **Link Type:** We flag H5L_TYPE_EXTERNAL to warn about potential missing dependencies.

```{r}
#| label: generate-dictionary
#| warning: false
#| message: false

message("Generating Data Dictionary...")

analyze_hdf5_structure <- function(file_path) {
  fname <- basename(file_path)
  
  tryCatch({
    # Open File (Read-Only)
    h5f <- H5File$new(file_path, mode = "r")
    on.exit(h5f$close_all()) # Ensure closure even if errors occur
    
    # 1. List all objects recursively
    # The 'ls' function returns a dataframe with columns: name, link.type, obj.type, etc.
    contents <- h5f$ls(recursive = TRUE)
    
    # 2. Iterate through objects to extract deep metadata
    purrr::map_dfr(seq_len(nrow(contents)), function(i) {
      
      obj_path <- contents$name[i]
      obj_type <- contents$obj_type[i] # H5I_GROUP or H5I_DATASET
      link_type <- contents$link.type[i] # H5L_TYPE_HARD, H5L_TYPE_EXTERNAL
      
      # Defaults
      dims <- NA_character_
      dtype <- NA_character_
      compression <- "None"
      layout <- NA_character_
      attrs_str <- ""
      
      # CASE A: External Link (Risk!)
      if (link_type == "H5L_TYPE_EXTERNAL") {
        return(tibble(
          FileName = fname, Path = obj_path, Type = "EXTERNAL_LINK",
          Dimensions = NA, DataType = NA, Compression = NA, 
          Attributes = "Warning: Points to external file", Status = "Risk: External Dependency"
        ))
      }
      
      # CASE B: Dataset (The actual data)
      if (obj_type == "H5I_DATASET") {
        tryCatch({
          dset <- h5f[[obj_path]]
          
          # Dimensions & Type
          dims <- paste(dset$dims, collapse = " x ")
          dtype <- dset$get_type()$to_text()
          
          # Advanced: Compression & Layout (Creation Properties)
          dcpl <- dset$create_plist
          layout <- dcpl$get_layout() # e.g., H5D_CHUNKED
          
          # Filters (Compression)
          n_filters <- dcpl$get_nfilters()
          if (n_filters > 0) {
            filters <- map_chr(0:(n_filters - 1), ~ dcpl$get_filter(.x)$name)
            compression <- paste(filters, collapse = ", ")
          }
          
          # Attributes
          attr_list <- names(h5attributes(dset))
          if (length(attr_list) > 0) attrs_str <- paste(head(attr_list, 5), collapse = "; ")
          
        }, error = function(e) {
          dims <<- "Error reading dataset"
        })
      }
      
      # CASE C: Group (Folder)
      if (obj_type == "H5I_GROUP") {
         tryCatch({
          grp <- h5f[[obj_path]]
          attr_list <- names(h5attributes(grp))
          if (length(attr_list) > 0) attrs_str <- paste(head(attr_list, 5), collapse = "; ")
         }, error = function(e) {})
      }
      
      tibble(
        FileName = fname,
        Path = obj_path,
        Type = obj_type,
        Dimensions = dims,
        DataType = dtype,
        Compression = compression,
        Attributes = substr(attrs_str, 1, 100),
        Status = "Success"
      )
    })
    
  }, error = function(e) {
    tibble(
      FileName = fname, Path = "ROOT", Type = "ERROR", 
      Dimensions = NA, DataType = NA, Compression = NA, Attributes = NA,
      Status = paste("File Read Failed:", e$message)
    )
  })
}

if (length(hdf5_files) > 0) {
  report <- purrr::map_dfr(hdf5_files, analyze_hdf5_structure)
  
  datatable(report, 
            caption = "Table 1: HDF5 Internal Structure & Metadata",
            options = list(scrollX = TRUE, pageLength = 15))
} else {
  message("No HDF5 files found.")
}
```

## Visualization

Understanding how the data is stored (Compressed vs. Uncompressed) helps assess software dependencies.

```{r}
#| label: viz-compression
#| fig-cap: "Figure 1: Compression Methods Used. Proprietary filters pose preservation risks."

if (exists("report") && nrow(report) > 0) {
  
  # Filter only for Datasets (exclude Groups)
  dataset_info <- report %>% filter(Type == "H5I_DATASET")
  
  if (nrow(dataset_info) > 0) {
    ggplot(dataset_info, aes(x = Compression)) +
      geom_bar(fill = "#4C78A8") +
      labs(
        title = "Compression Filters in Use",
        subtitle = "'None' or 'GZIP' are preferred. Custom filters (e.g., LZF) require plugins.",
        x = "Compression Filter",
        y = "Number of Datasets"
      ) +
      theme_minimal()
  } else {
    message("No datasets found to visualize.")
  }
}
```

## Save Results

Save the dictionary to a CSV file for review.

```{r}
#| label: save-results

# Define output directory
output_dir <- file.path("Results", "Inspect_hdf5")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Define filename
output_file <- file.path(output_dir, paste0("HDF5_Dictionary_", Sys.Date(), ".csv"))

# Write CSV
write.csv(report, output_file, row.names = FALSE)

print(paste("Data Dictionary saved to:", output_file))
```

## Curation Insights

Use the generated CSV to perform these checks:

-   **Proprietary Compression:** If the Compression column lists "SZIP" (Science Data Process) or "LZF," standard tools may fail to open the file in the future. GZIP (Deflate) is the safest standard.

-   **External Links:** If Type is EXTERNAL_LINK, the file is not self-contained. It relies on a "target file" that must also be present in the directory. If missing, the link is broken.

-   **Attribute Metadata:** HDF5 is "Self-Describing" only if the researcher adds attributes. If the Attributes column is empty for major datasets, the data units (e.g., "Meters" vs "Feet") are unknown.

## Additional Tools & Resources

While R is excellent for scripted inspection, visual tools are often better for exploratory curation.

-   **HDFView:** The official Java-based visual browser maintained by The HDF Group. It allows you to click through the hierarchy, plot simple graphs, and edit attributes manually (see https://hdfgroup.org/downloads/hdfview/).

-   **h5dump (Command Line):** A [utility](https://support.hdfgroup.org/documentation/hdf5/latest/_h5_t_o_o_l__d_p__u_g.html) that converts binary HDF5 content into human-readable text (ASCII or XML). It is essential for generating archival "dumps" of metadata.

-   **Panoply:** A cross-platform [data viewer](https://www.giss.nasa.gov/tools/panoply/) specifically for HDF5/NetCDF files that adhere to climate standards (CF Conventions).

## Using the Non-Interactive R Script

For users who want to run this analysis on a server (HPC), in a batch job, or from the command line, here is the pure R script version.

### The `Inspect_hdf5_Script.R` Script

Download the **R Script:** [**`Inspect_hdf5_Script.R`**](Scripts/Inspect_hdf5_Script.R)

### Example HPC Submission Script (`Inspect_hdf5_submit.sh`)

``` bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:20:00
#SBATCH --job-name=hdf5_check

# Load R module
module load R

# IMPORTANT: The 'hdf5r' package relies on the HDF5 system library.
# On many clusters, this must be loaded explicitly.
# Check your cluster docs (e.g. 'module avail hdf5')
module load hdf5 

# Define directories
DATA_DIR="/scratch/your_user/scientific_data"
OUTPUT_DIR="/scratch/your_user/hdf5_results"

# Run Script
Rscript Inspect_hdf5_Script.R $DATA_DIR $OUTPUT_DIR
```

## References

::: {#refs}
:::