---
title: "OCR and Document Intelligence using `daiR` and Google AI"
author: "Daniel Manrique-Castano"
date: "2025-12-16"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: cosmo
params:
  target_dir: "data/Inspect_OCR/" 
bibliography: references.bib
---

## Overview

Optical Character Recognition (OCR) is the process of converting images of text (typed, handwritten, or printed) into machine-encoded text.

::: {.callout-note title="Curation Goal"}
Bridge the gap between "dark data" (unsearchable scanned PDFs) and FAIR data. Our objective is to extract text and structure from image-based documents, ensuring they become indexable and accessible for future research.
:::

::: {.callout-warning title="Preservation Risk"}
Image-only documents are effectively invisible to automated indexing and assistive technologies. While AI-based OCR is powerful, it introduces risks of "hallucination," privacy concerns when using cloud APIs, and dependency on proprietary service models.
:::

**Key Curation Objectives:**

1.  **Digitization:** Extract text from static images and scans.
2.  **Quality Assessment:** Use "Confidence Scores" to flag poor scans for manual review.
3.  **Privacy Screening:** Automatically scan extracted text for sensitive information (PII).

## Setup

**Prerequisite:** To use `daiR`[@daiR], you must have a Google Cloud Project with the **Document AI API** enabled and a Service Account JSON key.

### R Packages

If you do not have the required packages, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "daiR", "magick", "jsonlite", "rstudioapi"))
```

::: {.callout-note title="How to Setup Google Document AI"}
To run this notebook, you need a Google Cloud Platform (GCP) account and a Service Account key. Follow these steps:

**Step 1:** Create a Google Cloud Project

1.  **Create a Project:** Go to the [Google Cloud Console](https://console.cloud.google.com/) and create a new project (e.g., "OCR-Curation").
2.  **Enable the API:** In the search bar, type **"Document AI API"** and click **Enable**.
3.  **Create a Service Account:**
    -   Navigate to **IAM & Admin \> Service Accounts**.
    -   Click **+ Create Service Account**, give it a name (e.g., "ocr-bot"), and click **Create**.
    -   **Role:** Assign the role **"Document AI API User"** or **"Owner"** (for testing).
4.  **Download JSON Key:**
    -   Click on your new service account email in the list.
    -   Go to the **Keys** tab -\> **Add Key** -\> **Create new key**.
    -   Select **JSON**. This will download a file (e.g., `project-id-12345.json`) to your computer.
5.  **Authenticate in R:** Rename this file to `service-account.json` and place it in your project folder, or set the path via `Sys.setenv(GCS_AUTH_FILE = "path/to/key.json")`.

**Step 2:** Get the Processor ID

1.  Go to the Google Cloud Console \> Document AI.

2.  Click "Explore Processors" and choose "Document Parser" (General text) or "Form Parser" (if you have forms). Click Create Processor.

3.  Give it a name (e.g., my-ocr-tool) and select a region (usually US or EU).

Once created, copy the Processor ID (it looks like a1b2c3d4e5f6).
:::

### Load libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(daiR)
library(magick)
library(jsonlite)
library(pdftools) # NEW: For reading digital PDFs for free

# --- CONFIGURATION ---
# 1. Google Cloud Project ID (from your error log)
my_proj_id <- "YOUR PROJECT KEY"  
# 2. Processor ID (from your error log)
my_proc_id <- "YOUR PROCESSOR ID"           
# 3. Location
my_loc <- "us"
# 4. JSON Key File
key_path <- "service-account.json" 

# --- AUTHENTICATION ---
# We use a flag 'run_ocr' to control execution. 
# If the key is missing (e.g. on GitHub), we skip the API calls.
run_ocr <- FALSE

if (file.exists(key_path)) {
  message("Key file found. Authenticating...")
  
  # 1. Set the environment variable explicitly
  Sys.setenv(GCS_AUTH_FILE = key_path)
  
  tryCatch({
    # 2. Authenticate
    dai_auth()
    
    # 3. Verify: Check if we can get an access token
    token <- dai_token()
    if (!is.null(token)) {
      message("✅ Authentication Successful! Token acquired.")
      run_ocr <- TRUE
    }
  }, error = function(e) {
    warning("❌ Authentication Failed: Could not get API token. Check your JSON key permissions.")
    run_ocr <- FALSE
  })
  
} else {
  message("⚠️ BIO: Service account key not found.")
  message("Running in DEMO MODE. No calls to Google Cloud will be made.")
  run_ocr <- FALSE
}
```

## Select Target Directory

We select the folder we want to apply OCR.

```{r}
#| label: select-directory

# Interactive selection logic (same as previous notebooks)
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select Document Directory")
} else {
  selected_dir <- NULL
}

if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))
```

## Inventory and Pre-processing

We identify PDF and Image files. Note: Google Document AI has file size limits (usually 20MB for synchronous processing). We filter out overly large files to prevent API errors.

```{r}
#| label: inventory

files <- list.files(
  path = target_dir,
  pattern = "\\.(pdf|jpg|png|jpeg|tif|tiff)$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

inventory <- tibble(file_path = files) %>%
  mutate(
    filename = basename(file_path),
    filesize_mb = file.size(file_path) / 1024^2,
    # We must explicitly create this column for the filter to work later
    is_large = filesize_mb > 20
  )

print(paste("Found", nrow(inventory), "documents."))
head(inventory)
```

## OCR Extraction (Processing)

We iterate through the files and send them to the API.

::: callout-warning
This step consumes Google Cloud credits.
:::

```{r}
#| label: ocr-processing
#| message: false

# Helper: Layout-Aware Digital Extraction (Fixes Truncation)
extract_digital_robust <- function(fp) {
  tryCatch({
    # pdf_data returns the X/Y coordinates of every word
    data_pages <- pdftools::pdf_data(fp)
    
    # We reconstruct the text line-by-line based on these coordinates
    full_text <- map_chr(data_pages, function(page_df) {
      if (nrow(page_df) == 0) return("")
      page_df %>%
        arrange(y, x) %>% # Sort by vertical (Y) then horizontal (X) position
        group_by(y) %>%
        summarise(line_text = paste(text, collapse = " "), .groups = "drop") %>%
        pull(line_text) %>%
        paste(collapse = "\n")
    }) %>% paste(collapse = "\n\n")
    
    return(full_text)
  }, error = function(e) return(""))
}

# Main Processing Function
process_doc_robust <- function(fp) {
  
  # --- STEP 1: Try Local Digital Extraction ---
  if (str_detect(fp, "(?i)\\.pdf$")) {
    digital_text <- extract_digital_robust(fp)
    
    # Validation: If we found substantial text, we trust it.
    if (nchar(digital_text) > 500) {
      return(tibble(
        file_path = fp,
        extracted_text = digital_text,
        ocr_confidence = 1.0, 
        source = "LOCAL_PDF_LAYOUT",
        status = "SUCCESS"
      ))
    }
  }
  
  # --- STEP 2: Google Cloud OCR (Fallback) ---
  tryCatch({
    # 1. API Call
    raw_response <- dai_sync(fp, proc_id = my_proc_id, proj_id = my_proj_id, loc = my_loc)
    
    # 2. Parse Content
    data <- httr::content(raw_response, as = "parsed")
    
    # 3. Get Text
    txt <- data$document$text
    if (is.null(txt)) txt <- ""
    
    # 4. Get Confidence
    confs <- tryCatch({
      data$document$pages %>% map(~ .x$blocks) %>% flatten() %>% map(~ .x$layout$confidence) %>% unlist()
    }, error = function(e) NULL)
    
    avg_conf <- if (is.numeric(confs) && length(confs) > 0) mean(confs, na.rm=TRUE) else 0
    
    tibble(
      file_path = fp,
      extracted_text = txt,
      ocr_confidence = round(avg_conf, 4),
      source = "GOOGLE_DOC_AI",
      status = "SUCCESS"
    )
  }, error = function(e) {
    tibble(
      file_path = fp,
      extracted_text = NA_character_,
      ocr_confidence = NA_real_,
      source = "FAILED",
      status = paste("ERROR:", e$message)
    )
  })
}

# Run Analysis
if (run_ocr) {
  ocr_results_raw <- inventory %>%
    filter(!is_large) %>% 
    pull(file_path) %>%
    map_dfr(process_doc_robust)
  
  # Join back to inventory
  ocr_results <- inventory %>%
    inner_join(ocr_results_raw, by = "file_path")
  
  print("Processing complete.")
  ocr_results %>% select(filename, source, ocr_confidence, status)
} else {
  message("Skipping OCR processing (Demo Mode).")
}
```

## Curation and PII identification

Now that we have extracted the text, we must analyze it for preservation risks. We will perform Content Analysis to:

-   **Detect PII:** Automatically scan for sensitive patterns like Email Addresses or Social Security Numbers.

-   **Assess Quality:** Flag documents with low confidence scores (\< 85%), which usually indicate blurry scans or faint handwriting that require manual review.

-   **Identify Empty Files:** Flag files where the extraction resulted in empty strings, indicating potential errors or blank pages.

```{r}
#| label: curation-logic

if (run_ocr && exists("ocr_results")) {
  curated_data <- ocr_results %>%
    mutate(
      # 1. PII Regex Patterns (Emails and SSN-like formats)
      has_email = str_detect(extracted_text, "[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}"),
      has_ssn = str_detect(extracted_text, "\\b\\d{3}-\\d{2}-\\d{4}\\b"),
      
      # 2. Apply Curation Flags
      flag_low_conf = ifelse(ocr_confidence < 0.85, "LOW_CONFIDENCE", NA),
      flag_pii = ifelse(has_email | has_ssn, "POSSIBLE_PII", NA),
      # Flag if the extracted text is suspiciously short (under 10 chars)
      flag_empty = ifelse(nchar(extracted_text) < 10, "NO_TEXT_EXTRACTED", NA)
    ) %>%
    # Combine flags into a single readable column
    unite("curation_flags", starts_with("flag_"), sep = "; ", na.rm = TRUE, remove = FALSE)
  
  # Preview files that need attention
  curated_data %>% 
    select(filename, ocr_confidence, curation_flags)
} else {
  message("Skipping Curation Logic (Demo Mode).")
}
```

## Export the results

The following chunck performs\_

-   **Saves Sidecar Files:** It creates a separate .txt file for every document containing its full extracted text. This is the archival best practice you asked for.

-   **Saves the Metadata Report:** It saves the CSV with all the flags and confidence scores, but excludes the massive text blocks to keep the CSV clean and usable.

```{r}
#| label: save-results

if (run_ocr && exists("curated_data")) {
  # 1. Setup Output Directory
  output_dir <- "Results/Inspect_OCR"
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
  
  # 2. Save Individual Text Files (Sidecars)
  # This loops through every file and writes its content to a matching .txt file
  curated_data %>%
    filter(!is.na(extracted_text) & extracted_text != "") %>%
    pwalk(function(filename, extracted_text, ...) {
      # Create a txt filename (e.g., "document.pdf.txt")
      txt_name <- paste0(output_dir, "/", filename, ".txt")
      writeLines(extracted_text, txt_name)
    })
  
  # 3. Save Administrative Metadata (CSV)
  # We exclude the full text from the CSV to keep it clean, now that we have .txt files
  timestamp <- format(Sys.Date(), "%Y%m%d")
  csv_file <- paste0(output_dir, "/Curation_Report_OCR_", timestamp, ".csv")
  
  final_export <- curated_data %>%
    select(filename, ocr_confidence, source, curation_flags, status) # Removed 'extracted_text'
  
  write.csv(final_export, csv_file, row.names = FALSE)
  
  print(paste("Process Complete."))
  print(paste("1. Metadata saved to:", csv_file))
  print(paste("2. Text files saved to:", output_dir))
} else {
  message("Skipping Save Results (Demo Mode).")
}
```

## Curation Insights

Use the curation_flags and source columns in the generated CSV to guide your preservation actions:

-   **NO_TEXT_EXTRACTED:** This implies the file was successfully processed, but no readable text characters were found. The curator can visually verify the file. If it is a blank page, it can be excluded from the archival package to save storage. If it is an image without text (e.g., a photo of a landscape): This is expected; no action needed. If it contains text that was missed, it is likely that the image resolution is too low for OCR.

-   **LOW_CONFIDENCE (\< 0.85):** The AI was unsure about the character shapes. This is common in documents with artifacts, faint handwriting, or complex multi-column layouts. These files are "High Risk" for searchability and should not be rely on for indexing without human review.

-   **POSSIBLE_PII (Privacy Risk):** The script detected patterns matching Email Addresses or Social Security Numbers. This files must be quarantined immediately. Open the original PDF to verify if the data is sensitive and create a redacted derivative before making the dataset public.

-   **Source Check (LOCAL_PDF_LAYOUT vs GOOGLE_DOC_AI):** Files marked LOCAL_PDF_LAYOUT were read directly from the digital layer. These are usually 100% accurate. Focus your manual QC efforts on the files marked GOOGLE_DOC_AI, as these relied on visual pattern recognition and are more prone to "typos."

Here are the Curation Insights and Additional Tools & Resources sections, tailored specifically for the results generated by this hybrid OCR notebook. You can copy and paste this directly at the end of your .qmd file.

Curation Insights Use the curation_flags and source columns in the generated CSV to guide your preservation actions:

NO_TEXT_EXTRACTED:

Context: The file was successfully processed, but no readable text characters were found.

Action: Visually verify the file.

If it is a blank page: Exclude it from the archival package to save storage.

If it is an image without text (e.g., a photo of a landscape): This is expected; no action needed.

If it contains text that was missed: The image resolution may be too low for OCR. Flag for manual transcription.

LOW_CONFIDENCE (\< 0.85):

Context: The AI was unsure about the character shapes. This is common in documents with coffee stains, faint handwriting, or complex multi-column layouts.

Action: These files are "High Risk" for searchability. Do not rely on this text for indexing without human review.

POSSIBLE_PII (Privacy Risk):

Context: The script detected patterns matching Email Addresses or Social Security Numbers.

Action: Quarantine these files immediately. Open the original PDF to verify if the data is sensitive. If so, you must create a redacted derivative before making the dataset public.

Source Check (LOCAL_PDF_LAYOUT vs GOOGLE_DOC_AI):

Context: Files marked LOCAL_PDF_LAYOUT were read directly from the digital layer.

Action: These are usually 100% accurate. Focus your manual QC efforts on the files marked GOOGLE_DOC_AI, as these relied on visual pattern recognition and are more prone to "typos."

## Additional Tools & Resources

While daiR provides state-of-the-art accuracy, other tools may be better suited for specific constraints (e.g., cost or data sovereignty).

-   **Tesseract OCR:** Is an optical [character recognition engine](https://github.com/tesseract-ocr/tesseract). It is essential for "Content Analysis." It can scan thousands of images to detect text, helping curators flag files that contain sensitive documents (PII) which might have been mixed into a photo dataset.

-   **Apache Tika:** Is a [toolkit](https://tika.apache.org/) that detects and extracts metadata and text from over a thousand different file types (such as PPT, XLS, PDF). It is excellent for extracting text from "born-digital" files without OCR.

-   **Label Studio:** This [utility](https://labelstud.io/) assists curators with "Low Confidence" scans, this open-source tool allows the set up a "Human-in-the-Loop" workflow to manually correct the OCR output.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server, in a batch job, or from the command line, here is a pure R script that performs the same process.

**Prerequisites:** Set up the Google Cloud and account and corresponding OCR Document AI.

Download the **R Script:** [**`Inspect_OCR_Script.R`**](Scripts/Inspect_OCR_Script.R)

### Example HPC Submission Script

``` bash
#!/bin/bash
#SBATCH --job-name=ocr_curation
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=4
#SBATCH --mem=8G
#SBATCH --time=02:00:00
#SBATCH --output=logs/ocr_%j.log

module load R

# Define the target directory containing your documents
TARGET_DIR="/scratch/user/project_data/documents"

# Ensure service-account.json is in the current folder!
echo "Starting OCR Curation Pipeline..."
Rscript OCR_Curator_Script_Robust.R "$TARGET_DIR"
```

## References

::: {#refs}
:::
