---
title: "MATLAB (.mat) Files"
author: "Daniel Manrique-Castano"
date: "2025-12-18"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
bibliography: references.bib
params:
  target_dir: "data/Inspect_mat/"
---

## Overview

MAT-files are the proprietary binary data container format used by MATLAB. They are widely used in engineering and science but pose a significant challenge for long-term preservation.

::: {.callout-note title="Curation Goal"}
Identify file versions and inventory internal variables. Our objective is to distinguish between classic Level 5 (v5) files and modern Level 7.3 (HDF5-based) files, ensuring the correct open-source libraries are used for long-term access.
:::

::: {.callout-warning title="Preservation Risk"}
Proprietary format lock-in is a major risk. Version 7.3 files require specialized HDF5 parsers, while older versions are opaque without MATLAB or compatible open-source alternatives like GNU Octave.
:::

**This notebook inspects `.mat` files to:**

1.  **Identify Version:** Distinguish between v5 and v7.3 without crashing.
2.  **Inventory Content:** List variable names and array dimensions to understand dataset complexity.

------------------------------------------------------------------------

## Setup

This workflow relies on the `R.matlab` package.

### R Packages

If you don't have the packages installed, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "R.matlab", "hdf5r",  "rstudioapi"))

if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager"); BiocManager::install("rhdf5")
```

### Load libraries

```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(R.matlab)
library(hdf5r)
library(rstudioapi)
```

## Select Target Directory

We select the folder containing the `.mat` files.

**Note:** If running interactively, a dialog box will appear. Otherwise, it defaults to the target_dir parameter.

```{r}
#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select Notebook Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))

```

## Find MATLAB files

We scan the directory for files ending in .mat.

```{r}
#| label: find-files

ipynb_files <- list.files(
  path = target_dir,
  pattern = "\\.mat$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

print(paste("Found", length(ipynb_files), "MATLAB files."))
head(ipynb_files)
```

## Generate Metadata Report

We try to read each .mat file as HDF5 first. If that succeeds, we extract:

-   variable (dataset) names
-   data type (HDF5 class)
-   array dimensions
-   number of dimensions
-   number of elements

If the file is not HDF5, the code falls back to R.matlab::readMat() and extracts the same information from a v4/v5 structure.

```{r}
#| label: extraction-logic
#| warning: false
#| message: false

# --- Helper 1: Extract v5 Content (R.matlab) ---
extract_v5_content <- function(fp) {
  tryCatch({
    # readMat returns a simple list
    content <- R.matlab::readMat(fp)
    var_names <- names(content)
    
    # Map each variable to its dimensions and class
    map_dfr(var_names, function(v) {
      obj <- content[[v]]
      dims <- paste(dim(obj), collapse = "x")
      if (dims == "") dims <- paste(length(obj), "(len)") # Handle vectors
      
      tibble(
        object_path = v,
        object_type = "MATLAB Variable",
        dimensions = dims,
        data_class = class(obj)[1]
      )
    })
  }, error = function(e) {
    tibble(object_path = "Parse Error", object_type = "Error", dimensions = NA, data_class = e$message)
  })
}

# --- Helper 2: Extract v7.3 Content (rhdf5) ---
extract_v73_content <- function(fp) {
  tryCatch({
    # h5ls recursively lists the HDF5 hierarchy
    content <- rhdf5::h5ls(fp, all = TRUE)
    
    content %>%
      mutate(
        object_path = paste0(group, "/", name),
        object_path = str_replace_all(object_path, "//", "/") # Clean paths
      ) %>%
      select(
        object_path,
        object_type = otype,  # e.g., H5I_DATASET
        dimensions = dim,
        data_class = dclass     # e.g., H5T_FLOAT
      )
  }, error = function(e) {
    tibble(object_path = "HDF5 Error", object_type = "Error", dimensions = NA, data_class = e$message)
  })
}

# --- Main Processing Function ---
inspect_mat_file <- function(fp) {
  fname <- basename(fp)
  
  # 1. Header Check (Identify Version)
  con <- file(fp, "rb")
  header_raw <- tryCatch(readBin(con, "raw", n = 128), finally = close(con))
  header_txt <- rawToChar(header_raw[header_raw > 31 & header_raw < 127])
  
  is_hdf5 <- str_detect(header_txt, "HDF5") || str_detect(header_txt, "MATLAB 7.3")
  version_label <- if (is_hdf5) "v7.3 (HDF5)" else "v5 (Standard)"
  
  # 2. Branching Logic
  if (is_hdf5) {
    content_df <- extract_v73_content(fp)
  } else {
    content_df <- extract_v5_content(fp)
  }
  
  # 3. Add Metadata
  content_df %>%
    mutate(
      filename = fname,
      version = version_label
    ) %>%
    select(filename, version, everything())
}

# Run Analysis
mat_files <- list.files(target_dir, pattern = "\\.mat$", full.names = TRUE, recursive = TRUE)

message(paste("Processing", length(mat_files), "files..."))

if (length(mat_files) > 0) {
  results <- map_dfr(mat_files, inspect_mat_file)
} else {
  results <- tibble(filename = character(), version = character())
}

print(paste("Extracted", nrow(results), "objects from", length(unique(results$filename)), "files."))
head(results, 10)
```

## Exports results

```{r}
#| label: save-results

output_dir <- "Results/Inspect_mat"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

output_file <- file.path(output_dir, paste0("MAT_Deep_Report_", format(Sys.Date(), "%Y%m%d"), ".csv"))

write.csv(results, output_file, row.names = FALSE)
print(paste("Report saved to:", output_file))
```

## Curation Insights

Use the generated CSV report to guide your preservation actions:

-   **Format Obsolescence (v5 vs v7.3):** The version column identifies the file architecture. v7.3 are essentially HDF5 files. They require specialized libraries (rhdf5 in R, h5py in Python) to read. Consider creating a Codebook that maps the HDF5 paths (e.g., /group1/dataset_A) to their scientific meaning.

-   **Workspace Dumps:** Files with high object counts (\>50) and generic names (var1, temp, data_copy) are likely RAM demanding. These files request a data dictionary or a cleaned version containing only the final variables.

-   **Complex Structures (H5T_COMPOUND):** In v7.3 files, H5T_COMPOUND indicates complex nested tables (structs). These are difficult to convert to CSV. Recommend converting these specific tables to JSON or separate CSVs to preserve the hierarchy.

### Notes and limitations

MATLAB .mat files can vary significantly in format and complexity. We have some limitations to keep in mind:

-   Very large datasets may not be fully loaded and may be summarized based on sampling.
-   Deeply nested MATLAB structs or objects may require custom handling.
-   Character encoding in .mat files may differ across MATLAB versions.
-   Complex cell arrays can contain heterogeneous types that cannot be summarized numerically.

## Additional Tools

-   **rhdf5 (Bioconductor):** This [library](https://bioconductor.org/packages/release/bioc/html/rhdf5.html) allows R users to browse and read v7.3 MAT-files as if they were standard HDF5 archives.

-   **GNU Octave:** Is an [open-source](https://octave.org/) high-level language, largely compatible with MATLAB. It can read and write most MAT-files without a MATLAB license.

-   **HDFView:** This is a [visual tool](https://www.hdfgroup.org/download-hdfview/) to browse the contents of v7.3 MAT-files without needing code.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server, in a batch job, or from the command line, here is a pure R script that performs the same process.

Download the **R Script:** [**`Inspect_mat_Script.R`**](Scripts/Inspect_mat_Script.R)

### Example HPC Submission Script (`Inspect_mat_Script.sh`)

``` bash
#!/bin/bash
#SBATCH --job-name=matlab_inspect
#SBATCH --cpus-per-task=2
#SBATCH --mem=8G
#SBATCH --time=01:00:00
#SBATCH --output=matlab_inspect_%j.out
#SBATCH --error=matlab_inspect_%j.err

module load R/4.3.0   # adapt to your cluster's module system

Rscript matlab_inspector.R \
  --target_dir /path/to/your/mat/files \
  --output_dir /path/to/output/folder \
  --max_full_numeric 1000000 \
  --max_sample_numeric 100000
  
```

## References

::: {#refs}
:::