---
title: "TIFF (.tif, .tiff) Images"
author: "Data Curation Team"
date: "2025-12-18"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: cosmo
params:
  target_dir: "data/Inspect_tiff/"
bibliography: references.bib
---

## Overview

The **Tag Image File Format (TIFF)** is the [dominant format](https://www.loc.gov/preservation/digital/formats/fdd/fdd000022.shtml) for scientific imaging (microscopy, astronomy, geospatial). Different to other image formats, TIFF images can hold:

-   **High Bit-Depths:** 16-bit or 32-bit floating-point data (common in sensors), which often appears "black" on standard monitors.
-   **Multi-Page Stacks:** A single `.tif` file can contain hundreds of individual images (e.g., a time-lapse recording or a 3D scan).
-   **Geo-Metadata:** Embedded coordinates (GeoTIFF) for satellite imagery.

**Hidden Compression** is a major preservation risk for TIFF images. While rhis format is generally considered lossless, it allows different internal compressions like the **JPEG compression**. This implies that a curator might archive a `.tif` file thinking it is pristine, only to discover later it contains lossy artifacts. In the same vein, "Multi-page" TIFFs can be often mistaken for single images if not properly inspected.

Furthermore, scientific software (like ImageJ/Fiji) often stores critical metadata (like "slices" or "physical pixel size") in generic "Description" tags rather than standard fields, making it invisible to standard photo viewers.

This notebook employs a **"Nuclear" Multi-Strategy** approach to ensure no metadata is missed:

1.  **LibTIFF Extraction:** Queries the raw C-level file header (IFD) for standard tags like Compression Code.
2.  **ImageMagick Fallback:** Uses the robust ImageMagick engine to infer properties when headers are non-standard.
3.  **Text Mining:** Scans the generic "Description" fields using Regular Expressions to find hidden ImageJ metadata (e.g., `slices=50`).

------------------------------------------------------------------------

## Setup

We use the `ijtiff` package, which is designed for scientific image handling and correctly processes multi-page and high-bit-depth files.

### R Packages

If you do not have the required packages, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "tiff", "magick", "stringr", "rstudioapi"))
```

### Load libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(tiff)       # Raw LibTIFF
library(magick)     # ImageMagick
library(stringr)    # Text Mining
library(rstudioapi) # UI interaction
```

## Select Target Directory

Select the folder containing the TiFF files.

**Note:** If running interactively, a dialog box will appear. Otherwise, it defaults to the target_dir parameter.

```{r}
#| label: select-target-dir

if (interactive() && .Platform$OS.type == "windows") {
  selected_dir <- rstudioapi::selectDirectory(caption = "Select TIFF Directory")
} else {
  selected_dir <- NULL
}

if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))

```

## Inventory of files

We scan for `.tif` and `.tiff` files.

```{r}
#| label: inventory

tiff_files <- list.files(
  path = target_dir,
  pattern = "\\.tiff?$", 
  recursive = TRUE, 
  full.names = TRUE, 
  ignore.case = TRUE
)

print(paste("Found", length(tiff_files), "TIFF files."))
```

## Deep Header Inspection

This function scans all file attributes for keywords (e.g., "depth", "compression") to find metadata even if it uses non-standard tag names.

```{r}
#| label: inspection-logic
#| warning: false
#| message: false

message("Generating Deep Inspection Report...")

inspect_tiff_fuzzy <- function(fp) {
  fname <- basename(fp)
  
  tryCatch({
    # Read Image Header
    img <- image_read(fp)
    info <- image_info(img)
    attrs <- image_attributes(img)
    
    # Fuzzy Extraction: Bit Depth
    # Find ANY attribute containing "depth" or "bits" (case insensitive)
    depth_check <- attrs %>% 
      filter(grepl("depth|bits", property, ignore.case = TRUE)) %>% 
      pull(value)
    
    # Logic: Use found attribute, or fallback to info$depth, or "Unknown"
    final_depth <- if(length(depth_check) > 0) {
      paste(unique(depth_check), collapse = "/") 
    } else if ("depth" %in% names(info)) {
      as.character(info$depth[1])
    } else {
      "Unknown"
    }

    # Fuzzy Extraction: Compression
    # Find ANY attribute containing "compression"
    comp_check <- attrs %>% 
      filter(grepl("compression", property, ignore.case = TRUE)) %>% 
      pull(value)
      
    final_comp <- if(length(comp_check) > 0) comp_check[1] else "Unknown"

    # Fuzzy Extraction: Resolution
    res_check <- attrs %>% 
      filter(grepl("resolution|density", property, ignore.case = TRUE)) %>% 
      pull(value)
      
    final_res <- if(length(res_check) > 0) paste(res_check[1], "DPI") else paste(info$density[1], "DPI")

    # Build Row
    tibble(
      FileName = fname,
      Dimensions = paste(info$width[1], "x", info$height[1]),
      BitDepth = final_depth,
      Compression = final_comp,
      ColorSpace = info$colorspace[1],
      Resolution = final_res,
      FileSize_MB = round(file.size(fp) / 1024^2, 2),
      Status = "Success"
    )
    
  }, error = function(e) {
    # Error Handling
    message(paste("Failed on:", fname, "-", e$message))
    tibble(
      FileName = fname, Dimensions = NA, BitDepth = NA, Compression = NA, 
      ColorSpace = NA, Resolution = NA, FileSize_MB = NA,
      Status = paste("Error:", e$message)
    )
  })
}

# Execute Analysis
report <- map_dfr(tiff_files, inspect_tiff_fuzzy)

# Display
print("--- Deep Inspection Preview ---")
print(head(report))
```

## Save Results

Save the report to a CSV file for review.

```{r}
#| label: save-results

output_dir <- "Results/Inspect_tiff"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

output_file <- file.path(output_dir, paste0("TIFF_DeepScan_", format(Sys.Date(), "%Y%m%d"), ".csv"))

write.csv(report, output_file, row.names = FALSE)
print(paste("Report saved to:", output_file))
```

## Curation Insights

Use this report to verify scientific integrity:

-   **Bit Depth (16 or 32):** These are true scientific images. Do not convert them to JPG or PNG, as those formats are typically limited to 8-bit and will discard data precision. If BitDepth remains "Unknown", the file likely lacks a proper header.

-   **Compression:** Recommended (Uncompressed, LZW, Deflate, Packbits); Not recommended (JPEG, OldJPEG). If you see JPEG compression inside a TIFF, the data integrity for pixel-based analysis is compromised.

-   **Color Space (Gray vs sRGB):** Scientific microscopy data is typically Gray (intensity only). If a raw microscope file is sRGB, it may have been converted or screenshotted, potentially losing its dynamic range.

## Additional Tools

-   **ImageJ / Fiji:** The absolute standard [open-source software](https://imagej.net/software/fiji/downloads) for viewing and analyzing scientific TIFFs (microscopy, astronomy). It handles multi-page stacks and high-bit-depth data that confuse standard viewers [@schindelin2012].

-   **Bio-Formats:** A [library](https://www.openmicroscopy.org/bio-formats/) specifically designed to read proprietary life sciences image formats (e.g., .czi, .lif) and convert them to standard OME-TIFF (see the [Bio-Formats](https://bio-formats.readthedocs.io/en/v8.3.0/about/index.html) documentation).

-   **ExifTool:** A robust [command-line tool](https://exiftool.org/) for reading and writing embedded metadata (EXIF, IPTC, XMP) in images (see the [documentation](https://exiftool.org/exiftool_pod.html)).

## Using the Non-Interactive R Script

For users who want to run this analysis on a server (HPC), in a batch job, or from the command line, here is the pure R script version.

Download the **R Script:** [**`Inspect_tiff_Script.R`**](Scripts/Inspect_tiff_Script.R)

### Example HPC Submission Script (`Inspect_tiff_submit.sh`)

``` bash
#!/bin/bash
#SBATCH --job-name=tiff_check
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:30:00
#SBATCH --mem=16G
#SBATCH --output=logs/tiff_check_%j.log

# Load R Module
# Note: ImageMagick is often a system library, but some clusters require a module.
# Check with 'module avail imagemagick' if the script fails.
module load R
module load imagemagick/7.1  # Adjust version as needed for your cluster

# Define Target Directory
# Replace with the actual path to your microscopy/image data
DATA_DIR="/scratch/user/project_data/microscopy"

# Prepare Environment
mkdir -p Results/Inspect_tiff
mkdir -p logs

# Run Analysis
echo "Starting TIFF Deep Scan on $DATA_DIR"
Rscript Inspect_tiff_Script.R "$DATA_DIR"
```

## References

::: {#refs}
:::
