---
title: "JSON Files"
author: "Daniel Manrique-Castano"
date: "2025-11-19"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
bibliography: references.bib
params:
  target_dir: "data/Inspect_json/"
---

## Overview

**JSON (JavaScript Object Notation)** has become the de standard for data interchange on the web and in NoSQL databases, replacing XML due to its lightweight and human-readable nature. In data curation, JSON is frequently used for metadata, configuration settings, and hierarchical datasets.

Unlike tabular data (CSV), JSON is hierarchical and semi-structured. A file might appear valid but contain deeply nested structures that make it difficult to flatten or analyze. Pezoa et al. (2016) highlight that while JSON syntax is simple, the lack of a mandatory schema often leads to "structural heterogeneity," where different records in the same dataset have different fields [@pezoa2016].

This notebook evaluates JSON files on three levels:

1.  **Syntactic Validity:** Can the parser read the file, or are there missing brackets/commas?
2.  **Structural Complexity:** We calculate the **Nesting Depth** and **Key Count**. High depth values often indicate complex objects that require specialized flattening tools.
3.  **Data Type:** Distinguishing between **Objects** (single records) and **Arrays** (lists of records).

------------------------------------------------------------------------

## Setup

This workflow relies on `jsonlite` [@jsonlite] for parsing and `tidyverse` [@tidyverse-2] for data manipulation.

### R Packages

If you don't have these packages installed, run this command once in your R console:

```{r}
# install.packages(c("tidyverse", "jsonlite", "purrr", "rstudioapi"))

```

## Load libraries

```{r}

#| label: load-libraries
#| message: false

library(tidyverse)
library(jsonlite)   # Standard parser
library(purrr)      # For robust iteration
library(rstudioapi) # For directory selection
```

## Select Target Directory

Select the directory containing your .json files.

Note: Interactive selection works in RStudio. If rendering, it defaults to params\$target_dir.

```{r}

#| label: select-target-dir

# 1. Try to select interactively if in RStudio
if (interactive() && .Platform$OS.type == "windows") { 
  selected_dir <- rstudioapi::selectDirectory(caption = "Select JSON Directory")
} else {
  selected_dir <- NULL
}

# 2. Logic to determine final directory
if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}

print(paste("Analyzing directory:", target_dir))
```

## Find Json files

```{r}
#| label: find-files

# Find all .json files
json_files <- list.files(
  path = target_dir,
  pattern = "\\.json$", 
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

print(paste("Found", length(json_files), "JSON files."))
head(json_files)
```

## Inventory & Extraction

We iterate through the JSON files to determine their R-structure (Data Frame vs. List) and complexity.

```{r}
#| label: extraction-logic
#| warning: false
#| message: false

# Helper: Calculate Nesting Depth recursively
get_depth <- function(x) {
  if (is.list(x) && length(x) > 0) {
    1 + max(vapply(x, get_depth, numeric(1)), 0)
  } else {
    0
  }
}

# Initialize results list
json_summary_list <- list()

# Loop through files
for (file_path in json_files) {
  file_name <- basename(file_path)
  
  tryCatch({
    # Attempt to parse the JSON
    # simplifyVector = TRUE allows R to detect tables automatically
    json_data <- jsonlite::fromJSON(file_path, simplifyVector = TRUE)
    
    # Calculate Complexity (Depth)
    # We re-parse without simplification just to check the raw tree depth accurately
    raw_data <- jsonlite::fromJSON(file_path, simplifyVector = FALSE)
    depth_val <- get_depth(raw_data)
    
    # Determine Class (Data Frame, List, or Vector)
    obj_class <- class(json_data)[1]
    
    # Determine Dimensions & Keys based on type
    if (is.data.frame(json_data)) {
      dims_str <- paste(dim(json_data), collapse = " x ")
      keys_str <- paste(head(colnames(json_data), 5), collapse = ", ")
      if (ncol(json_data) > 5) keys_str <- paste(keys_str, "...")
      
    } else if (is.list(json_data)) {
      dims_str <- paste(length(json_data), "elements")
      keys_str <- paste(head(names(json_data), 5), collapse = ", ")
      if (length(json_data) > 5) keys_str <- paste(keys_str, "...")
      
    } else {
      # Atomic vectors or arrays
      dims_str <- paste(length(json_data), "length")
      keys_str <- "(No Keys - Flat Array)"
    }
    
    # Store success result
    json_summary_list[[length(json_summary_list) + 1]] <- tibble(
      filename = file_name,
      is_valid = TRUE,
      structure_type = obj_class,
      dimensions = dims_str,
      top_level_keys = keys_str,
      max_depth = depth_val,
      error_msg = NA_character_
    )
    
  }, error = function(e) {
    # Store error result
    json_summary_list[[length(json_summary_list) + 1]] <- tibble(
      filename = file_name,
      is_valid = FALSE,
      structure_type = "Error",
      dimensions = "NA",
      top_level_keys = "NA",
      max_depth = NA_real_,
      error_msg = e$message
    )
  })
}

# Combine Results
results <- bind_rows(json_summary_list)

print(paste("Analyzed", nrow(results), "JSON files."))
head(results)
```

## Visualization: Data Types

We visualize the distribution of JSON structures. A prevalence of data.frame types indicates a dataset that is easy to analyze in R, while list types imply nested metadata.

```{r}
#| label: visual-json
#| fig-cap: "Distribution of JSON Structure Types"

ggplot(results %>% filter(is_valid), aes(x = structure_type, fill = structure_type)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "JSON Structure Types",
    subtitle = "Identified R Objects (Data Frame = Tabular, List = Nested)",
    x = "Structure Type",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Export Results

```{r}
#| label: save-results

output_dir <- "Results/Inspect_json"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

output_file <- file.path(output_dir, paste0("JSON_Report_", format(Sys.Date(), "%Y%m%d"), ".csv"))

write.csv(results, output_file, row.names = FALSE)
print(paste("Report saved to:", output_file))
```

## Curation Insights

Use the generated CSV report to guide your preservation actions:

-   **Validity Check (is_valid):** Files marked FALSE are corrupt. Common causes include "Trailing Commas" (allowed in JavaScript but forbidden in standard JSON) or truncation (incomplete download). You can validate the syntax using a linter. If truncated, attempt to recover the source. If the error is encoding-related (e.g., "Invalid char"), check if the file is saved as UTF-8.

-   **Structure Check (structure_type):** Files identified as list often contain nested hierarchies that cannot be easily flattened into a table. Ensure these files are documented with a data dictionary explaining the nesting logic.

-   **Complexity Check (max_depth \> 5):** Deeply nested files are difficult for future researchers to query without specialized tools. Curation best practice suggests generating a "JSON Schema" to validate that all files in the batch adhere to the same structure.

## Additional Tools

-   **jq:** Is a lightweight and flexible [command-line JSON processor](https://jqlang.org/), known as the industry standard for slicing, filtering, and transforming JSON data.

-   **JSON Schema Validator:** Is a [tool](https://json-schema.org/) to define the expected structure of your JSON (required fields, data types), allowing curators to automate the quality control of metadata.

Source: Pezoa, F. et al. (2016). Foundations of JSON Schema.

-   **VS Code (JSON Tools):** [Visual Studio Code](https://code.visualstudio.com/docs/languages/json) has built-in formatters that can automatically fix minor syntax errors (like indentation) and highlight invalid characters.

## Using the Non-Interactive R Script

For users who want to run this analysis on a server, in a batch job, or from the command line, here is a pure R script that performs the same process.

Download the **R Script:** [**`Inspect_json_Script.R`**](Scripts/Inspect_json_Script.R)

-   **For clusters**

Alternatively, adapt the folling code script to match your cluster's environment and your data path, then submit it using `sbatch`.

``` bash
#!/bin/bash
#SBATCH --job-name=json_inspect
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:15:00
#SBATCH --mem=4G
#SBATCH --output=logs/inspect_json_%j.log

# Load R module (Adjust based on your cluster's specific module name)
module load R

# Define target directory containing JSON files
TARGET_DIR="/scratch/user/project_data/metadata"

# Create output directory structure if it doesn't exist
mkdir -p Results/Inspect_json
mkdir -p logs

# Run the R script
echo "Starting JSON Inspection on $TARGET_DIR"
Rscript Scripts/Inspect_json_Script.R "$TARGET_DIR"
```


## References

::: {#refs}
:::

