---
title: "OME-Zarr Image Stores (.zarr, .tar)"
author: "Data Curation Team"
date: "2026-10-06"
format:
  html:
    toc: true
    toc-location: left
    code-fold: true
    theme: cosmo
params:
  target_dir: "data/Inspect_OMEZarr/"
bibliography: references.bib
---

```{r}
#| label: eval-gate
#| include: false

knitr::opts_chunk$set(eval = dir.exists(params$target_dir))
```

## Overview // Aperçu

**Zarr** stores chunked, compressed arrays as a tree of small files, and **OME-Zarr** (OME-NGFF) adds a metadata convention for bioimaging: axes, units, voxel scale, channels, and multiple resolution levels. Large lightsheet and microscopy datasets are often handed to a repository as `<name>.zarr` folders, or packed into uncompressed `.tar` files so that thousands of small chunk files travel as one.

//

**Zarr** stocke des tableaux fragmentés et compressés sous forme d'arborescence de petits fichiers, et **OME-Zarr** (OME-NGFF) y ajoute une convention de métadonnées pour la bio-imagerie : axes, unités, échelle des voxels, canaux et niveaux de résolution multiples. Les grands jeux de données de microscopie à feuillet de lumière sont souvent remis à un dépôt sous forme de dossiers `<nom>.zarr`, ou regroupés dans des fichiers `.tar` non compressés afin que des milliers de petits fichiers de fragments voyagent en un seul.

::: {.callout-note title="Curation Goal // Objectif de la curation"}
Check the **structure and metadata** of each store without decompressing anything: whether all expected chunks are present, how each array is described, whether the OME-NGFF metadata is complete and consistent, and whether the container is labelled correctly.
//
Vérifier la **structure et les métadonnées** de chaque magasin sans rien décompresser : si tous les fragments attendus sont présents, comment chaque tableau est décrit, si les métadonnées OME-NGFF sont complètes et cohérentes et si le conteneur est correctement étiqueté.
:::

::: {.callout-warning title="Identifying Risks // Identification des risques"}
Missing or empty chunks make part of an image unreadable. Template text left in the metadata, or acquisition file names that carry a different subject code than the container they were packed into, point to mislabelled data that a file-name check cannot see. Tar headers can also carry a user account name.
//
Des fragments manquants ou vides rendent une partie d'une image illisible. Du texte de gabarit laissé dans les métadonnées, ou des noms de fichiers d'acquisition portant un code de sujet différent de celui du conteneur dans lequel ils ont été regroupés, signalent des données mal étiquetées qu'une vérification des noms de fichiers ne peut pas voir. Les en-têtes tar peuvent aussi contenir un nom de compte d'utilisateur.
:::

**What the notebook reads.** For tars, one listing pass plus one extraction pass of the small JSON metadata files; the chunk payloads are skipped by seeking, never read. This keeps it practical on tars of tens of gigabytes. For each store and resolution level it reports:

1.  **Layout and completeness:** The expected chunk grid is rebuilt from `.zarray` (shape and chunks) and compared with the chunk files present, handling both `.` and `/` chunk-key separators.
2.  **Array description:** Shape, chunk shape, data type, compressor, and fill value.
3.  **OME-NGFF metadata (`.zattrs`):** Axes, units, voxel scale, channel labels, orientation, and version.
4.  **BIDS-Microscopy sidecar and leftover template text:** Acquisition file names, pixel size, and boilerplate such as "Author Name 1" or "TODO", plus subject-ID consistency.
5.  **Container hygiene:** Whether the tar headers carry a named user or group, recorded as a flag and a count, **not** the account name.

**By design, chunk payloads are not decoded** (no blosc, lz4 or zstd decoding), so pixel values are not checked.

//

**Ce que le cahier lit.** Pour les fichiers tar, un passage de listage plus un passage d'extraction des petits fichiers de métadonnées JSON ; les données des fragments sont sautées par déplacement dans le fichier, jamais lues. Cela reste praticable sur des fichiers tar de plusieurs dizaines de gigaoctets. Pour chaque magasin et chaque niveau de résolution, il indique :

1.  **Disposition et complétude :** La grille de fragments attendue est reconstruite à partir de `.zarray` (forme et fragments) et comparée aux fichiers de fragments présents, en gérant les séparateurs de clés `.` et `/`.
2.  **Description du tableau :** Forme, forme des fragments, type de données, compresseur et valeur de remplissage.
3.  **Métadonnées OME-NGFF (`.zattrs`) :** Axes, unités, échelle des voxels, noms des canaux, orientation et version.
4.  **Fichier annexe BIDS-Microscopy et texte de gabarit résiduel :** Noms des fichiers d'acquisition, taille des pixels et texte type comme « Author Name 1 » ou « TODO », ainsi que la cohérence de l'identifiant du sujet.
5.  **Hygiène du conteneur :** Si les en-têtes tar contiennent un nom d'utilisateur ou de groupe, consigné sous forme d'indicateur et de nombre, **et non** le nom du compte.

**Par conception, les données des fragments ne sont pas décodées** (aucun décodage blosc, lz4 ou zstd) ; les valeurs des pixels ne sont donc pas vérifiées.

::: {.callout-tip title="Sample Data // Données d'exemple"}
No sample data is bundled with this notebook. If the folder `data/Inspect_OMEZarr/` does not exist, the code is displayed but not run. Place your own files there (or change `target_dir` in the YAML header) and render again to see the results.
//
Aucune donnée d'exemple n'est fournie avec ce cahier. Si le dossier `data/Inspect_OMEZarr/` n'existe pas, le code est affiché mais n'est pas exécuté. Placez-y vos propres fichiers (ou modifiez `target_dir` dans l'en-tête YAML) et générez à nouveau le document pour voir les résultats.
:::

------------------------------------------------------------------------

## Setup // Configuration

We use `jsonlite` to read the Zarr and OME-NGFF metadata [@jsonlite]. Tars are listed with the `tar` program found on your `PATH` (GNU tar or Windows `bsdtar`). Set the `TAR_PATH` environment variable to override it.

//

Nous utilisons `jsonlite` pour lire les métadonnées Zarr et OME-NGFF [@jsonlite]. Les fichiers tar sont listés avec le programme `tar` trouvé dans votre `PATH` (GNU tar ou `bsdtar` de Windows). Définissez la variable d'environnement `TAR_PATH` pour le remplacer.

### R Packages // Packages R

```{r}
#| label: install-packages
#| eval: false

# install.packages(c("tidyverse", "jsonlite", "rstudioapi"))
```

### Load Libraries // Charger les bibliothèques

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)  # Data manipulation [@tidyverse]
library(jsonlite)   # Reading JSON metadata [@jsonlite]
library(rstudioapi) # Interactive folder selection
```

## Select Target Directory // Sélectionner le répertoire cible

This block allows for interactive selection of the directory. If running in a non-interactive environment, it defaults to the path defined in the YAML header.

//

Ce bloc permet de sélectionner le répertoire de manière interactive. En cas d'exécution dans un environnement non interactif, le chemin par défaut est celui défini dans l'en-tête YAML.

```{r}
#| label: select-target-dir

if (interactive() && .Platform$OS.type == "windows") {
  selected_dir <- rstudioapi::selectDirectory(caption = "Select Directory with OME-Zarr stores or tars")
} else {
  selected_dir <- NULL
}

if (!is.null(selected_dir)) {
  target_dir <- selected_dir
} else {
  target_dir <- params$target_dir
}
target_dir <- normalizePath(target_dir, winslash = "/", mustWork = TRUE)
dir_label <- gsub("[^A-Za-z0-9_.-]", "_", basename(sub("[/\\\\]+$", "", target_dir)))
output_dir <- "Results/Inspect_OMEZarr"

print(paste("Analyzing directory:", target_dir))
```

## Tar Tool // Outil tar

This block locates the `tar` program and detects whether it is GNU tar, which needs the `--force-local` option for paths such as `D:/...`.

//

Ce bloc localise le programme `tar` et détecte s'il s'agit de GNU tar, qui exige l'option `--force-local` pour les chemins comme `D:/...`.

```{r}
#| label: tar-tool

tar_bin <- Sys.getenv("TAR_PATH", unset = "")
if (tar_bin == "") tar_bin <- Sys.which("tar")[[1]]
have_tar <- nzchar(tar_bin)
is_gnu_tar <- FALSE
if (have_tar) {
  ver <- tryCatch(system2(tar_bin, "--version", stdout = TRUE, stderr = TRUE), error = function(e) "")
  is_gnu_tar <- any(grepl("GNU tar", ver))
}
tar_args <- function(...) c(if (is_gnu_tar) "--force-local", ...)
```

## Inventory of Containers // Liste des conteneurs

We scan for `.tar` files and for `.zarr` folders. Folders named `Curation_Results` are skipped.

//

Nous recherchons les fichiers `.tar` et les dossiers `.zarr`. Les dossiers nommés `Curation_Results` sont ignorés.

```{r}
#| label: inventory

tar_files <- list.files(target_dir, pattern = "\\.tar$", recursive = TRUE, full.names = TRUE, ignore.case = TRUE)
tar_files <- tar_files[!grepl("Curation_Results", tar_files, ignore.case = TRUE)]
zarr_dirs <- list.dirs(target_dir, recursive = TRUE, full.names = TRUE)
zarr_dirs <- zarr_dirs[grepl("\\.zarr$", zarr_dirs, ignore.case = TRUE) & !grepl("Curation_Results", zarr_dirs, ignore.case = TRUE)]

message(sprintf("Found %d tar file(s) and %d on-disk .zarr folder(s).", length(tar_files), length(zarr_dirs)))
if (length(tar_files) == 0 && length(zarr_dirs) == 0) {
  stop("No .tar files or .zarr folders found.", call. = FALSE)
}
if (length(tar_files) > 0 && !have_tar) stop("No `tar` executable found on PATH; set TAR_PATH.", call. = FALSE)
```

## Helper Functions // Fonctions auxiliaires

These functions list a tar or an on-disk store, extract only the small metadata files, read JSON safely, and rebuild the expected chunk keys.

//

Ces fonctions listent un fichier tar ou un magasin sur disque, extraient uniquement les petits fichiers de métadonnées, lisent le JSON de façon sûre et reconstruisent les clés de fragments attendues.

```{r}
#| label: helpers

# Parse one line of `tar -tv` output (GNU or bsdtar) into path/size/owner.
parse_tar_line <- function(x) {
  gnu <- str_match(x, "^(\\S{10})\\s+(\\S+)\\s+(\\d+)\\s+\\d{4}-\\d{2}-\\d{2}\\s+\\d{2}:\\d{2}(?::\\d{2})?\\s+(.*)$")
  bsd <- str_match(x, "^(\\S{10})\\s+\\d+\\s+(\\S+)\\s+(\\S+)\\s+(\\d+)\\s+\\w{3}\\s+\\d{1,2}\\s+(?:\\d{2}:\\d{2}|\\d{4})\\s+(.*)$")
  tibble(
    is_dir = substr(x, 1, 1) == "d",
    owner = ifelse(!is.na(gnu[, 3]), gnu[, 3], paste(bsd[, 3], bsd[, 4], sep = "/")),
    size = as.numeric(ifelse(!is.na(gnu[, 4]), gnu[, 4], bsd[, 5])),
    path = ifelse(!is.na(gnu[, 5]), gnu[, 5], bsd[, 6])
  )
}

list_tar <- function(fp) {
  out <- system2(tar_bin, c(tar_args("-tvf", shQuote(fp))), stdout = TRUE, stderr = FALSE)
  parse_tar_line(out) %>% filter(!is.na(path))
}

extract_meta <- function(fp, dest) {
  dir.create(dest, recursive = TRUE, showWarnings = FALSE)
  pats <- c("*.json", "*.tsv", "*.md", "*.zattrs", "*.zgroup", "*.zarray")
  system2(tar_bin, c(tar_args("-xf", shQuote(fp), "-C", shQuote(dest), if (is_gnu_tar) "--wildcards", shQuote(pats))),
          stdout = FALSE, stderr = FALSE)
}

list_disk_zarr <- function(root) {
  fl <- list.files(root, recursive = TRUE, all.files = TRUE, full.names = TRUE, include.dirs = FALSE)
  fi <- file.info(fl)
  parent <- dirname(root)
  tibble(is_dir = FALSE, owner = NA_character_, size = fi$size,
         path = sub(paste0("^", gsub("([.|()\\^{}+$*?]|\\[|\\])", "\\\\\\1", parent), "/"), "", fl))
}

read_json_safe <- function(p) {
  if (!file.exists(p)) return(NULL)
  tryCatch(fromJSON(p, simplifyVector = FALSE), error = function(e) NULL)
}

read_text_safe <- function(p) {
  if (!file.exists(p)) return(character(0))
  tryCatch(readLines(p, warn = FALSE, encoding = "UTF-8"), error = function(e) character(0))
}

# Expected chunk keys from shape/chunks; keys can be "0.12.0.0" or "0/12/0/0".
expected_chunk_keys <- function(shape, chunks, sep) {
  ranges <- map2(shape, chunks, ~ seq_len(ceiling(.x / .y)) - 1L)
  grid <- do.call(expand.grid, c(ranges, KEEP.OUT.ATTRS = FALSE))
  do.call(paste, c(as.list(grid), sep = sep))
}

placeholder_hits <- function(desc, readme_lines) {
  hits <- character(0)
  if (!is.null(desc)) {
    if (identical(desc$Name, "Name of the dataset")) hits <- c(hits, "Name")
    if (!is.null(desc$License) && grepl("^The license for the dataset$", paste(desc$License, collapse = ""))) hits <- c(hits, "License")
    au <- unlist(desc$Authors)
    if (length(au) > 0 && any(grepl("^Author Name [0-9]+$", au))) hits <- c(hits, "Authors")
  }
  if (any(grepl("^\\s*-\\s*\\[ \\]|TODO", readme_lines))) hits <- c(hits, "README_TODO")
  paste(hits, collapse = "; ")
}
```

## Per-Store Inspection // Inspection de chaque magasin

This function produces one row per store and resolution level, combining the structure, the array description, the metadata, and the consistency checks.

//

Cette fonction produit une ligne par magasin et par niveau de résolution, en combinant la structure, la description du tableau, les métadonnées et les contrôles de cohérence.

```{r}
#| label: store-inspection

# `listing` has one row per archive entry (path relative to the tar root, or to
# the parent of the .zarr folder on disk); `meta_root` is where metadata files can
# be read (extracted tar dir, or the dataset dir for on-disk stores).
inspect_container <- function(container_label, container_path, listing, meta_root, is_tar) {
  tar_base <- sub("\\.tar$", "", basename(container_path), ignore.case = TRUE)
  files <- listing %>% filter(!is_dir)
  stores <- unique(sub("(\\.zarr)/.*$", "\\1", files$path[grepl("\\.zarr/", files$path, ignore.case = TRUE)]))
  top_folder <- unique(sub("/.*$", "", listing$path))
  n_owner <- if (is_tar) n_distinct(na.omit(listing$owner)) else NA_integer_
  named_owner <- if (is_tar) any(!is.na(listing$owner) & !grepl("^(root|0|nobody)(/|$)", listing$owner)) else NA

  if (length(stores) == 0) {
    return(tibble(Container = container_label, Store = NA_character_, Level = NA_character_,
                  Status = "No .zarr store found in container"))
  }

  subject_dir <- str_extract(container_path, "LD[A-Za-z]*\\d+|sub-[A-Za-z0-9]+")
  map_dfr(stores, function(st) {
    st_files <- files %>% filter(startsWith(path, paste0(st, "/")))
    rel <- substring(st_files$path, nchar(st) + 2)
    levels <- sort(unique(sub("/.*$", "", rel[grepl("/", rel)])))
    levels <- levels[grepl("^[0-9]+$", levels)]

    zattrs <- read_json_safe(file.path(meta_root, st, ".zattrs"))
    ms <- if (!is.null(zattrs$multiscales)) zattrs$multiscales[[1]] else NULL
    axes <- if (!is.null(ms$axes)) map_chr(ms$axes, ~ .x$name) else character(0)
    units <- if (!is.null(ms$axes)) map_chr(ms$axes, ~ if (is.null(.x$unit)) "-" else .x$unit) else character(0)
    ds_paths <- if (!is.null(ms$datasets)) map_chr(ms$datasets, ~ .x$path) else character(0)
    ch <- zattrs$omero$channels
    ch_labels <- if (!is.null(ch)) map_chr(ch, ~ if (is.null(.x$label)) NA_character_ else .x$label) else character(0)

    # Sidecar next to the store: <store base>.json  (BIDS-Microscopy _SPIM.json)
    sidecar_path <- file.path(meta_root, sub("\\.ome\\.zarr$|\\.zarr$", ".json", st, ignore.case = TRUE))
    side <- read_json_safe(sidecar_path)
    tile_cfg <- if (!is.null(side$ExtraMetadata$TileConfiguration$TileConfiguration)) side$ExtraMetadata$TileConfiguration$TileConfiguration else ""
    # TileConfiguration is "<n>  <file>;;(x, y)  <file>;;(x, y) ...". File names can
    # contain spaces ("[00 x 01]"), so split on the ";;(x, y)" coordinate markers.
    tile_names <- trimws(str_split(tile_cfg, ";;\\([^)]*\\)")[[1]])
    tile_names <- trimws(sub("^\\d+\\s+", "", tile_names))
    tile_names <- tile_names[nzchar(tile_names)]
    # Only names that begin with a time-of-day prefix (hh-mm-ss_) carry a sample code;
    # bare names such as "Z0000.ome.tif" (one file per z-slice) do not.
    coded <- tile_names[grepl("^\\d{2}-\\d{2}-\\d{2}_", tile_names)]
    tile_stem <- sub("^\\d{2}-\\d{2}-\\d{2}_", "", unique(coded))
    # Everything before "_Blaze" (or the first "[") is the sample/stain descriptor,
    # e.g. "LDH16_cfos_left". A sample code is not always one token ("LD_H18_cfos_left"),
    # so the ID check compares on the digits found in the descriptor.
    tile_desc <- unique(sub("_?Blaze.*$|\\[.*$", "", tile_stem))
    tile_code <- unique(unlist(str_extract_all(tile_desc, "\\d+")))
    tile_style <- if (length(tile_names) == 0) NA_character_ else if (length(coded) == 0) "generic (no sample code in names)" else if (length(coded) == length(tile_names)) "sample-coded" else "mixed"
    subj_id <- str_extract(st, "sub-[A-Za-z0-9]+")
    subj_num <- str_extract(ifelse(is.na(subj_id), container_path, subj_id), "\\d+")
    id_mismatch <- if (length(tile_code) == 0 || is.na(subj_num)) NA else !(as.integer(subj_num) %in% as.integer(tile_code))
    pix <- if (!is.null(side$PixelSize)) paste(unlist(side$PixelSize), collapse = " x ") else NA_character_

    # Dataset-level template files at the tar root
    root_prefix <- if (is_tar) top_folder[1] else NULL
    desc <- read_json_safe(if (is_tar) file.path(meta_root, root_prefix, "dataset_description.json") else file.path(dirname(meta_root), "none"))
    rd <- if (is_tar) read_text_safe(file.path(meta_root, root_prefix, "README.md")) else character(0)
    placeholders <- if (is_tar) placeholder_hits(desc, rd) else NA_character_
    samples <- if (is_tar) {
      p <- file.path(meta_root, root_prefix, "samples.tsv")
      if (file.exists(p)) tryCatch(read_tsv(p, show_col_types = FALSE), error = function(e) NULL) else NULL
    } else NULL
    samples_ok <- if (!is.null(samples) && "participant_id" %in% names(samples) && !is.na(subj_id)) all(samples$participant_id == subj_id) else NA

    map_dfr(if (length(levels) == 0) NA_character_ else levels, function(lv) {
      lv_files <- if (is.na(lv)) st_files[0, ] else st_files %>% filter(startsWith(path, paste0(st, "/", lv, "/")))
      za <- if (is.na(lv)) NULL else read_json_safe(file.path(meta_root, st, lv, ".zarray"))
      chunk_files <- lv_files %>% mutate(key = substring(path, nchar(st) + nchar(lv) + 3)) %>%
        filter(!grepl("(^|/)\\.z(array|attrs|group)$", key))
      sep <- if (nrow(chunk_files) > 0 && any(grepl("/", chunk_files$key))) "/" else "."
      exp_keys <- if (!is.null(za)) expected_chunk_keys(unlist(za$shape), unlist(za$chunks), sep) else character(0)
      missing_n <- if (!is.null(za)) length(setdiff(exp_keys, chunk_files$key)) else NA_integer_
      extra_n <- if (!is.null(za)) length(setdiff(chunk_files$key, exp_keys)) else NA_integer_
      tibble(
        Container = container_label,
        Store = st,
        Top_Level_Folder = paste(top_folder, collapse = "; "),
        Top_Level_Matches_Container_Name = if (is_tar) identical(top_folder, tar_base) else NA,
        Level = lv,
        Levels_Present = paste(levels, collapse = ","),
        Levels_In_Metadata = paste(ds_paths, collapse = ","),
        Metadata_Level_Present = if (length(ds_paths) > 0) all(ds_paths %in% levels) else NA,
        Entry_Count = nrow(listing),
        Chunk_Files = nrow(chunk_files),
        Chunk_GB = round(sum(chunk_files$size, na.rm = TRUE) / 1024^3, 3),
        Empty_Chunk_Files = sum(chunk_files$size == 0, na.rm = TRUE),
        Shape = if (!is.null(za)) paste(unlist(za$shape), collapse = " x ") else NA_character_,
        Chunk_Shape = if (!is.null(za)) paste(unlist(za$chunks), collapse = " x ") else NA_character_,
        Dtype = if (!is.null(za)) za$dtype else NA_character_,
        Compressor = if (!is.null(za) && !is.null(za$compressor)) paste0(za$compressor$id, "/", za$compressor$cname) else NA_character_,
        Zarr_Format = if (!is.null(za)) za$zarr_format else NA_integer_,
        Expected_Chunks = length(exp_keys),
        Missing_Chunks = missing_n,
        Unexpected_Chunks = extra_n,
        Axes = paste(axes, collapse = ","),
        Axis_Units = paste(units, collapse = ","),
        Voxel_Scale = if (!is.null(ms$datasets[[1]]$coordinateTransformations[[1]]$scale)) paste(unlist(ms$datasets[[1]]$coordinateTransformations[[1]]$scale), collapse = " x ") else NA_character_,
        Channel_Labels = paste(ch_labels, collapse = ","),
        NGFF_Version = if (!is.null(ms$version)) ms$version else NA_character_,
        Orientation = if (!is.null(zattrs$orientation)) zattrs$orientation else NA_character_,
        Sidecar_Found = !is.null(side),
        Sidecar_PixelSize = pix,
        Sidecar_Instrument = if (!is.null(side$InstrumentModel)) side$InstrumentModel else NA_character_,
        Sidecar_Tile_Files = length(tile_names),
        Sidecar_Tile_Name_Style = tile_style,
        Sidecar_Tile_Source_Codes = paste(tile_code, collapse = "; "),
        Sidecar_Tile_Source_Descriptor = paste(tile_desc, collapse = "; "),
        Subject_ID = subj_id,
        Tile_Code_vs_Subject_Mismatch = id_mismatch,
        Samples_Tsv_Matches_Subject = samples_ok,
        Template_Placeholders = placeholders,
        Tar_Distinct_Owner_Accounts = n_owner,
        Tar_Has_Named_Owner_Account = named_owner,
        Status = "Success"
      )
    })
  })
}
```

## Running the Inspection // Exécution de l'inspection

Every tar and every on-disk store is inspected. Metadata extracted from tars goes to a temporary folder that is deleted at the end.

//

Chaque fichier tar et chaque magasin sur disque est inspecté. Les métadonnées extraites des fichiers tar sont placées dans un dossier temporaire supprimé à la fin.

```{r}
#| label: execution

results <- list()
tmp_root <- file.path(tempdir(), "omezarr_meta")

for (i in seq_along(tar_files)) {
  fp <- tar_files[i]
  message(sprintf("[%d/%d] %s (%.1f GB)", i, length(tar_files), basename(fp), file.size(fp) / 1024^3))
  res <- tryCatch({
    listing <- list_tar(fp)
    if (!any(grepl("\\.zarr(/|$)", listing$path, ignore.case = TRUE))) {
      tibble(Container = basename(fp), Store = NA_character_, Level = NA_character_,
             Status = "Tar does not contain a .zarr store")
    } else {
      dest <- file.path(tmp_root, sprintf("t%03d", i))
      extract_meta(fp, dest)
      inspect_container(basename(fp), fp, listing, dest, is_tar = TRUE)
    }
  }, error = function(e) {
    tibble(Container = basename(fp), Store = NA_character_, Level = NA_character_, Status = paste("Failed:", conditionMessage(e)))
  })
  results[[length(results) + 1]] <- res
}

for (zd in zarr_dirs) {
  message(sprintf("On-disk store: %s", zd))
  res <- tryCatch({
    listing <- list_disk_zarr(zd)
    inspect_container(basename(zd), zd, listing, dirname(zd), is_tar = FALSE)
  }, error = function(e) {
    tibble(Container = basename(zd), Store = NA_character_, Level = NA_character_, Status = paste("Failed:", conditionMessage(e)))
  })
  results[[length(results) + 1]] <- res
}

report <- bind_rows(results)
unlink(tmp_root, recursive = TRUE)
```

## Save Results // Enregistrer les résultats

```{r}
#| label: save-results

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

output_file <- file.path(output_dir, paste0("OMEZarr_Report_", dir_label, "_", format(Sys.Date(), "%Y%m%d"), ".csv"))
write_excel_csv(report, output_file)

desc_map <- c(
  Container = "Tar file name (or .zarr folder name) inspected.",
  Store = "Path of the .zarr store inside the container.",
  Top_Level_Folder = "Top-level folder name(s) inside the tar.",
  Top_Level_Matches_Container_Name = "TRUE if the single top-level folder equals the tar name without .tar (what a README usually promises: 'extracts to <name>/').",
  Level = "Resolution level (numeric folder inside the store) this row describes.",
  Levels_Present = "All numeric resolution-level folders found in the store.",
  Levels_In_Metadata = "Dataset paths listed in .zattrs multiscales.",
  Metadata_Level_Present = "TRUE if every level named in the metadata exists in the store.",
  Entry_Count = "Number of entries (files and folders) in the container.",
  Chunk_Files = "Number of chunk files at this level (everything except .zarray/.zattrs/.zgroup).",
  Chunk_GB = "Total size of the chunk files at this level, in GiB (as stored, compressed).",
  Empty_Chunk_Files = "Chunk files of 0 bytes (would fail to decode).",
  Shape = "Array shape from .zarray.",
  Chunk_Shape = "Chunk shape from .zarray.",
  Dtype = "Numpy-style data type from .zarray (e.g. <u2 = little-endian uint16).",
  Compressor = "Compressor id/codec from .zarray.",
  Zarr_Format = "zarr_format from .zarray (2 or 3).",
  Expected_Chunks = "Chunks the shape/chunk grid implies.",
  Missing_Chunks = "Expected chunk keys with no file present (nonzero = incomplete store; all-zero background chunks are sometimes legitimately omitted by writers, check the fill_value before treating as an error).",
  Unexpected_Chunks = "Chunk files present that the shape/chunk grid does not account for.",
  Axes = "Axis names from the OME-NGFF multiscales metadata.",
  Axis_Units = "Axis units from the metadata ('-' = none declared, e.g. channel axis).",
  Voxel_Scale = "Scale coordinate transformation of the first dataset (per axis, in Axis_Units).",
  Channel_Labels = "Channel labels from the omero metadata.",
  NGFF_Version = "OME-NGFF version from the multiscales metadata.",
  Orientation = "Non-standard 'orientation' string from .zattrs, if present.",
  Sidecar_Found = "TRUE if a BIDS-Microscopy .json sidecar sits next to the store.",
  Sidecar_PixelSize = "PixelSize from the sidecar (acquisition resolution, may differ from the level stored).",
  Sidecar_Instrument = "InstrumentModel from the sidecar.",
  Sidecar_Tile_Files = "Number of acquisition tile file names found in TileConfiguration.",
  Sidecar_Tile_Name_Style = "sample-coded (names start with a time-of-day prefix and a sample code), generic (bare names such as Z0000.ome.tif, one per z-slice, no sample code to check), or mixed.",
  Sidecar_Tile_Source_Codes = "Number(s) found in the sample descriptor of the acquisition tile file names.",
  Sidecar_Tile_Source_Descriptor = "Sample descriptor(s) from the tile file names without the time-of-day prefix (e.g. sample code, stain, side).",
  Subject_ID = "Subject ID parsed from the store path (sub-...).",
  Tile_Code_vs_Subject_Mismatch = "TRUE if the numeric part of the tile source code differs from the subject number: the acquisition files may belong to a different subject than the container name says. Confirm with the depositor.",
  Samples_Tsv_Matches_Subject = "TRUE if samples.tsv participant_id equals the store's subject ID.",
  Template_Placeholders = "Leftover boilerplate in dataset_description.json / README.md (Name, License, Authors, README_TODO).",
  Tar_Distinct_Owner_Accounts = "Number of distinct owner/group strings in the tar headers (values not recorded).",
  Tar_Has_Named_Owner_Account = "TRUE if the tar headers carry a user account name other than root/0 (value not recorded; consider whether it identifies a person).",
  Status = "Success, or the reason the container could not be read."
)
codebook <- tibble(
  Variable = names(report),
  Type = map_chr(report, function(col) {
    if (is.logical(col)) "Logical" else if (is.integer(col)) "Integer" else if (is.numeric(col)) "Numeric" else "Text"
  }),
  Description = unname(ifelse(names(report) %in% names(desc_map), desc_map[names(report)], ""))
)
write_excel_csv(codebook, file.path(output_dir, "OMEZarr_Report_Codebook.csv"))

ok <- report %>% filter(Status == "Success")
message(sprintf("Process complete. %d store-level row(s) from %d container(s); %d failed or without a store.",
                nrow(ok), n_distinct(ok$Container), nrow(report) - nrow(ok)))
if (nrow(ok) > 0) {
  message("  NOTE: structure and metadata only; chunk payloads were not decoded (no pixel-value check).")
  if (any(ok$Missing_Chunks > 0, na.rm = TRUE)) message(sprintf("  WARNING: %d row(s) have missing chunks.", sum(ok$Missing_Chunks > 0, na.rm = TRUE)))
  if (any(ok$Unexpected_Chunks > 0, na.rm = TRUE)) message(sprintf("  WARNING: %d row(s) have unexpected chunk files.", sum(ok$Unexpected_Chunks > 0, na.rm = TRUE)))
  if (any(ok$Empty_Chunk_Files > 0, na.rm = TRUE)) message(sprintf("  WARNING: %d row(s) have 0-byte chunk files.", sum(ok$Empty_Chunk_Files > 0, na.rm = TRUE)))
  if (any(ok$Top_Level_Matches_Container_Name == FALSE, na.rm = TRUE)) message("  WARNING: top-level folder name does not match the tar name for some containers.")
  if (any(ok$Tile_Code_vs_Subject_Mismatch, na.rm = TRUE)) message(sprintf("  WARNING: %d container(s) have acquisition tile source codes that do not match the subject ID.", sum(ok$Tile_Code_vs_Subject_Mismatch, na.rm = TRUE)))
  if (any(nzchar(ok$Template_Placeholders), na.rm = TRUE)) message(sprintf("  NOTE: %d container(s) still hold template placeholder text.", sum(nzchar(ok$Template_Placeholders), na.rm = TRUE)))
  if (any(ok$Tar_Has_Named_Owner_Account, na.rm = TRUE)) message("  NOTE: tar headers carry a named user account; check whether it identifies a person.")
}
message(paste("Report saved to:", output_file))
```

## Curation Insights // Aperçu de la curation

Use this report to decide what to ask the depositor:

-   **Missing_Chunks \> 0:** The store is incomplete. Some writers legitimately omit all-zero background chunks, so check the `fill_value` before treating it as an error.

-   **Unexpected_Chunks \> 0 or Empty_Chunk_Files \> 0:** Chunk files that the shape does not account for, or 0-byte chunk files that would fail to decode, point to a damaged or mis-packed store.

-   **Top_Level_Matches_Container_Name = FALSE:** A README often promises that a tar extracts to a folder with the tar's name. A different top-level folder name means that the documentation does not match the container.

-   **Metadata_Level_Present = FALSE:** The metadata lists a resolution level that is not in the store.

-   **Axes, Axis_Units, Voxel_Scale:** Missing units or scale make the image impossible to measure. Check them against the instrument.

-   **Tile_Code_vs_Subject_Mismatch = TRUE:** The acquisition files carry a different subject code than the container. Confirm with the depositor that the right data was packed.

-   **Template_Placeholders:** Boilerplate text (such as `Author Name 1`, `README_TODO`) means the descriptive metadata was never filled in.

-   **Tar_Has_Named_Owner_Account = TRUE:** The tar headers carry a user account name. The value is deliberately not recorded here, so inspect the tar and decide whether it identifies a person.

-   **Structure only:** The notebook confirms structure and metadata. It does not decode the chunks, so pixel values are not validated.

//

Utilisez ce rapport pour décider de ce qu'il faut demander au déposant :

-   **Missing_Chunks \> 0 :** Le magasin est incomplet. Certains outils d'écriture omettent légitimement les fragments d'arrière-plan entièrement à zéro ; vérifiez donc la valeur `fill_value` avant d'y voir une erreur.

-   **Unexpected_Chunks \> 0 ou Empty_Chunk_Files \> 0 :** Des fichiers de fragments que la forme n'explique pas, ou des fragments de 0 octet qui ne pourraient pas être décodés, indiquent un magasin endommagé ou mal empaqueté.

-   **Top_Level_Matches_Container_Name = FALSE :** Un README promet souvent qu'un fichier tar s'extrait dans un dossier portant le nom du tar. Un nom de dossier de premier niveau différent signifie que la documentation ne correspond pas au conteneur.

-   **Metadata_Level_Present = FALSE :** Les métadonnées énumèrent un niveau de résolution absent du magasin.

-   **Axes, Axis_Units, Voxel_Scale :** Des unités ou une échelle manquantes rendent l'image impossible à mesurer. Vérifiez-les par rapport à l'instrument.

-   **Tile_Code_vs_Subject_Mismatch = TRUE :** Les fichiers d'acquisition portent un code de sujet différent de celui du conteneur. Confirmez avec le déposant que les bonnes données ont été empaquetées.

-   **Template_Placeholders :** Du texte type (comme `Author Name 1` ou `README_TODO`) signifie que les métadonnées descriptives n'ont jamais été remplies.

-   **Tar_Has_Named_Owner_Account = TRUE :** Les en-têtes tar contiennent un nom de compte d'utilisateur. La valeur n'est volontairement pas consignée ici ; examinez le fichier tar et décidez si elle identifie une personne.

-   **Structure seulement :** Le cahier confirme la structure et les métadonnées. Il ne décode pas les fragments ; les valeurs des pixels ne sont donc pas validées.

## Additional Tools // Outils supplémentaires

-   **OME-NGFF specification:** The [specification](https://ngff.openmicroscopy.org/) of the OME-Zarr metadata convention.

-   **Zarr:** The [Zarr project](https://zarr.dev/) documents the array storage format and its implementations.

-   **ome-zarr-py:** A Python [reference reader](https://ome-zarr.readthedocs.io/) for OME-Zarr that can decode the pixel data this notebook does not read.

-   **Checksum verification:** For a full check of every chunk file, combine this notebook with the [checksum manifest verification](Verify_ChecksumManifest_Notebook.qmd).

//

-   **Spécification OME-NGFF :** La [spécification](https://ngff.openmicroscopy.org/) de la convention de métadonnées OME-Zarr.

-   **Zarr :** Le [projet Zarr](https://zarr.dev/) documente le format de stockage des tableaux et ses implémentations.

-   **ome-zarr-py :** Un [lecteur de référence](https://ome-zarr.readthedocs.io/) Python pour OME-Zarr, capable de décoder les données de pixels que ce cahier ne lit pas.

-   **Vérification des sommes de contrôle :** Pour un contrôle complet de chaque fichier de fragment, combinez ce cahier avec la [vérification d'un manifeste de sommes de contrôle](Verify_ChecksumManifest_Notebook.qmd).

## Using the Non-Interactive R Script // Utilisation du script R non interactif

For users who want to run this analysis on a server, in a batch job, or from the command line, here is a pure R script that performs the same process.

Download the **R Script:** [**`Inspect_OMEZarr_Script.R`**](Scripts/Inspect_OMEZarr_Script.R)

//

Pour les utilisateurs qui souhaitent exécuter cette analyse sur un serveur, dans le cadre d'un traitement par lots ou depuis la ligne de commande, voici un script R pur qui effectue le même processus.

Télécharger le **script R :** [**`Inspect_OMEZarr_Script.R`**](Scripts/Inspect_OMEZarr_Script.R)

### Example HPC Submission Script // Exemple de script de soumission HPC

`Inspect_OMEZarr_submit.sh`

``` bash
#!/bin/bash
#SBATCH --job-name=omezarr_check
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=04:00:00
#SBATCH --mem=8G
#SBATCH --output=logs/omezarr_check_%j.log

# Load R Module
module load R

# The tar program is normally available on cluster nodes.
# export TAR_PATH="/usr/bin/tar"   # only if needed

# Define directories
DATA_DIR="/scratch/user/project_data/lightsheet"
OUTPUT_DIR="/scratch/user/omezarr_results"

# Prepare Environment
mkdir -p logs

# Run Analysis
echo "Starting OME-Zarr inspection on $DATA_DIR"
Rscript Inspect_OMEZarr_Script.R "$DATA_DIR" "$OUTPUT_DIR"
```

## References // Références

::: {#refs}
:::
