---
title: "File extensions and formats"
author: "Natalie Williams"
date: "2025-11-13"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: true
bibliography: references.bib
params:
  target_dir: "data/Inspect_Extensions/"
---

## Overview

This notebook performs the initial triage of a data package by inventorying file extensions.

::: {.callout-note title="Curation Goal"}
Establish an initial inventory of all file types within a submission. Our objective is to distinguish between extension-based labels and internal format signatures to identify mislabeled, obsolete, or high-risk files.
:::

::: {.callout-warning title="Identifying Risks"}
A file extension is merely a textual label. Relying on it can hide malicious executables renamed as media files or proprietary formats that lack a long-term migration path [@brown2013].
:::

**Curation Objectives:**

1.  **Content Profiling:** Establish the composition of the dataset (e.g., Tabular vs. Proprietary).
2.  **Junk Detection:** Identify system artifacts (`.DS_Store`, `Thumbs.db`) that clutter the archive.
3.  **Risk Flagging:** Detect executables or zero-byte files that indicate transfer errors.

------------------------------------------------------------------------

## Setup

We utilize the `tidyverse` [@tidyverse] for data manipulation and `tools` [@tools] for path parsing.

### R Packages

If you don't have the `tidyverse` package installed, run this command once in your R console:

```{r}
# install.packages("tidyverse", "DT", "tools", "rstudioapi", "exiftoolr")
```

------------------------------------------------------------------------

### Load Libraries

This chunk loads the necessary library for the session.

**Note:** The `exiftoolr` package requires the [ExifTool software](https://exiftool.org/). The code below checks for it and attempts to install it locally if missing.

```{r}
#| label: load-libraries
#| message: false

library(tidyverse)
library(DT)         # For interactive tables
library(tools)      # For file path handling
library(rstudioapi) # For interactive selection
library(exiftoolr)  # R wrapper for ExifTool

# Check if ExifTool is available; if not, try to install locally
if (is.null(exif_version())) {
  message("ExifTool not found. Attempting local installation...")
  tryCatch({
    install_exiftool()
    message("ExifTool installed successfully.")
  }, error = function(e) {
    warning("Could not install ExifTool. Deep metadata extraction may fail.")
  })
}
```

------------------------------------------------------------------------

## Select Target Directory

This section identifies the directory to be analyzed and creates a complete list of every file within it, including those in sub-folders and hidden files. The directory is set by the `target_dir` parameter at the top of this document.

```{r}
#| label: select-target-dir

# Hybrid Selection Logic
if (interactive() && requireNamespace("rstudioapi", quietly = TRUE)) {
    message("Running in interactive mode. Please select a directory.")
    selected_dir <- rstudioapi::selectDirectory(caption = "Select Data Directory")

    if (is.null(selected_dir)) {
        stop("No directory selected. Analysis aborted.")
    }
    target_dir <- selected_dir
} else {
    # Fallback for rendering/batch mode
    target_dir <- params$target_dir
}

# Validation
if (!dir.exists(target_dir)) {
    stop(paste("Target directory does not exist:", target_dir))
}

message("Inventorying directory: ", target_dir)
```

------------------------------------------------------------------------

## Generate File Inventory

We construct a detailed table including the relative path to locate files and detect risks.

Data Collected:

-   **Path:** The relative path to the file (critical for locating "bad" files).

-   **Extension:** The file extension (normalized to lowercase).

-   **Size:** File size in Bytes and Megabytes (MB).

-   **Flags:** Automatic tags for "Empty", "System Junk", or "Executable".

```{r}
#| label: generate-inventory
#| warning: false
#| message: false

message("Scanning files... This may take a moment for large directories.")

# Normalize to forward slashes so path arithmetic below is Windows-safe
target_dir <- normalizePath(target_dir, winslash = "/", mustWork = TRUE)

# 1. Get all files recursively
all_files <- list.files(
  path = target_dir,
  recursive = TRUE,
  full.names = TRUE,
  all.files = TRUE
)

message("Found ", length(all_files), " total files.")

if (length(all_files) > 0) {

  # 2. Build Inventory Dataframe
  inventory <- tibble(FullPath = all_files) %>%
    mutate(
      FullPath = normalizePath(FullPath, winslash = "/", mustWork = FALSE),
      FileName = basename(FullPath),
      # Calculate relative path for readability
      RelativePath = str_remove(FullPath, fixed(paste0(target_dir, "/"))),

      # Extract extension: Get text after last dot, lowercase it.
      Extension = tolower(file_ext(FileName)),
      Extension = if_else(Extension == "", "(no extension)", paste0(".", Extension)),
      
      # Get File Size
      Size_Bytes = file.size(FullPath),
      Size_MB = round(Size_Bytes / 1024^2, 4)
    )

  # 3. Add Risk Flags
  junk_patterns <- c("\\.ds_store", "thumbs\\.db", "__macosx")
  exec_patterns <- c("\\.exe$", "\\.bat$", "\\.sh$", "\\.bin$", "\\.jar$")

  inventory <- inventory %>%
    mutate(
      Risk_Flag = case_when(
        Size_Bytes == 0 ~ "Zero-Byte File",
        str_detect(tolower(FileName), paste(junk_patterns, collapse = "|")) ~ "System Junk",
        str_detect(tolower(FileName), paste(exec_patterns, collapse = "|")) ~ "Executable",
        TRUE ~ "Clean"
      )
    )

  # Preview
  datatable(inventory %>% select(RelativePath, Extension, Size_MB, Risk_Flag),
            options = list(pageLength = 10, scrollX = TRUE),
            caption = "Table 1: Full File Inventory with Risk Flags")
            
} else {
  message("Directory is empty.")
  inventory <- tibble()
}
```

## Deep Metadata Extraction (ExifTool)

This section goes beyond the filename. We use ExifTool to read the file headers. This allows us to:

-   **Verify Formats:** Does the MIMEType match the extension?

-   **Check Integrity:** Does ExifTool report any Warning (e.g., "Corrupted header")?

-   **Extract Context:** Get Author, CreateDate, or ImageSize metadata.

```{r}
#| label: run-exiftool
#| warning: false
#| message: false

if (nrow(inventory) > 0 && !is.null(exif_version())) {

  message("Running ExifTool on ", nrow(inventory), " files...")

  # We select specific tags to keep the report manageable
  tags_to_extract <- c("FileName", "MIMEType", "FileType", "Author", "CreateDate", "Warning", "ImageSize")

  # Read files one at a time: a single unreadable/unrecognized file (e.g. no
  # extension, or a warning ExifTool emits outside the JSON stream) can
  # otherwise corrupt the JSON for the whole batch and abort every file's metadata.
  safe_exif_read <- purrr::safely(function(fp) {
    exif_read(fp, tags = tags_to_extract, quiet = TRUE)
  })

  exif_results <- purrr::map(inventory$FullPath, safe_exif_read)
  n_failed <- sum(purrr::map_lgl(exif_results, ~ !is.null(.x$error)))
  if (n_failed > 0) {
    message("Warning: ExifTool could not read metadata for ", n_failed, " file(s); leaving those blank.")
  }

  exif_data <- purrr::map(exif_results, "result") %>%
    purrr::compact() %>%
    bind_rows()

  if (nrow(exif_data) > 0) {
    # Clean up column names (ExifTool returns "SourceFile" as the path)
    exif_data <- exif_data %>%
      rename(FullPath = SourceFile)
    if ("FileName" %in% names(exif_data)) exif_data <- select(exif_data, -FileName)

    # Join with main inventory
    inventory <- left_join(inventory, exif_data, by = "FullPath")
  }

  # Ensure expected columns always exist, even if every read failed
  for (col in c("MIMEType", "FileType", "Author", "CreateDate", "Warning")) {
    if (!col %in% names(inventory)) inventory[[col]] <- NA_character_
  }

  message("Metadata extraction complete.")
}

# Preview Combined Data
if (nrow(inventory) > 0) {
  cols_to_show <- c("RelativePath", "Extension", "MIMEType", "Risk_Flag")
  if ("Warning" %in% names(inventory)) cols_to_show <- c(cols_to_show, "Warning")
  
  datatable(inventory %>% select(any_of(cols_to_show)),
            options = list(pageLength = 10, scrollX = TRUE),
            caption = "Table 1: File Inventory with ExifTool Metadata")
}
```

## Extension Summary and Visualization

We aggregate the inventory to visualize the dataset's content profile. This helps curators identify if the submission matches the depositor's description (e.g., "They said it was images, but I see mostly .docx files").

```{r}
#| label: viz-extensions
#| fig-cap: "Figure 1: Top 15 File Formats by Frequency"

if (nrow(inventory) > 0) {
  
  # Use MIMEType if available, otherwise Extension
  plot_data <- inventory %>%
    mutate(Format_Label = if_else(!is.na(MIMEType) & MIMEType != "", MIMEType, Extension)) %>%
    count(Format_Label, name = "Count", sort = TRUE) %>%
    head(15)

  ggplot(plot_data, aes(x = reorder(Format_Label, Count), y = Count)) +
    geom_col(fill = "#4C78A8") +
    coord_flip() +
    labs(
      title = "File Format Distribution",
      subtitle = "Based on MIME Type (ExifTool) where available",
      x = "Format / MIME Type",
      y = "File Count"
    ) +
    theme_minimal()
}
```

## Save Results

We export the full inventory to help curators locate and delete the specific files flagged as "Junk" or "Zero-Byte."

```{r}
#| label: save-results

output_dir <- file.path("Results", "Inspect_Extensions")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Label for output filenames: name of the folder that was explored
dir_label <- gsub("[^A-Za-z0-9_.-]", "_", basename(sub("[/\\\\]+$", "", target_dir)))

summary_file <- file.path(output_dir, paste0("Format_Summary_", dir_label, "_", Sys.Date(), ".csv"))
inventory_file <- file.path(output_dir, paste0("Full_Inventory_ExifTool_", dir_label, "_", Sys.Date(), ".csv"))

if (nrow(inventory) > 0) {
  # Generate summary
  summary_table <- inventory %>%
    count(Extension, MIMEType, Risk_Flag, name = "Count") %>%
    arrange(desc(Count))

  write_excel_csv(summary_table, summary_file)
  write_excel_csv(inventory, inventory_file)

  message("Reports saved:")
  message("1. ", summary_file)
  message("2. ", inventory_file)
}
```

------------------------------------------------------------------------

## Curation Insights

-   **Zero-Byte Risks:** A file with 0 bytes is technically valid in the filesystem but semantically empty. It may indicate a failed file transfer (e.g., an FTP interruption). Check this is the case in the original deposit for this file, then check with the depositor to confirm the file is safe to remove.

-   **Extension Mismatches:** If you see a generic extension like .dat or .file, the depositor may have removed the extension. Use tools like DROID (see below) to identify the true format.

-   **Corrupted Files:** Filter for rows where the Warning column is not empty. ExifTool often detects truncated files that standard file system checks miss.

-   **Preservation vs. Access:** Distinguish between "Originals" (e.g., .wav, .tiff) and "Derivatives" (e.g., .mp3, .jpg). Ideally, the archive should preserve the uncompressed originals (see https://www.loc.gov/preservation/digital/formats/).

-   **Metadata Privacy:** Check the Author or GPS columns, if extracted. Check if these contain real names or locations for a de-identified dataset, and discuss with the depositor to confirm the files can and should be scrubbed before publishing.

## Additional Tools for Researchers

This notebook performs some file identification. Here are some other tools for file identification.

-   **Siegfried / DROID:** A tool for [format identification](nationalarchives.gov.uk/information-management/manage-information/preserving-digital-records/droid/). It scans the file's binary signature (Magic Number) and matches it against PRONOM, the technical registry maintained by The National Archives UK (see https://www.itforarchivists.com/siegfried).

-   **Apache Tika:** A [toolkit](tika.apache.org) that detects thousands of file types and extracts metadata (e.g., Author, Date) from within the files.

-   **FIDO (Format Identification for Digital Objects):** A command-line tool designed for simplified integration into preservation pipelines (see https://openpreservation.org/tools/fido/).

## Using the Non-Interactive R Script

For users who want to run this analysis on a server or from the command line, a pure R script is available to perform the same process.

**Prerequisites:** \* R with `tidyverse`, `magick`, and `exiftoolr` installed. \* **ExifTool** command-line software must be installed on the system.

Download the **R Script:** [**`Inspect_Extensions_Script.R`**](Scripts/Inspect_Extensions_Script.R)

### Example HPC Submission Script

``` bash
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --time=00:30:00
#SBATCH --job-name=ext_check_exif

# Load R Module
module load R

# Load Perl module (Often required for ExifTool to run)
module load perl 

# Define directories
DATA_DIR="/scratch/your_user/data_folder"
OUTPUT_DIR="/scratch/your_user/inventory_results"

# Run Script
Rscript Inspect_Extensions_Script.R $DATA_DIR $OUTPUT_DIR
```

## References

::: {#refs}
:::
